---
title: ":some_gene: expression in various datasets"
date: today
date-format: iso
format: 
  html: 
    embed-resources: true
    toc: true
code-fold: true
execute: 
  eval: true
  echo: true
  output: true
params:
  gene: ""
---

# Instructions for generating this report

Use Quarto in the command line to generate the gene expression report for your gene of interest. For example, to run the code for *BRCA1*, run the following in the command line:

```bash
# set gene of interest
goi=BRCA1

# search and replace in the file
sed -i "s/:some_gene:/$goi/g" copy_of_gene_expression_report.qmd

# run quarto to generate report
quarto render copy_of_gene_expression_report.qmd -P gene:$goi
```

Alternatively, add BRCA1 to the YAML header of this file and then render the report.

```yaml
params:
  gene: "BRCA1"
```

# Initizalization

```{r}
library(tidyverse)
library(Seurat)
library(patchwork)

source("/CBL_scRNAseq/software/utilities/cell_labelling.R")
```

```{r}
integ_srat <- readRDS("/CBL_scRNAseq/results/integrated/vladoiu_liam_RL.rds")

# add cell type to integ_clusters
integ_srat <- label_integ_clusters(
  integ_srat,
  integ_method = "cca",
  datasets = "rl"
)
```


# Developing human cerebellum

Datasets of the developing human cerebellum.

## Aldinger et al., Nat Neurosci (2021)

Load the scRNA-seq dataset that contains the RL lineage cells only (file received from Liam Hendrikse).

```{r}
srat <- readRDS(
  "/isilon/CBL_scRNAseq-archived/data/human/Aldinger/glutamatergic_dev_Liam.RDS"
)
```

Copy the annotations from the mouse/human integration.

```{r}
srat$integ_clusters <- integ_srat@meta.data[row.names(srat@meta.data), "integ_clusters"]
```

Show basic UMAPs

```{r}
#| label: fig-aldinger-annot
#| fig-cap: "UMAP of human RL lineage cells coloured by (A) Liam's annotations or (B) the human/mouse integration annotations."
#| fig-height: 12

.plt1 <- DimPlot(srat, group.by = "new_cell_type", label = TRUE, repel = TRUE) + 
  NoLegend()

.plt2 <- DimPlot(srat, group.by = "integ_clusters", label = TRUE, repel = TRUE) + 
  NoLegend()

wrap_plots(
  list(.plt1, .plt2),
  nrow = 2
) + 
  plot_annotation(tag_levels = "A")
```

Show expression of *`r params$gene`*

::: {.callout-note}
The RNA assay in the Seurat object is being used to visualize the expression data.
:::

```{r}
#| label: fig-aldinger-gene-expr
#| fig-cap: !expr glue::glue("{params$gene} expression in the Aldinger scRNA-seq dataset. (A-B) UMAP of {params$gene} expression without (A) and with (B) expression cutoffs. (C-D) Violin plot of {params$gene} expression split by (C) Liam's annotations or (D) the human/mouse integration annotations.")
#| fig-height: 24

DefaultAssay(srat) <- "RNA"

.plt1 <- FeaturePlot(
  srat,
  features = params$gene,
  order = TRUE
) + 
  labs(title = "no expression cutoff")

.plt2 <- FeaturePlot(
  srat,
  features = params$gene,
  order = TRUE,
  min.cutoff = "q10",
  max.cutoff = "q90"
) + 
  labs(title = "min.cutoff/max.cutoff = q10/q90")

.plt3 <- VlnPlot(
  srat,
  features = params$gene,
  group.by = "new_cell_type"
) + NoLegend()

.plt4 <- VlnPlot(
  srat,
  features = params$gene,
  group.by = "integ_clusters"
) + NoLegend()

wrap_plots(
  list(.plt1, .plt2, .plt3, .plt4),
  nrow = 4
) + 
  plot_annotation(tag_levels = "A")
```

<!-- ## Cao et al., Science (2020) -->

<!-- Load the scRNA-seq dataset -->


# Medulloblastoma

## Vladoiu et al., Nature (2019)

Load the tumour scRNA-seq dataset containing eight tumour samples: two SHH, two Group 3, and four Group 4. Samples were integrated using fastMNN.

```{r}
srat <- readRDS("/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds")
```

Show basic UMAPs

```{r}
#| label: fig-vladoiu-annot
#| fig-cap: "UMAP of MB tumour cells annotated by (A) sample ID, (B) MB subtype, and (C) Seurat clusters."
#| fig-height: 18

.plt <- purrr::map(
  .x = c("orig.ident", "subtype", "seurat_clusters"),
  .f = \(x) {
    p <- DimPlot(
      srat,
      group.by = x,
      label = TRUE,
      repel = TRUE
    ) + 
      NoLegend()
    return(p)
  }
)

wrap_plots(
  .plt,
  nrow = 3
) + 
  plot_annotation(tag_levels = "A")
```

Show expression of *`r params$gene`*

::: {.callout-note}
The RNA assay in the Seurat object is being used to visualize the expression data.
:::

```{r}
#| label: fig-vladoiu-gene-expr
#| fig-cap: !expr glue::glue("{params$gene} expression in the Vladoiu tumour scRNA-seq dataset. (A-B) UMAP of {params$gene} expression without (A) and with (B) expression cutoffs. (C-E) Violin plot of {params$gene} expression split by (C) sample ID, (D) MB subtype, or (E) Seurat clusters.")
#| fig-height: 30

DefaultAssay(srat) <- "RNA"

.plt1 <- FeaturePlot(
  srat,
  features = params$gene,
  order = TRUE
) + 
  labs(title = "no expression cutoff")

.plt2 <- FeaturePlot(
  srat,
  features = params$gene,
  order = TRUE,
  min.cutoff = "q10",
  max.cutoff = "q90"
) + 
  labs(title = "min.cutoff/max.cutoff = q10/q90")

.plt <- purrr::map(
  .x = c("orig.ident", "subtype", "seurat_clusters"),
  .f = \(x) {
    p <- VlnPlot(
      srat,
      features = params$gene,
      group.by = x
    ) + 
      NoLegend()
    return(p)
  }
)

wrap_plots(
  c(list(.plt1, .plt2), .plt),
  nrow = 5
) + 
  plot_annotation(tag_levels = "A")
```

# Session info

```{r}
print(sessionInfo())
```
