---
title: Differential gene expression in integrated dataset
date: 2024-05-08
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine the differential gene expression in integrated dataset.

## Files and directories

### Input

Original CCA integration file can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240506/cca/20240506_cca_integ.qs`

### Output

Results can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca`

## Process and Results

### Dataset Filtering
Loading in necessary libraries. Note we have to use the clustree conda environment due to presto requiring R version 4.3
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
library(presto)
library(ggrepel)
```
```{r eval=FALSE} 
integ_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240506/cca/20240506_cca_integ.qs")
```

The CCA integration dataset was filtered to only contain the 5000 variables genes that the integration was done on. The variable genes was found using `/u/llau/software/mb_scrnaseq/MB_scRNAseq/scrnaseq_Leo/pipeline_scripts/integration_prep.R`. Specifically the below line which only found 4906 genes to be variable features.
```{r eval=FALSE}
features <- SelectIntegrationFeatures(dataset_list, nfeatures = 5000)
```

The integrated dataset was filtered to only contain the variable genes using:
```{r eval=FALSE}
variable_integ <- subset(integ_cca, features = integ_cca@assays$integrated@var.features)

# Save metadata
write.csv(variable_integ@meta.data, "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/variable_integ.csv")
```

The dataset was further filtered to only contain human cells:
```{r eval=FALSE}
human_variable_integ <- subset(variable_integ, subset = species == "human")

# Save metadata
write.csv(human_variable_integ@meta.data, "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/human_variable_integ.csv")
```

### Differential Gene Analysis

For all the differential gene analysis, we are using the 0.4 Resolution FindCluster() results.

The differential gene analysis was first conducted on the integrated dataset with filtered to only have the variable genes.
```{r eval=FALSE}
variable_markers <- wilcoxauc(
  variable_integ,
  group_by = "snn_res.0.4",
  assay = "data",
  seurat_assay = "RNA"
)

write.csv(variable_markers, file = "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/variable_markers.csv")
top_markers <- top_markers(variable_markers, n = 100)
write.csv(top_markers, file = "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/top_variable_markers.csv")
```

The same process was done for the human-only variable genes dataset
```{r eval=FALSE}
human_variable_markers <- wilcoxauc(
  human_variable_integ,
  group_by = "snn_res.0.4",
  assay = "data",
  seurat_assay = "RNA"
)

write.csv(human_variable_markers, file = "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/human_variable_markers.csv")
top_markers <- top_markers(human_variable_markers, n = 100)
write.csv(top_markers, file = "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/top_human_variable_markers.csv")
```

Using these markers, the volcano plots were generated.
```{r eval=FALSE}
source("/u/llau/software/mb_scrnaseq/MB_scRNAseq/software/utilities/plotting.R")
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/variable_genes_volcano_plots"

variable_markers$direction <- ifelse(variable_markers$logFC >= 0, "up", "down")
variable_markers$pval <- variable_markers$pval
log_fc_thresh <- 0.75
variable_markers$feature_labelled <- ifelse(abs(variable_markers$logFC) > log_fc_thresh & variable_markers$pval < 0.05, variable_markers$feature, NA)
variable_markers_list <- split(variable_markers, variable_markers$group)

for(cluster in variable_markers_list) {
  non_zero <- cluster$pval[cluster$pval != 0]
  min_val <- min(non_zero)
  cluster$pval <- ifelse(cluster$pval == 0, min_val*exp(-1), cluster$pval)
  plt <- make_volcano(
    data = cluster,
    log_fc = logFC,
    log_pval = -log10(pval),
    direction = direction,
    gene_labels = feature_labelled,
    log_fc_thresh = c(-log_fc_thresh, log_fc_thresh),
    log_pval_thresh = -log10(0.05),
    label_num_genes = TRUE
  ) + theme_classic()
  volcano_name <- paste0("cluster_", cluster$group[1], "_volcano_plot.png")
  ggsave(filename = volcano_name, plot = plt, path = out_directory, 
          width = 10, height = 10, units = "in", dpi = 600)
}
write.csv(variable_markers, file = "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/variable_genes_volcano_plots/variable_markers_adjusted.csv")
```
```{r eval=FALSE}
source("/u/llau/software/mb_scrnaseq/MB_scRNAseq/software/utilities/plotting.R")
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/human_variable_genes_volcano_plots"

human_variable_markers$direction <- ifelse(human_variable_markers$logFC >= 0, "up", "down")
human_variable_markers$pval <- human_variable_markers$pval
log_fc_thresh <- 0.75
human_variable_markers$feature_labelled <- ifelse(abs(human_variable_markers$logFC) > log_fc_thresh & human_variable_markers$pval < 0.05,         human_variable_markers$feature, NA)
human_variable_markers_list <- split(human_variable_markers, human_variable_markers$group)

for(cluster in human_variable_markers_list) {
  non_zero <- cluster$pval[cluster$pval != 0]
  min_val <- min(non_zero)
  cluster$pval <- ifelse(cluster$pval == 0, min_val*exp(-1), cluster$pval)
  plt <- make_volcano(
    data = cluster,
    log_fc = logFC,
    log_pval = -log10(pval),
    direction = direction,
    gene_labels = feature_labelled,
    log_fc_thresh = c(-log_fc_thresh, log_fc_thresh),
    log_pval_thresh = -log10(0.05),
    label_num_genes = TRUE
  ) + theme_classic()
  volcano_name <- paste0("cluster_", cluster$group[1], "_volcano_plot.png")
  ggsave(filename = volcano_name, plot = plt, path = out_directory, 
          width = 10, height = 10, units = "in", dpi = 600)
}
write.csv(human_variable_markers, file = "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca/human_variable_genes_volcano_plots/human_variable_markers_adjusted.csv")
```

## 0.4 Resolution Clustering Plots

Due to some of the plots being generated in 0.8 resolution, I will need to regenerate some plots.
```{r eval=FALSE}
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240508/cca"

integ_cca$snn_res.0.4 <- fct_relevel(integ_cca$snn_res.0.4, str_sort(levels(integ_cca$snn_res.0.4), numeric = TRUE))

# Cluster 0.4 UMAP
plt <- DimPlot(integ_cca, reduction = "umap", group.by = "snn_res.0.4", label = TRUE, label.size = 3, raster = FALSE) + 
        ggtitle("Resolution: 0.4")
      filename <- "CCA_0.4_integ_umap_clusters.png"
      ggsave(filename = filename, plot = plt, path = out_directory, 
              width = 10, height = 10, dpi = 600)

## Bar plots

# Species
source("/u/llau/software/mb_scrnaseq/MB_scRNAseq/software/utilities/cluster_barplot.R")
species_bar_plot <- cluster_barplot(integ_cca, split.by = "species", group.by = "snn_res.0.4")
species_bar_plot_name <- "CCA_0.4_integ_species_bar_plot.png"
ggsave(filename = species_bar_plot_name, plot = species_bar_plot, path = out_directory, 
          width = 20, height = 10, units = "in", dpi = 600)

# Common cell name
common_cell_type_bar_plot <- cluster_barplot(integ_cca, split.by = "common_cell_name", group.by = "snn_res.0.4")
common_cell_type_bar_plot_name <- "CCA_0.4_integ_common_cell_type_bar_plot.png"
ggsave(filename = common_cell_type_bar_plot_name, plot = common_cell_type_bar_plot, path = out_directory, 
          width = 20, height = 10, units = "in", dpi = 600)

# Cell type
cell_type_bar_plot <- cluster_barplot(integ_cca, split.by = "cell_type", group.by = "snn_res.0.4")
cell_type_bar_plot_name <- "CCA_0.4_integ_cell_type_bar_plot.png"
ggsave(filename = cell_type_bar_plot_name, plot = cell_type_bar_plot, path = out_directory, 
          width = 20, height = 10, units = "in", dpi = 600)

# Dataset
dataset_bar_plot <- cluster_barplot(integ_cca, split.by = "dataset_name", group.by = "snn_res.0.4")
dataset_bar_plot_name <- "CCA_0.4_integ_dataset_bar_plot.png"
ggsave(filename = dataset_bar_plot_name, plot = dataset_bar_plot, path = out_directory, 
          width = 20, height = 10, units = "in", dpi = 600)

## Dot plot

# Genes of interest
genes <- read.csv("/u/llau/software/mb_scrnaseq/MB_scRNAseq/scrnaseq_Leo/utilities/cell_gene_mapping.csv")
genes <- genes$gene
DefaultAssay(integ_cca) <- "RNA"
dot_plot <- DotPlot(object = integ_cca, features = genes, group.by = "snn_res.0.4") + theme_classic()
  dot_plot_name <- "CCA_0.4_integ_dot_plot.png"
  ggsave(filename = dot_plot_name, plot = dot_plot, path = out_directory, 
            width = 20, height = 20, units = "in", dpi = 600)
```

