---
title: Conducting pathway analysis
date: 2024-07-10
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
engine: knitr
---

# Goal of this analysis

1. Use Ian's script to conduct pathway analysis

# Files and directories

## Input

List of markers can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240708/ubc_markers.csv`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240710`

# Process and Results

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
```

Lets load in the list of markers
```{r eval=FALSE}
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240710"

markers <- read.csv("/.mounts/labs/pailab/private/llau/results/integrated/20240708/ubc_markers.csv")

# Subset for only UBC_1 markers
ubc_1_markers <- subset(markers, subset = cluster == "UBC_1")

write.csv(ubc_1_markers, file = paste0(out_directory, "/ubc_1_markers.csv"), row.names = FALSE)
```

Now lets run the pathway analysis script. 
```{bash eval=FALSE}
conda activate clustree
cd /u/llau/software/mb_scrnaseq/MB_scRNAseq/scrnaseq_Leo/utilities
Rscript ubcs_gost_enrichment.R --de_gene_file /.mounts/labs/pailab/private/llau/results/integrated/20240710/ubc_1_markers.csv --out_dir /.mounts/labs/pailab/private/llau/results/integrated/20240710
```

