---
title: Using new markers for UBC sub clusters in SingleR
date: 2024-07-17
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

# Goal of this analysis

1. Using the new cell type markers for SingleR

# Files and directories

## Input

CCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs`

Tumour file is found at `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240717`

# Process and Results

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
```

```{r eval=FALSE}
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240717"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
rl_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs")
```

## Marker determination

Step 1: Filter RL lineage for human cells
```{r eval=FALSE}
rl_cca <- rl_cca %>%
  subset(subset = species == "human") 
```

Step 2: Separate the UBC cells by dataset
```{r eval=FALSE}
ald <- subset(rl_cca, subset = dataset_name == "Aldinger_full_cerebellum_human")
sepp <- subset(rl_cca, subset = dataset_name == "Sepp_full_cerebellum_human")
```

Step 3: Take row sums for each gene in count matrix, keep genes expressed in 1% or more of cells
```{r eval=FALSE}
# Get counts slot from Seurat object
counts_ald <- GetAssayData(object = ald, slot = "counts")
counts_sepp <- GetAssayData(object = sepp, slot = "counts")

# If a count for a gene in a cell is greater than 0, set as TRUE (= 1)
nonzero_ald <- counts_ald > 0
nonzero_sepp <- counts_sepp > 0

# If 1% or more cells are TRUE, keep the gene. Each TRUE value = 1. Taking the sum of all the cells for that gene
keep_ald_genes <- Matrix::rowSums(nonzero_ald) >= (0.01*length(Cells(ald)))
keep_sepp_genes <- Matrix::rowSums(nonzero_sepp) >= (0.01*length(Cells(sepp)))

# Get the genes names that we are keeping
ald_genes <- rownames(ald)[keep_ald_genes]
sepp_genes <- rownames(sepp)[keep_sepp_genes]
```

Step 4: Take intersecting genes between the datasets
```{r eval=FALSE}
intersecting_genes <- intersect(ald_genes, sepp_genes)  
```

```{r eval=FALSE}
# Note that the number of genes in the Aldinger and Sepp datasets are the same as they were merged together before
> print(length(rownames(ald)))
[1] 52672

> print(length(rownames(sepp)))
[1] 52672

> print(length(ald_genes))
[1] 13796

> print(length(sepp_genes))
[1] 14376

> print(length(intersecting_genes))
[1] 10604
```

Step 5: Filter each dataset for intersecting genes
```{r eval=FALSE}
ald <- subset(ald, features = intersecting_genes)
sepp <- subset(sepp, features = intersecting_genes)
```

Step 6: Run SCTransform on each dataset, regressing out (CC.Difference)
```{r eval=FALSE}
ald <- SCTransform(ald, vars.to.regress = "CC.Difference", variable.features.n = 5000)
sepp <- SCTransform(sepp, vars.to.regress = "CC.Difference", variable.features.n = 5000)
```

Step 7: Merge the datasets back together
```{r eval=FALSE}
rl_merged <- merge(x = ald, y = sepp)
```

Step 8: Run PrepSCTFindMarkers()
```{r eval=FALSE}
rl_merged <- PrepSCTFindMarkers(object = rl_merged)
```

Step 9: Run FindConservedMarkers()
```{r eval=FALSE}
# Releveling based on alphabetical then numeric order
levels_index <- str_order(
  unique(rl_merged$broad_cell_type),
  numeric = TRUE
)
rl_merged$broad_cell_type <- fct_relevel(rl_merged$broad_cell_type, levels(rl_merged$broad_cell_type)[levels_index])

# Setting the Idents to be the cluster name because FindMarkers() uses the Ident
Idents(rl_merged) <- "broad_cell_type"

# Creating empty list to store markers
rl_marker_list <- list()

# Find markers for each cluster
for (cluster in levels(Idents(rl_merged))) {
  rl_markers <- FindConservedMarkers(
    object = rl_merged,
    ident.1 = cluster,
    grouping.var = "dataset_name",
    logfc.threshold = 0,
    min.pct = 0.01,
    only.pos = FALSE,
    assay = "SCT"
  )
  # Append markers to list
  rl_marker_list[[cluster]] <- rl_markers
}

# Creating a dataframe from the list of markers
df_rl <- rl_marker_list
for (name in names(df_rl)) {
  # Adding cluster name as a column
  df_rl[[name]]$cluster <- name
  # Adding gene name as a column
  df_rl[[name]]$feature <- rownames(df_rl[[name]])
}

# Stack into one df
combined_df <- do.call(bind_rows,df_rl) %>%
  filter(
    Aldinger_full_cerebellum_human_p_val_adj < 0.05, 
    Sepp_full_cerebellum_human_p_val_adj < 0.05, 
    Aldinger_full_cerebellum_human_avg_log2FC * Sepp_full_cerebellum_human_avg_log2FC > 0) %>%
  mutate(
    p_val_adj = pmax(Aldinger_full_cerebellum_human_p_val_adj, Sepp_full_cerebellum_human_p_val_adj),
    avg_log2FC = pmin(Aldinger_full_cerebellum_human_avg_log2FC, Sepp_full_cerebellum_human_avg_log2FC))

write.csv(combined_df, file = paste0(out_directory, "/sig_rl_markers.csv"), row.names = FALSE)
```
