---
title: Trying out singleCellNet
date: 2024-08-08
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
execute:
  eval: false
---

# Goal of this analysis

1. Try out singleCellNet

# Files and directories

## Input

CCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs`

Tumour file is found at `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240808`

# Process and Results

Using the `singleCellNet` conda environment

Lets load in the required libraries
```{r}
library(tidyverse)
library(Seurat)
library(qs)
library(singleCellNet)
```

```{r}
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240808"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
```

```{r}
rl_srat <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs")
tumour_srat <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds")
```

```{r}
# Subset out mice cells
rl_srat <- subset(x = rl_srat, subset = species == "human")
```

Take row sums for each gene in count matrix, keep genes expressed in 1% or more of cells
```{r}
# Get counts slot from Seurat object
counts_rl <- GetAssayData(object = rl_srat, slot = "counts")
counts_tumour <- GetAssayData(object = tumour_srat, slot = "counts")

# If a count for a gene in a cell is greater than 0, set as TRUE (= 1)
nonzero_rl <- counts_rl > 0
nonzero_tumour <- counts_tumour > 0

# If 1% or more cells are TRUE, keep the gene. Each TRUE value = 1. Taking the sum of all the cells for that gene
keep_rl_genes <- Matrix::rowSums(nonzero_rl) >= (0.01*length(Cells(rl_srat)))
keep_tumour_genes <- Matrix::rowSums(nonzero_tumour) >= (0.01*length(Cells(tumour_srat)))

# Get the genes names that we are keeping
rl_genes <- rownames(rl_srat)[keep_rl_genes]
tumour_genes <- rownames(tumour_srat)[keep_tumour_genes]
```

Take intersecting genes between the datasets
```{r}
intersecting_genes <- intersect(rl_genes, tumour_genes)  
```

```{r}
> print(length(rownames(rl_srat)))
[1] 52672

> print(length(rownames(tumour_srat)))
[1] 21624

> print(length(rl_genes))
[1] 15128

> print(length(tumour_genes))
[1] 13370

> print(length(intersecting_genes))
[1] 11077
```

Filter each dataset for intersecting genes
```{r}
rl_srat <- subset(rl_srat, features = intersecting_genes)
tumour_srat <- subset(tumour_srat, features = intersecting_genes)
```

Normalize Seurat object
```{r}
# Normalize RNA assay before converting
rl_srat <- NormalizeData(rl_srat, assay = "RNA")
tumour_srat  <- NormalizeData(tumour_srat, assay = "RNA")
```

