---
title: Using FindConservedMarkers() to determine UBC markers using the RNA assay instead
date: 2024-10-18
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
engine: knitr
execute:
  eval: false
---

# Goal of this analysis

1. Find markers with the FindConservedMarkers() workflow but using the RNA assay instead of the SCTransform assay

# Files and directories

## Input

CCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20241018`

# Process and Results

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
library(RColorBrewer)
library(patchwork)
library(ggrepel)
```

Loading in the Seurat object
```{r eval=FALSE}
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20241018"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
rl_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs")
```

## Marker determination

Step 1: Filter RL lineage for human UBC cells
```{r eval=FALSE}
ubc_cells <- rl_cca %>%
  subset(subset = species == "human") %>%
  subset(subset = broad_cell_type == "UBC") 
```

Step 2: Separate the UBC cells by dataset
```{r eval=FALSE}
ald <- subset(ubc_cells, subset = dataset_name == "Aldinger_full_cerebellum_human")
sepp <- subset(ubc_cells, subset = dataset_name == "Sepp_full_cerebellum_human")
```

Step 3: Take row sums for each gene in count matrix, keep genes expressed in 1% or more of cells
```{r eval=FALSE}
# Get counts slot from Seurat object
counts_ald <- GetAssayData(object = ald, slot = "counts")
counts_sepp <- GetAssayData(object = sepp, slot = "counts")

# If a count for a gene in a cell is greater than 0, set as TRUE (= 1)
nonzero_ald <- counts_ald > 0
nonzero_sepp <- counts_sepp > 0

# If 1% or more cells are TRUE, keep the gene. Each TRUE value = 1. Taking the sum of all the cells for that gene
keep_ald_genes <- Matrix::rowSums(nonzero_ald) >= (0.01*length(Cells(ald)))
keep_sepp_genes <- Matrix::rowSums(nonzero_sepp) >= (0.01*length(Cells(sepp)))

# Get the genes names that we are keeping
ald_genes <- rownames(ald)[keep_ald_genes]
sepp_genes <- rownames(sepp)[keep_sepp_genes]
```

Step 4: Take intersecting genes between the datasets
```{r eval=FALSE}
intersecting_genes <- intersect(ald_genes, sepp_genes)  
```

```{r eval=FALSE}
# Note that the number of genes in the Aldinger and Sepp datasets are the same as they were merged together before
> print(length(rownames(ald)))
[1] 52672

> print(length(rownames(sepp)))
[1] 52672

> print(length(ald_genes))
[1] 9882

> print(length(sepp_genes))
[1] 12658

> print(length(intersecting_genes))
[1] 8122
```

Step 5: Filter each dataset for intersecting genes
```{r eval=FALSE}
ald <- subset(ald, features = intersecting_genes)
sepp <- subset(sepp, features = intersecting_genes)
```

Step 6: Run NormalizeData() on each dataset, then FindVariableFeatures(), then ScaleData() regressing out (CC.Difference)
```{r eval=FALSE}
ald <- NormalizeData(ald, assay = "RNA") %>%
  FindVariableFeatures() %>%
  ScaleData(vars.to.regress = "CC.Difference")

sepp <- NormalizeData(sepp, assay = "RNA") %>%
  FindVariableFeatures() %>%
  ScaleData(vars.to.regress = "CC.Difference")
```

Step 7: Merge the datasets back together
```{r eval=FALSE}
ubc_merged <- merge(x = ald, y = sepp)
```

Step 8: Run FindConservedMarkers()
```{r eval=FALSE}
# Releveling based on alphabetical then numeric order
levels_index <- str_order(
  unique(ubc_merged$new_clusters),
  numeric = TRUE
)
ubc_merged$new_clusters <- fct_relevel(ubc_merged$new_clusters, levels(ubc_merged$new_clusters)[levels_index])

# Setting the Idents to be the cluster name because FindConservedMarkers() uses the Ident
Idents(ubc_merged) <- "new_clusters"

# Creating empty list to store markers
ubc_marker_list <- list()

# Find markers for each cluster
for (cluster in levels(Idents(ubc_merged))) {
  ubc_markers <- FindConservedMarkers(
    object = ubc_merged,
    ident.1 = cluster,
    grouping.var = "dataset_name",
    logfc.threshold = 0,
    min.pct = 0.01,
    only.pos = FALSE,
    assay = "RNA"
  )
  # Append markers to list
  ubc_marker_list[[cluster]] <- ubc_markers
}

# Creating a dataframe from the list of markers
df_ubc <- ubc_marker_list
for (name in names(df_ubc)) {
  # Adding cluster name as a column
  df_ubc[[name]]$cluster <- name
  # Adding gene name as a column
  df_ubc[[name]]$feature <- rownames(df_ubc[[name]])
}

# Stack into one df
combined_df <- do.call(rbind,df_ubc) %>%
  filter(
    # Filtering to make sure that a gene is significant in both datasets as well as log2FC is same direction
    Aldinger_full_cerebellum_human_p_val_adj < 0.05, 
    Sepp_full_cerebellum_human_p_val_adj < 0.05, 
    Aldinger_full_cerebellum_human_avg_log2FC * Sepp_full_cerebellum_human_avg_log2FC > 0) %>%
  mutate(
    # Adding p_val_adj and avg_log2FC column for pathway analysis script
    p_val_adj = pmax(Aldinger_full_cerebellum_human_p_val_adj, Sepp_full_cerebellum_human_p_val_adj),
    avg_log2FC = pmin(Aldinger_full_cerebellum_human_avg_log2FC, Sepp_full_cerebellum_human_avg_log2FC))

# Saving the background for each cluster for pathway analysis
background <- do.call(rbind,df_ubc) %>%
  select(feature, cluster) 

all_genes <- do.call(rbind,df_ubc)

write.csv(combined_df, file = paste0(out_directory, "/sig_ubc_markers.csv"), row.names = FALSE)

write.csv(background, file = paste0(out_directory, "/background.csv"), row.names = FALSE)

write.csv(all_genes, file = paste0(out_directory, "/all_tested_genes.csv"), row.names = FALSE)
```

I have the markers for each UBC cluster stored in the `ubc_marker_list` variable. Now I can create the necessary plots.

## Volcano Plots

```{r}
ald_markers <- all_genes %>%
  select(
    p_val = Aldinger_full_cerebellum_human_p_val, 
    avg_log2FC = Aldinger_full_cerebellum_human_avg_log2FC, 
    pct.1 = Aldinger_full_cerebellum_human_pct.1,
    pct.2 = Aldinger_full_cerebellum_human_pct.2,
    p_val_adj = Aldinger_full_cerebellum_human_p_val_adj,
    cluster,
    feature) %>%
  split(.$cluster)

sepp_markers <- all_genes %>%
  select(
    p_val = Sepp_full_cerebellum_human_p_val, 
    avg_log2FC = Sepp_full_cerebellum_human_avg_log2FC, 
    pct.1 = Sepp_full_cerebellum_human_pct.1,
    pct.2 = Sepp_full_cerebellum_human_pct.2,
    p_val_adj = Sepp_full_cerebellum_human_p_val_adj,
    cluster,
    feature) %>%
  split(.$cluster)
```

Aldinger genes
```{r}
source("/u/llau/software/mb_scrnaseq/MB_scRNAseq/software/utilities/plotting.R")

for(cluster in names(ald_markers)){
    variable_markers <- ald_markers[[cluster]]

    # Setting log2FC threshold
    log_fc_thresh <- 0.5
    
    # Adding column to determine if gene is up regulated or down regulated
    variable_markers <- mutate(
      variable_markers,
      direction = case_when(
        variable_markers$avg_log2FC >= log_fc_thresh & variable_markers$p_val_adj < 0.05 ~ "Upregulated",
        variable_markers$avg_log2FC <= -log_fc_thresh & variable_markers$p_val_adj < 0.05 ~ "Downregulated",
        TRUE ~ "Not significant"
        )
      )
    variable_markers$direction <- factor(variable_markers$direction, levels = c("Upregulated", "Downregulated", "Not significant"))

    # Adding column denotating genes to label
    variable_markers$feature_labelled <- ifelse(abs(variable_markers$avg_log2FC) > log_fc_thresh & variable_markers$p_val_adj < 0.05, variable_markers$feature, NA)

    # Dealing with padjval = 0. -log10(0) = infinity therefore we need to deal with the 0 value by setting it as something small
    non_zero <- variable_markers$p_val_adj[variable_markers$p_val_adj != 0]
    # Find the smallest value
    min_val <- min(non_zero)
    # Multiply the smallest value by 0.1 and assign to p_val_adj values
    variable_markers$p_val_adj <- ifelse(variable_markers$p_val_adj == 0, min_val*exp(-1), variable_markers$p_val_adj)
    
    n_up <- sum(variable_markers$avg_log2FC > log_fc_thresh & variable_markers$p_val_adj < 0.05)
    n_down <- sum(variable_markers$avg_log2FC < -log_fc_thresh & variable_markers$p_val_adj < 0.05)
    
    gene_numbers <- paste0(
      "n = ", nrow(variable_markers),
      "; down = ", n_down,
      "; up = ", n_up
    )

    # Creating the volcano plot
    plt <- make_volcano(
      data = variable_markers,
      log_fc = avg_log2FC,
      log_pval = -log10(p_val_adj),
      direction = direction,
      #gene_labels = feature_labelled,
      log_fc_thresh = c(-log_fc_thresh, log_fc_thresh),
      log_pval_thresh = -log10(0.05),
      label_num_genes = TRUE
    ) + 
    theme_classic() +
    labs(x = expression(log[2]~"(fold change)"), y = expression(-log[10]~"(adj. p-value)")) +
    scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey")) +
    ggtitle(cluster) +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_text_repel(aes(label = feature_labelled), size = 2, show.legend = FALSE) +
    labs(subtitle = gene_numbers)

    volcano_name <- paste0("aldinger_", cluster, "_volcano_plot.png")
    ggsave(filename = volcano_name, plot = plt, path = out_directory, 
            width = 10, height = 10, units = "in", dpi = 600)
}
```

Sepp genes
```{r}
for(cluster in names(sepp_markers)){
    variable_markers <- sepp_markers[[cluster]]

    # Setting log2FC threshold
    log_fc_thresh <- 0.5
    
    # Adding column to determine if gene is up regulated or down regulated
    variable_markers <- mutate(
      variable_markers,
      direction = case_when(
        variable_markers$avg_log2FC >= log_fc_thresh & variable_markers$p_val_adj < 0.05 ~ "Upregulated",
        variable_markers$avg_log2FC <= -log_fc_thresh & variable_markers$p_val_adj < 0.05 ~ "Downregulated",
        TRUE ~ "Not significant"
        )
      )
    variable_markers$direction <- factor(variable_markers$direction, levels = c("Upregulated", "Downregulated", "Not significant"))

    # Adding column denotating genes to label
    variable_markers$feature_labelled <- ifelse(abs(variable_markers$avg_log2FC) > log_fc_thresh & variable_markers$p_val_adj < 0.05, variable_markers$feature, NA)

    # Dealing with padjval = 0. -log10(0) = infinity therefore we need to deal with the 0 value by setting it as something small
    non_zero <- variable_markers$p_val_adj[variable_markers$p_val_adj != 0]
    # Find the smallest value
    min_val <- min(non_zero)
    # Multiply the smallest value by 0.1 and assign to p_val_adj values
    variable_markers$p_val_adj <- ifelse(variable_markers$p_val_adj == 0, min_val*exp(-1), variable_markers$p_val_adj)
    
    n_up <- sum(variable_markers$avg_log2FC > log_fc_thresh & variable_markers$p_val_adj < 0.05)
    n_down <- sum(variable_markers$avg_log2FC < -log_fc_thresh & variable_markers$p_val_adj < 0.05)
    
    gene_numbers <- paste0(
      "n = ", nrow(variable_markers),
      "; down = ", n_down,
      "; up = ", n_up
    )

    # Creating the volcano plot
    plt <- make_volcano(
      data = variable_markers,
      log_fc = avg_log2FC,
      log_pval = -log10(p_val_adj),
      direction = direction,
      #gene_labels = feature_labelled,
      log_fc_thresh = c(-log_fc_thresh, log_fc_thresh),
      log_pval_thresh = -log10(0.05),
      label_num_genes = TRUE
    ) + 
    theme_classic() +
    labs(x = expression(log[2]~"(fold change)"), y = expression(-log[10]~"(adj. p-value)")) +
    scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey")) +
    ggtitle(cluster) +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_text_repel(aes(label = feature_labelled), size = 2, show.legend = FALSE) +
    labs(subtitle = gene_numbers)

    volcano_name <- paste0("sepp_", cluster, "_volcano_plot.png")
    ggsave(filename = volcano_name, plot = plt, path = out_directory, 
            width = 10, height = 10, units = "in", dpi = 600)
}
```