---
title: Results for Ian's committee meeting
date: 2024-07-18
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

# Goal of this analysis

1. For human and mouse cells, proportion of cells at each time point
2. Run scProportionTest on all clusters with UBC subclusters in one cluster
3. Run Wilcoxon test on proportions from propeller test

# Files and directories

## Input

CCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs`

RPCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240613/rpca/combined_clusters_rl.qs`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240718`

# Process and Results

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
library(scProportionTest)
library(RColorBrewer)
```

```{r eval=FALSE}
# Setting seed for scproportiontest
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240718"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
# Loading in Seurat objects
rl_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs")
rl_rpca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240613/rpca/combined_clusters_rl.qs")
```

## Proportion of Cells at each Time Point

Filtering out non-RL lineage cells (endothelial, microglia, OPC)
```{r eval=FALSE}
rl_only_cca <- rl_cca
Idents(rl_only_cca) <- "new_clusters"
rl_only_cca <- subset(x = rl_only_cca, idents = c("Cluster_11", "Cluster_13", "Cluster_14", "Cluster_15"), invert = TRUE)
```

Plotting the age proportions per species
```{r eval=FALSE}
# Refactoring age column
rl_only_cca$combined_age <- factor(rl_only_cca$combined_age, levels = unique(rl_only_cca$combined_age))
levels_index <- str_order(
  levels(rl_only_cca$combined_age),
  numeric = TRUE
)
rl_only_cca$combined_age <- fct_relevel(rl_only_cca$combined_age, levels(rl_only_cca$combined_age)[levels_index])

source("/u/llau/software/mb_scrnaseq/MB_scRNAseq/software/utilities/cluster_barplot.R")

#Creating pallete
palette1 <- brewer.pal(12, "Set3") 
palette2 <- brewer.pal(8, "Pastel2") 
palette3 <- brewer.pal(12, "Paired") 
combined_palette <- c(palette1, palette2, palette3)

plt <- cluster_barplot(rl_only_cca, split.by = "combined_age", group.by = "species", position = "fill") + ggtitle("Ages Per Species in Rhombic Lip Lineage") + scale_fill_manual(values = combined_palette)

bar_plot_name <- "rl_age_bar_plot.png"

ggsave(filename = bar_plot_name, plot = plt, path = out_directory, 
          width = 10, height = 10, units = "in", dpi = 600)
```

## scProportion Test

### CCA

Proportion test for all RL-lineage clusters with all timepoints
```{r eval=FALSE}
# Converting seurat object to scproportion object 
rl_test <- sc_utils(rl_only_cca)

# Running proportion test
rl_test <- permutation_test(
	rl_test, cluster_identity = "snn_res.0.4",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

# Plotting proportion test results
plt <- permutation_plot(rl_test)
filename <- "proportion_test/cca/cca_rl_clusters_all_timepoints_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

# Saving the permutation statistics
write.csv(
  x = rl_test@results$permutation,
  file = file.path(out_directory, "proportion_test/cca/cca_rl_clusters_all_timepoints_proportion_test.csv"),
  row.names = FALSE
)
```

Filtering out early mouse ages (E10 - E12.5) and latest mouse age (P14) and rerunning proportion tests
```{r eval=FALSE}
Idents(rl_only_cca) <- "combined_age"
rl_only_cca <- subset(x = rl_only_cca, idents = c("E10", "E10.5", "E11.5", "E12", "E12.5", "P14"), invert = TRUE)
```

Proportion test for all RL-lineage clusters
```{r eval=FALSE}
rl_test <- sc_utils(rl_only_cca)

rl_test <- permutation_test(
	rl_test, cluster_identity = "snn_res.0.4",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(rl_test)
filename <- "proportion_test/cca/cca_rl_clusters_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

write.csv(
  x = rl_test@results$permutation,
  file = file.path(out_directory, "proportion_test/cca/cca_rl_clusters_proportion_test.csv"),
  row.names = FALSE
)
```

Proportion test for all RL-lineage clusters with the UBC cluster split into the UBC subclusters
```{r eval=FALSE}
rl_test <- sc_utils(rl_only_cca)

rl_test <- permutation_test(
	rl_test, cluster_identity = "new_clusters",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(rl_test)
filename <- "proportion_test/cca/cca_rl_ubc_clusters_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

write.csv(
  x = rl_test@results$permutation,
  file = file.path(out_directory, "proportion_test/cca/cca_rl_ubc_clusters_proportion_test.csv"),
  row.names = FALSE
)
```

Proportion test for only UBC subclusters
```{r eval=FALSE}
# Filtering for UBC subcluster cells only
Idents(rl_only_cca) <- "new_clusters"
ubc_cells <- subset(rl_only_cca, idents = c("UBC_0", "UBC_1", "UBC_2", "UBC_3", "UBC_4", "UBC_5"))
```
```{r eval=FALSE}
ubc_test <- sc_utils(ubc_cells)

ubc_test <- permutation_test(
	ubc_test, cluster_identity = "new_clusters",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(ubc_test) + 
  theme(
    legend.position = c(0.27, 0.1),
    legend.text = element_text(size = 7),  # Adjust legend text size
    legend.title = element_text(size = 9),  # Adjust legend title size
    legend.key.size = unit(0.1, "in")  # Adjust legend key size
  ) + 
  labs(x = NULL, y = expression("Observed"~log[2]~"fold difference"))

filename <- "proportion_test/cca/cca_ubc_clusters_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 5, height = 5, dpi = 600)

write.csv(
  x = ubc_test@results$permutation,
  file = file.path(out_directory, "proportion_test/cca/cca_ubc_clusters_proportion_test.csv"),
  row.names = FALSE
)
```

### RPCA

Before we run the proportion test, we need to subset out the control clusters (endothelial, microglia, OPC). For RPCA, they are Cluster 8, 10, 11, 12
```{r eval=FALSE}
rl_only_rpca <- rl_rpca
Idents(rl_only_rpca) <- "new_clusters"
rl_only_rpca <- subset(x = rl_only_rpca, idents = c("Cluster_8", "Cluster_10", "Cluster_11", "Cluster_12"), invert = TRUE)
```

Proportion test for all RL-lineage clusters with all timepoints
```{r eval=FALSE}
rl_test <- sc_utils(rl_only_rpca)

rl_test <- permutation_test(
	rl_test, cluster_identity = "snn_res.0.2",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(rl_test)
filename <- "proportion_test/rpca/rpca_rl_clusters_all_timepoints_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

write.csv(
  x = rl_test@results$permutation,
  file = file.path(out_directory, "proportion_test/rpca/rpca_rl_clusters_all_timepoints_proportion_test.csv"),
  row.names = FALSE
)
```

Filtering out early mouse ages (E10 - E12.5) and latest mouse age (P14) and rerunning proportion tests
```{r eval=FALSE}
Idents(rl_only_rpca) <- "combined_age"
rl_only_rpca <- subset(x = rl_only_rpca, idents = c("E10", "E10.5", "E11.5", "E12", "E12.5", "P14"), invert = TRUE)
```

Proportion test for all RL-lineage clusters
```{r eval=FALSE}
rl_test <- sc_utils(rl_only_rpca)

rl_test <- permutation_test(
	rl_test, cluster_identity = "snn_res.0.2",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(rl_test)
filename <- "proportion_test/rpca/rpca_rl_clusters_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

write.csv(
  x = rl_test@results$permutation,
  file = file.path(out_directory, "proportion_test/rpca/rpca_rl_clusters_proportion_test.csv"),
  row.names = FALSE
)
```

Proportion test for all RL-lineage clusters with the UBC cluster split into the UBC subclusters
```{r eval=FALSE}
rl_test <- sc_utils(rl_only_rpca)

rl_test <- permutation_test(
	rl_test, cluster_identity = "new_clusters",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(rl_test)
filename <- "proportion_test/rpca/rpca_rl_ubc_clusters_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

write.csv(
  x = rl_test@results$permutation,
  file = file.path(out_directory, "proportion_test/rpca/rpca_rl_ubc_clusters_proportion_test.csv"),
  row.names = FALSE
)
```

Proportion test for only UBC subclusters
```{r eval=FALSE}
# Filtering for UBC subclusters only
Idents(rl_only_rpca) <- "new_clusters"
ubc_cells <- subset(rl_only_rpca, idents = c("UBC_0", "UBC_1", "UBC_2", "UBC_3", "UBC_4", "UBC_5", "UBC_6"))
```

```{r eval=FALSE}
ubc_test <- sc_utils(ubc_cells)

ubc_test <- permutation_test(
	ubc_test, cluster_identity = "new_clusters",
	sample_1 = "mouse", sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

plt <- permutation_plot(ubc_test)
filename <- "proportion_test/rpca/rpca_ubc_clusters_permutation_test.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 10, dpi = 600)

write.csv(
  x = ubc_test@results$permutation,
  file = file.path(out_directory, "proportion_test/rpca/rpca_ubc_clusters_proportion_test.csv"),
  row.names = FALSE
)
```

## Wilcoxon test on proportions

For this portion we will need to switch environments to stay consistant

Using the `propeller` conda environment

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
```

Setting the seed and creating the output directory
```{r eval=FALSE}
set.seed(1)

out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240718/wilcox_proportion"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
```

Creating functions to streamline the process
```{r eval=FALSE}
# Refactors a given column based on numerical order
level_sort <- function(column) {
  column <- as.factor(as.character(column))
  column <- fct_relevel(column, str_sort(levels(column), numeric = TRUE))
  return(column)
}

# Converts a dataframe in a list of lists
wilcox_list <- function(x) {
  # Split dataframe by species
  species_df <- split(x, x$species)
  # For each species, split by cluster
  clusters_list <- lapply(species_df, function(sub_df) {
    split(sub_df, sub_df$clusters)
  })
  # Returns list$species$cluster
  return(clusters_list)
}

# Runs wilcox test on species proportions 
wilcox_test <- function(mouse, human) {
  wilcox_results <- wilcox.test(mouse$proportion, human$proportion, alternative = "two.sided", conf.int = TRUE)
  return(wilcox_results)
}

# wilcox.test returns a list with class 'htest' with components. Need to parse components in order to get p-val
get_p_val <- function(wilcox_results) {
  # Parses the wilcox results for p.value and makes it into a dataframe
  p_values <- map_dbl(wilcox_results, ~ .x$p.value) %>%
    data.frame(p_val = .) 
  # The cluster names are set as the rownames, need to add back as a column
  p_values$clusters <- rownames(p_values)
  return(p_values)
}
```

Function taken from `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/propeller_helpers.R`

Needed to make minor adjustments to the function
```{r eval=FALSE}
#' Make boxplot showing proportion of cells in each cluster.
#'
#' @param props The output of `get_props_per_sample` (a dataframe with the
#'   proportion of cells in each cluster/cell type for each sample).
#' @param facet_var Name of variable (passed to `facet_wrap`).
#' @param facet_ncol Number of columns (passed to `facet_wrap`).
#' @param fdrs The output of `get_fdrs_for_plotting` (a data frame with the FDRs
#'   for each cluster).
#' @param fdr_y The y-value at which to plot the FDR values. If `NULL`
#'   (default), the y-values will be calculated automatically for each facet.
#'
#' @return A `ggplot2` object.
#'
boxplot_props_per_cluster <- function(
  props,
  facet_var,
  facet_ncol = 4,
  fdrs = NULL,
  fdr_y = NULL
) {
  # 
  .plt <- ggplot(props, aes(x = species, y = proportion)) + 
    geom_boxplot(width = 0.6, outlier.shape = NA) + 
    geom_jitter(aes(shape = dataset_name), width = 0.2)
  
  if (!is.null(fdrs)) {
    # where to show the FDR value (how high on the plot)
    if (is.null(fdr_y)) {
      fdrs <- props %>%
        group_by(clusters) %>%
        summarise(fdr_y = max(proportion) * 1.05) %>%
        full_join(
          x = fdrs,
          y = .,
          by = join_by(clusters)
        )
    }

    # show FDR value on plot
    .plt <- .plt + 
      geom_text(
        data = fdrs,
        mapping = aes(x = "mouse", y = fdr_y, label = sprintf("p_val = %.3f", p_val)),
        nudge_x = 0.5,
        inherit.aes = FALSE
      )
  }

  .plt <- .plt + 
    facet_wrap(
      facets = vars({{facet_var}}),
      ncol = facet_ncol,
      scales = "free"
    ) + 
    labs(y = "proportion of all RL lineage cells") + 
    theme_light()
  
  return(.plt)
}
```

Running workflow on RL clusters (UBCs in one cluster)
```{r eval=FALSE}
# Load in proportions
rl_clusters <- read.csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240528/all_timepoints_proportions.csv")
# Factor $species as the plot required the mouse to be first and then the human
rl_clusters$species <- factor(rl_clusters$species, levels = c("mouse", "human"))
# Factor $cluster column numerically
rl_clusters$clusters <- level_sort(rl_clusters$clusters)

# Transform df into list
rl_clusters_list <- wilcox_list(rl_clusters)

# For applies wilcox_test() for each corresponding cluster in list
wilcox_results_rl_clusters <- map2(
  rl_clusters_list$mouse,
  rl_clusters_list$human,
  wilcox_test
)

# Get p_val for each cluster
p_val_rl_clusters <- get_p_val(wilcox_results_rl_clusters)
write.csv(p_val_rl_clusters, file = paste0(out_directory, "/pval_rl_cluster.csv"), row.names = FALSE)

# Factor $cluster column numerically
p_val_rl_clusters$clusters <- level_sort(p_val_rl_clusters$clusters)

# Plot proportions with p_val calculated with wilcox.test()
plt <- boxplot_props_per_cluster(
  props = rl_clusters,
  facet_var = clusters,
  fdrs = p_val_rl_clusters 
) + 
  labs(caption = "(CCA integration)") + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(7/8, 1/12)
  )

filename <- "boxplot_props_wilcox_rl_clusters.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 30, dpi = 600)
```

Running workflow on RL clusters + UBC subclusters
```{r eval=FALSE}
# Load in proportions
rl_ubc_clusters <- read.csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240617/cca_proportions.csv")
# Factor $species as the plot required the mouse to be first and then the human
rl_ubc_clusters$species <- factor(rl_ubc_clusters$species, levels = c("mouse", "human"))
# Factor $cluster column numerically
rl_ubc_clusters$clusters <- level_sort(rl_ubc_clusters$clusters)

# Transform df into list
rl_ubc_clusters_list <- wilcox_list(rl_ubc_clusters)

# For applies wilcox_test() for each corresponding cluster in list
wilcox_results_rl_ubc_clusters <- map2(
  rl_ubc_clusters_list$mouse,
  rl_ubc_clusters_list$human,
  wilcox_test
)

# Get p_val for each cluster
p_val_rl_ubc_clusters <- get_p_val(wilcox_results_rl_ubc_clusters)
write.csv(p_val_rl_ubc_clusters, file = paste0(out_directory, "/pval_rl_ubc_cluster.csv"), row.names = FALSE)
# Factor $cluster column numerically
p_val_rl_ubc_clusters$clusters <- level_sort(p_val_rl_ubc_clusters$clusters)

# Plot proportions with p_val calculated with wilcox.test()
plt <- boxplot_props_per_cluster(
  props = rl_ubc_clusters,
  facet_var = clusters,
  fdrs = p_val_rl_ubc_clusters 
) + 
  labs(caption = "(CCA integration)") + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(7/8, 1/12)
  )

filename <- "boxplot_props_wilcox_rl_ubc_clusters.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 30, dpi = 600)
```

Running workflow on just UBC subclusters
```{r eval=FALSE}
# Load in proportions
ubc_clusters <- read.csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240606/four_dataset_cca_ubc_proportions.csv")
# Factor $species as the plot required the mouse to be first and then the human
ubc_clusters$species <- factor(ubc_clusters$species, levels = c("mouse", "human"))
# Factor $cluster column numerically
ubc_clusters$clusters <- level_sort(ubc_clusters$clusters)

# Transform df into list
ubc_clusters_list <- wilcox_list(ubc_clusters)

# For applies wilcox_test() for each corresponding cluster in list
wilcox_results_ubc_clusters <- map2(
  ubc_clusters_list$mouse,
  ubc_clusters_list$human,
  wilcox_test
)

# Get p_val for each cluster
p_val_ubc_clusters <- get_p_val(wilcox_results_ubc_clusters)
write.csv(p_val_ubc_clusters, file = paste0(out_directory, "/pval_ubc_cluster.csv"), row.names = FALSE)
# Factor $cluster column numerically
p_val_ubc_clusters$clusters <- level_sort(p_val_ubc_clusters$clusters)

# Plot proportions with p_val calculated with wilcox.test()
plt <- boxplot_props_per_cluster(
  props = ubc_clusters,
  facet_var = clusters,
  fdrs = p_val_ubc_clusters 
) + 
  labs(caption = "(CCA integration)") + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(7/8, 1/12)
  )

filename <- "boxplot_props_wilcox_ubc_clusters.png"
ggsave(filename = filename, plot = plt, path = out_directory, width = 10, height = 30, dpi = 600)
```