---
title: Determining the best DE method for our dataset
date: 2024-07-12
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

# Goal of this analysis

1. Read **[Benchmarking integration of single-cell differential expression](https://www.nature.com/articles/s41467-023-37126-3)** 

2. Determine their metrics for classifying a dataset

3. Attempt recommended methods

4. Check Sepp and Aldinger datasets to see if they have same gene symbols

# Files and directories

## Input

CCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs`

Aldinger dataset can be found at `/.mounts/labs/pailab/private/llau/data/Aldinger_2021/seurat.rds`

Sepp dataset can be found at `/.mounts/labs/pailab/private/llau/data/Sepp_2024/sepp2024_human_sct.rds`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240712`

# Process and Results

For their 46 workflows, they had 3 integrative stategies

1. Batch Effect Correction (Use of batch-effect correction algorithms)
2. Covariate Models (Statistical model with batch covariate model)
3. Meta Analysis (DE analysis performed for each batch, resulting statistics combined for each gene)

They provided a [table](https://www.nature.com/articles/s41467-023-37126-3/tables/1) that documents what method to use in what scenario.

In this table, we have to determine for our dataset the:

1. Sparsity
2. Depth
3. Batch effects 

Lets investigate these metrics within our dataset

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
library(RColorBrewer)
```

Loading in the Seurat object
```{r eval=FALSE}
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240712"
rl_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs")
```

```{r eval=FALSE}
# Filtering for human UBCs
ubc_cells <- rl_cca %>%
  subset(subset = species == "human") %>%
  subset(subset = broad_cell_type == "UBC") 

# Subsetting by dataset
ald <- subset(ubc_cells, subset = dataset_name == "Aldinger_full_cerebellum_human")
sepp <- subset(ubc_cells, subset = dataset_name == "Sepp_full_cerebellum_human")

# Get counts slot from Seurat object
counts_ubc <- GetAssayData(object = ubc_cells, slot = "counts")
counts_ald <- GetAssayData(object = ald, slot = "counts")
counts_sepp <- GetAssayData(object = sepp, slot = "counts")

# If a count for a gene in a cell is greater than 0, set as TRUE (= 1)
nonzero_ubc <- counts_ubc > 0
nonzero_ald <- counts_ald > 0
nonzero_sepp <- counts_sepp > 0

# Note in the paper they filter for sparsely expressed genes (zero rate > 0.95)
# Here it is saying that keep gene if it is expressed in more than 5% of cells
keep_ubc_genes <- Matrix::rowSums(nonzero_ubc) >= (length(Cells(ubc_cells)) * 0.05)
keep_ald_genes <- Matrix::rowSums(nonzero_ald) >= (length(Cells(ald)) * 0.05)
keep_sepp_genes <- Matrix::rowSums(nonzero_sepp) >= (length(Cells(sepp)) * 0.05)

# Get the genes names that we are keeping
ubc_genes <- rownames(ubc_cells)[keep_ubc_genes]
ald_genes <- rownames(ald)[keep_ald_genes]
sepp_genes <- rownames(sepp)[keep_sepp_genes]

# Determine intersecting genes
intersecting_genes <- intersect(ald_genes, sepp_genes)  
```
```{r eval=FALSE}
# Note that the number of genes in the Aldinger and Sepp datasets are the same as they were merged together before
> print(length(rownames(ald)))
[1] 52672

> print(length(rownames(sepp)))
[1] 52672

> print(length(ald_genes))
[1] 4219

> print(length(sepp_genes))
[1] 5291

> print(length(intersecting_genes))
[1] 3314

> print(length(ubc_genes))
[1] 4836
```
```{r eval=FALSE}
# Subset both datasets for intersecting genes
ald <- subset(ald, features = intersecting_genes)
sepp <- subset(sepp, features = intersecting_genes)
ubc <- subset(ubc_cells, features = ubc_genes)
```

## Sparsity

Sparsity is the measure of zero values in the matrix relative to the total number of elements in the matrix.
```{r eval=FALSE}
sparsity <- function(matrix) {
  # Count zero elements
  zero_count <- sum(matrix == 0)

  # Total elements in the matrix
  total_elements <- length(matrix)

  # Compute sparsity
  sparsity <- zero_count / total_elements

  return(sparsity)
}
```
```{r eval=FALSE}
ubc_matrix <- as.matrix(GetAssayData(object = ubc, slot = "counts"))
ald_matrix <- as.matrix(GetAssayData(object = ald, slot = "counts"))
sepp_matrix <- as.matrix(GetAssayData(object = sepp, slot = "counts"))

ubc_sparsity <- sparsity(ubc_matrix)
ald_sparsity <- sparsity(ald_matrix)
sepp_sparsity <- sparsity(sepp_matrix)
```
```{r eval=FALSE}
> ubc_sparsity
[1] 0.8614414

> ald_sparsity
[1] 0.8263131

> sepp_sparsity
[1] 0.8431703
```

Overall, the sparsity hovers at around **84%** . For the benchmarking table, we will use the 80% value category.

## Depth

Depth is the average nonzero count in each cell after filtering sparsely expressed genes
```{r eval=FALSE}
depth <- function(x) {
  # Calculate the depth
  depth <- mean((x[x > 0]))
  return(depth)
}
```
```{r eval=FALSE}
ubc_depth <- depth(ubc_matrix)
ald_depth <- depth(ald_matrix)
sepp_depth <- depth(sepp_matrix)
```
```{r eval=FALSE}
> ubc_depth
[1] 1.619809

> ald_depth
[1] 1.814789

> sepp_depth
[1] 1.581016
```

The average depth is around **1.6**. Lets take the closest value of 4 in the table.

## Batch Effect

The whole reason we are investigating the best method for DE is because of the batch effects we saw between genes. For this category, we will select **substantial** for this column.

## Methods

This leaves us with 4 recommended workflows. LogN_FEM, limmatrend, Raw_Wilcox and RISC_QP.

We already impliment the Raw_Wilcox which should be the .
Lets take a look at LogN_FEM.

This appears to be a workflow of LogNormalize + FEM which stands for "fixed effects model" from the MetaDE library. The paper has [code](https://github.com/noobCoding/Benchmarking-integration-of-differential-expression/blob/main/code/DE-Workflow-Implementation/ref_script/META_FEM.R) showing their workflow.


## Checking gene names from datasets
```{r eval=FALSE}
ald_orig <- readRDS("/.mounts/labs/pailab/private/llau/data/Aldinger_2021/seurat.rds")
sepp_orig <- readRDS("/.mounts/labs/pailab/private/llau/data/Sepp_2024/sepp2024_human_sct.rds")

DefaultAssay(ald_orig) <- "RNA"
DefaultAssay(sepp_orig) <- "RNA"
```

```{r eval=FALSE}
# Determine intersecting genes
ald_genes <- rownames(ald_orig)
sepp_genes <- rownames(sepp_orig)
intersecting <- intersect(ald_genes, sepp_genes)  
```
```{r eval=FALSE}
> length(ald_genes)
[1] 49166

> length(sepp_genes)
[1] 22142

> length(intersecting)
[1] 17475
```
```{r eval=FALSE}
write.csv(ald_genes, file = paste0(out_directory, "/ald_genes.csv"), row.names = FALSE)
```