---
title: Feature Plots for UBC Cluster
date: 2024-06-12
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

1. Create feature plots for UBC subclusters 

## Files and directories

### Input

CCA UBC file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs`

RPCA UBC file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240606/rpca_20_analysis/ubc/ubc.qs`

### Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240612`

## Process and Results

Lets load in the required libraries
```{r eval=FALSE}
library(tidyverse)
library(Seurat)
library(qs)
```

### UBC CCA
```{r eval=FALSE}
ubc_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs")
```

```{r eval=FALSE}
setwd("/u/llau/software/mb_scrnaseq/MB_scRNAseq")
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240612/cca"

# Retrieve UMAP coordinates
umap_coords <- Embeddings(ubc_cca, "umap")

# Calculate minimum and maximum values for each dimension
min_umap <- apply(umap_coords, 2, min) # Min (x,y)
max_umap <- apply(umap_coords, 2, max) # Max (x,y)

min_x <- ceiling(min_umap[1])
min_y <- ceiling(min_umap[2])
max_x <- ceiling(max_umap[1])
max_y <- ceiling(max_umap[2])

buffer_x <- (max_x - min_x)*0.05
buffer_y <- (max_y - min_y)*0.05

min_x <- min_x - buffer_x
min_y <- min_y - buffer_y
max_x <- max_x + buffer_x
max_y <- max_y + buffer_y

```


```{r eval=FALSE}
# Genes of interest

genes <- read.csv("scrnaseq_Leo/utilities/cell_gene_mapping.csv")
genes <- genes$gene

DefaultAssay(ubc_cca) <- "RNA"

# Feature plots
for(gene in genes){
  feature_plot <- FeaturePlot(object = ubc_cca, features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                  ggtitle(paste("Integrated: ", gene)) +
                  xlim(min_x, max_x) +       # Set x-axis limits
                  ylim(min_y, max_y)         # Set y-axis limits

  feature_plot_human <- FeaturePlot(object = ubc_cca, cells = rownames(ubc_cca@meta.data[ubc_cca$species == "human", ]),  
                        features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                        ggtitle(paste("Human: ", gene)) +
                        xlim(min_x, max_x) +       # Set x-axis limits
                        ylim(min_y, max_y)         # Set y-axis limits

  feature_plot_mouse <- FeaturePlot(object = ubc_cca, cells = rownames(ubc_cca@meta.data[ubc_cca$species == "mouse", ]), 
                        features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                        ggtitle(paste("Mouse: ", gene)) +
                        xlim(min_x, max_x) +       # Set x-axis limits
                        ylim(min_y, max_y)         # Set y-axis limits

  plot <- feature_plot | feature_plot_human | feature_plot_mouse
  out_path <- paste0(out_directory, "/feature_plots")
  feature_plot_name <- paste0(gene, ".png")
  ggsave(filename = feature_plot_name, plot = plot, path = out_path, 
        width =30, height = 10, units = "in", dpi = 600)
}
```

### UBC RPCA

```{r eval=FALSE}
ubc_rpca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240606/rpca_20_analysis/ubc/ubc.qs")
```
```{r eval=FALSE}
setwd("/u/llau/software/mb_scrnaseq/MB_scRNAseq")
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240612/rpca"

# Retrieve UMAP coordinates
umap_coords <- Embeddings(ubc_rpca, "umap")

# Calculate minimum and maximum values for each dimension
min_umap <- apply(umap_coords, 2, min) # Min (x,y)
max_umap <- apply(umap_coords, 2, max) # Max (x,y)

min_x <- ceiling(min_umap[1])
min_y <- ceiling(min_umap[2])
max_x <- ceiling(max_umap[1])
max_y <- ceiling(max_umap[2])

buffer_x <- (max_x - min_x)*0.05
buffer_y <- (max_y - min_y)*0.05

min_x <- min_x - buffer_x
min_y <- min_y - buffer_y
max_x <- max_x + buffer_x
max_y <- max_y + buffer_y

```
```{r eval=FALSE}
# Genes of interest

genes <- read.csv("scrnaseq_Leo/utilities/cell_gene_mapping.csv")
genes <- genes$gene

DefaultAssay(ubc_rpca) <- "RNA"

# Feature plots
for(gene in genes){
  feature_plot <- FeaturePlot(object = ubc_rpca, features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                  ggtitle(paste("Integrated: ", gene)) +
                  xlim(min_x, max_x) +       # Set x-axis limits
                  ylim(min_y, max_y)         # Set y-axis limits

  feature_plot_human <- FeaturePlot(object = ubc_rpca, cells = rownames(ubc_rpca@meta.data[ubc_rpca$species == "human", ]),  
                        features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                        ggtitle(paste("Human: ", gene)) +
                        xlim(min_x, max_x) +       # Set x-axis limits
                        ylim(min_y, max_y)         # Set y-axis limits

  feature_plot_mouse <- FeaturePlot(object = ubc_rpca, cells = rownames(ubc_rpca@meta.data[ubc_rpca$species == "mouse", ]), 
                        features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                        ggtitle(paste("Mouse: ", gene)) +
                        xlim(min_x, max_x) +       # Set x-axis limits
                        ylim(min_y, max_y)         # Set y-axis limits

  plot <- feature_plot | feature_plot_human | feature_plot_mouse
  out_path <- paste0(out_directory, "/feature_plots")
  feature_plot_name <- paste0(gene, ".png")
  ggsave(filename = feature_plot_name, plot = plot, path = out_path, 
        width =30, height = 10, units = "in", dpi = 600)
}
```