---
title: Running DEA on Medulloblastoma samples
date: 2024-07-25
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
execute:
  eval: false
---

# Goal of this analysis

1. Make feature plots for genes of interest

# Files and directories

## Input

Tumour file is found at `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240725`

# Process and Results

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r}
library(tidyverse)
library(Seurat)
library(qs)
```

```{r}
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240725"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
```

```{r}
tumour_srat <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds")
```

```{r}
tumour_srat <- SCTransform(tumour_srat, vars.to.regress = "CC.Difference", variable.features.n = 5000)
```

```{r eval=FALSE}
DefaultAssay(tumour_srat) <- "mnn.reconstructed"
plt <- DimPlot(tumour_srat, reduction = "umap", label = TRUE, repel = TRUE, group.by = "seurat_clusters")
ggsave(filename = "tumour_srat_clusters_umap.png", plot = plt, path = out_directory, 
            width = 10, height = 10, units = "in", dpi = 600)
```

```{r}
source("/u/llau/software/mb_scrnaseq/MB_scRNAseq/software/utilities/cluster_barplot.R")

tumour_bar_plot <- cluster_barplot(tumour_srat, split.by = "subtype", group.by = "seurat_clusters")
bar_plot_name <- "tumour_cluster_bar_plot.png"
ggsave(filename = bar_plot_name, plot = tumour_bar_plot, path = out_directory, 
          width = 20, height = 10, units = "in", dpi = 600)

```

Looking at the cluster breakdowns, it appears that Clusters 3, 7, 9, 14 are pretty purely Group 4. Lets run a DEA for each of those clusters

```{r}
Idents(tumour_srat) <- "seurat_clusters"
DefaultAssay(tumour_srat) <- "SCT"
g4_clusters <- c("3", "7", "9", "14")

g4_markers <- list()
for(cluster in g4_clusters)
{
  markers <- list(cluster = FindMarkers(tumour_srat, ident.1 = cluster))
  g4_markers <- c(g4_markers, markers)
}
```

