---
title: Investigating genes of interest in UBC_1
date: 2024-07-23
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
execute:
  eval: false
---

# Goal of this analysis

1. Make feature plots for genes of interest

# Files and directories

## Input

CCA RL file is found at `/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs`

Tumour file is found at `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240723`

# Process and Results

Using the `scrnaseq_env` conda environment

Lets load in the required libraries
```{r}
library(tidyverse)
library(Seurat)
library(qs)
```

```{r}
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/integrated/20240723/UBC_1_genes"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
```

```{r}
rl_cca <- qread("/.mounts/labs/pailab/private/llau/results/integrated/20240618/rl_cca.qs")
tumour_srat <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20230510/mb_mnn.rds")
```

```{r eval=FALSE}
# Retrieve UMAP coordinates
umap_coords <- Embeddings(rl_cca, "umap")

# Calculate minimum and maximum values for each dimension
min_umap <- apply(umap_coords, 2, min) # Min (x,y)
max_umap <- apply(umap_coords, 2, max) # Max (x,y)

min_x <- ceiling(min_umap[1])
min_y <- ceiling(min_umap[2])
max_x <- ceiling(max_umap[1])
max_y <- ceiling(max_umap[2])

buffer_x <- (max_x - min_x)*0.05
buffer_y <- (max_y - min_y)*0.05

min_x <- min_x - buffer_x
min_y <- min_y - buffer_y
max_x <- max_x + buffer_x
max_y <- max_y + buffer_y

DefaultAssay(rl_cca) <- "RNA"
DefaultAssay(tumour_srat) <- "RNA"
tumour_srat <- NormalizeData(tumour_srat)

genes = c("CNTNAP2", "FOXP2", "EBF1", "EBF3", "DACH1", "RBFOX1", "TMTC2", "LRP1B", "SOX11", "SOX4", "MKI67", "NFIA", "SOX2", "RBFOX3", "PAX6", "NES", "HNRNPH1", "CTNN1", "BACH1", "HIVEP2")

#genes = c("USH2A") Xinghan was curious about this gene, has no relation to UBC_1

for(gene in genes){
  feature_plot <- FeaturePlot(object = rl_cca, features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                  ggtitle(paste("Integrated: ", gene)) +
                  xlim(min_x, max_x) +       # Set x-axis limits
                  ylim(min_y, max_y)         # Set y-axis limits

  feature_plot_human <- FeaturePlot(object = rl_cca, cells = rownames(rl_cca@meta.data[rl_cca$species == "human", ]),  
                        features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                        ggtitle(paste("Human: ", gene)) +
                        xlim(min_x, max_x) +       # Set x-axis limits
                        ylim(min_y, max_y)         # Set y-axis limits

  feature_plot_mouse <- FeaturePlot(object = rl_cca, cells = rownames(rl_cca@meta.data[rl_cca$species == "mouse", ]), 
                        features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) +
                        ggtitle(paste("Mouse: ", gene)) +
                        xlim(min_x, max_x) +       # Set x-axis limits
                        ylim(min_y, max_y)         # Set y-axis limits

  plot <- feature_plot | feature_plot_human | feature_plot_mouse
  feature_plot_name <- paste0(gene, ".png")
  ggsave(filename = feature_plot_name, plot = plot, path = out_directory, 
        width =30, height = 10, units = "in", dpi = 600)
  
  feature_plot <- FeaturePlot(object = tumour_srat, features = gene, order = TRUE, min.cutoff = "q10", max.cutoff = "q90", raster = FALSE) 

  feature_plot_name <- paste0(gene, "_tumour.png")
  ggsave(filename = feature_plot_name, plot = feature_plot, path = out_directory, 
        width =10, height = 10, units = "in", dpi = 600)

  }
```