---
title: Determining statistical significance of BRC1 kd and plotting results
date: 2024-08-07
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

# Goal of this analysis

1. Determining statistical significance of BRCA1 kd 
2. Plot results

# Files and directories

## Input

CSV is found at `/.mounts/labs/pailab/private/llau/results/brca1_experiment/20240807/siRNA_v2.csv`

## Output

The output can be found at `/.mounts/labs/pailab/private/llau/results/brca1_experiment/20240807`

# Process and Results

Using the `stats` conda environment

Lets load in the required libraries
```{r}
library(tidyverse)
library(ggpubr)
library(rstatix)
```

```{r}
set.seed(1)
out_directory <- "/.mounts/labs/pailab/private/llau/results/brca1_experiment/20240807"
if (!dir.exists(out_directory)) {
  dir.create(out_directory, recursive = TRUE)
}
```

Refactor the dataset and add mean and standard deviation of the cells
```{r}
si_rna <- read.csv("/.mounts/labs/pailab/private/llau/results/brca1_experiment/20240807/siRNA_v2.csv")
si_rna$time <- factor(si_rna$time, levels = c("48h", "72h", "96h", "120h", "144h"))
si_rna$sample <- factor(si_rna$sample, levels = c("WT", "Lipofectamine only control", "NT siRNA Knockdown", "BRCA1 siRNA Knockdown"))
#si_rna$id <- paste0(si_rna$time, si_rna$sample, si_rna$replicate)

summary_data <- si_rna %>%
  group_by(time, sample) %>%
  summarize(
    mean = mean(percent_cell_viability_normalized_WT_control),
    sd = sd(percent_cell_viability_normalized_WT_control),
    .groups = 'drop'
  )
```

Check assumptions for repeated measures ANOVA
```{r}
shapiro_results <- si_rna %>%
  group_by(time, sample) %>%
  shapiro_test(percent_cell_viability_normalized_WT_control)

shapiro_results

outlier_results <- si_rna %>%
  group_by(time, sample) %>%
  identify_outliers(percent_cell_viability_normalized_WT_control) %>%
  select(time, sample, replicate, percent_cell_viability_normalized_WT_control, is.outlier, is.extreme)

outlier_results
```

All samples have normal distribution except for 96h WT.

There is only one extreme outlier: 72h BRCA1 siRNA Knockdown Replicate 1.

Lets compute the repeated measures ANOVA
```{r}
anova_result <- si_rna %>%
  anova_test(
    formula = percent_cell_viability_normalized_WT_control ~ time * sample + Error(replicate/(time * sample))
  )
get_anova_table(anova_result)
```

There is a statistically significant two-way interactions between sample and time, F(12, 60) = 10.431, p = 1.03e-10.

Lets look at the effect of the sample at each time point
```{r}
# Effect of treatment at each time point
one.way <- si_rna %>%
  group_by(time) %>%
  anova_test(dv = percent_cell_viability_normalized_WT_control, wid = replicate, within = sample) %>%
  get_anova_table() %>%
  adjust_pvalue(method = "bonferroni")
one.way
```

Lets look at the pairwise comparisons between sample groups
```{r}
# Pairwise comparisions between sample groups for each time point
pwc <- si_rna %>%
  group_by(time) %>%
  pairwise_t_test(
    percent_cell_viability_normalized_WT_control ~ sample, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
write.csv(pwc, file = paste0(out_directory, "/pwc.csv"), row.names = FALSE)
```

```{r}
plt <- ggplot(summary_data, aes(x = time, y = mean, color = sample, group = sample)) +
    geom_line() + 
    geom_point() +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +
    labs(title = NULL,
        x = "Timepoint (hours)",
        y = "% cell viability normalized to its respective WT control",
        color = "Condition") +
    scale_x_discrete(
      breaks = c("48h", "72h", "96h", "120h", "144h"),  # Specify the categories to display
      labels = c("48", "72", "96", "120", "144")   # Specify the custom labels
    ) +
    theme(
      panel.background = element_rect(fill = "white"),
      plot.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black"),              # Axis lines in black
      axis.ticks = element_line(color = "black"),             # Axis ticks in black
      axis.text = element_text(color = "black"),              # Axis text in black
      axis.title = element_text(color = "black"),    
      #panel.grid.major = element_line(color = "grey90"),  # Light grey grid lines
      #panel.grid.minor = element_line(color = "grey95"),
      legend.text = element_text(size = 6),  # Adjust legend text size
      legend.title = element_text(size = 8),  # Adjust legend title size
      legend.key.size = unit(0.1, "in")  # Adjust legend key size
  )

ggsave(filename = "sirna_v2_plot.png", plot = plt, path = out_directory, 
            width = 7, height = 5, units = "in", dpi = 600)
```