---
title: "Different integration methods"
---

Code for figures showing the different integration methods. Note that the code chunks are run interactively using the `scrnaseq_env` conda environment. This file was **not** rendered with `quarto render`.

## Initialization

Packages:

```{r}
library(tidyverse)
library(patchwork)
library(cowplot)
library(Seurat)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))
source(file.path(thesis_dir, "../software/utilities/plotting.R"))

out_dir <- "./integ_methods"
my_pals <- get_custom_pals()
```

Function for making the plots for each integration method:

```{r}
make_plots <- function(
  srat,
  cluster_res
) {
  .plt <- list()
  base_size <- 6

  # UMAP coloured by species
  .plt$species <- DimPlot(
    srat,
    reduction = "umap",
    cells.highlight = WhichCells(srat, expression = species == "human"),
    sizes.highlight = 0.01,
    label = FALSE
  ) + 
    labs(title = "Species") + 
    scale_colour_manual(
      labels = c("mouse", "human"),
      values = my_pals$species
    ) + 
    guides(colour = guide_legend(override.aes = list(size = 1))) + 
    theme_classic2(base_size = base_size) + 
    theme(axis.text = element_blank(), axis.ticks = element_blank())

  # UMAP of clusters
  .plt$clusters <- DimPlot(
    srat,
    reduction = "umap",
    group.by = cluster_res,
    label = TRUE,
    label.size = 2,
    repel = TRUE
  ) + 
    labs(title = "Cluster") + 
    theme_classic2(base_size = base_size) + 
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    )

  # UMAP of RL-VZ/SVZ cells
  .plt$highlight_rl <- DimPlot(
    srat,
    reduction = "umap",
    cells.highlight = highlight_ids,
    sizes.highlight = 0.01,
    label = FALSE
  ) + 
    labs(title = "RL-VZ/SVZ cells") + 
    scale_colour_manual(
      labels = c("other", "VZ", "SVZ"),
      values = c("grey", highlight_cols)
    ) + 
    guides(colour = guide_legend(override.aes = list(size = 1))) + 
    theme_classic2(base_size = base_size) + 
    theme(axis.text = element_blank(), axis.ticks = element_blank())

  # bar plot showing cluster of RL-VZ/SVZ cells
  md <- srat[[]] %>%
    rownames_to_column(var = "cell_ids") %>%
    mutate(which_rl = case_when(
      cell_ids %in% highlight_ids$VZ ~ "VZ",
      cell_ids %in% highlight_ids$SVZ ~ "SVZ",
      .default = NA
    ))

  df <- table(md$which_rl, md[[cluster_res]]) %>%
    as.data.frame() %>%
    rename(which_rl = Var1, cluster = Var2, number = Freq) %>%
    mutate(
      which_rl = fct_rev(which_rl),
      cluster = fct_inseq(cluster)
    )

  .plt$bar_rl <- ggplot(df, aes(x = cluster, y = number, fill = which_rl)) + 
    geom_col(position = "stack", width = 0.75) + 
    labs(title = "Clustering of RL-VZ/SVZ", y = "number of cells", fill = "RL zone") + 
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) + 
    scale_fill_manual(values = highlight_cols) + 
    theme_classic2(base_size = base_size)

  return(.plt)
}
```

Load Seurat objects:

```{r}
srat_qs <- get_srat_paths()
all_srat <- load_srat(srat_qs[c("rl", "rl_rpca", "rl_harmony")]) %>%
  setNames(nm = c("cca", "rpca", "harm"))
```

Load cell IDs for the RL-VZ/SVZ cells:

```{r}
vz <- read.csv("/.mounts/labs/pailab/private/llau/results/integrated/20240509/cca/vz_cells.csv")
svz <- read.csv("/.mounts/labs/pailab/private/llau/results/integrated/20240509/cca/svz_cells.csv")

highlight_ids <- list(VZ = pull(vz, 2), SVZ = pull(svz, 2))
highlight_cols <- scales::pal_hue()(3)[2:3]
```


## CCA plots

```{r}
cca_plots <- make_plots(srat = all_srat$cca, cluster_res = "snn_res.0.4")

ggsave(
  filename = "cca_plots.png",
  plot = wrap_plots(cca_plots, nrow = 2),
  path = out_dir,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300
)
```


## RPCA plots

```{r}
rpca_plots <- make_plots(srat = all_srat$rpca, cluster_res = "snn_res.0.2")

ggsave(
  filename = "rpca_plots.png",
  plot = wrap_plots(rpca_plots, nrow = 2),
  path = out_dir,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300
)
```


## Harmony plots

```{r}
harm_plots <- make_plots(srat = all_srat$harm, cluster_res = "snn_res.0.3")

ggsave(
  filename = "harmony_plots.png",
  plot = wrap_plots(harm_plots, nrow = 2),
  path = out_dir,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300
)
```


## Final figure

```{r}
# extract legends
rl_lgd <- cowplot::get_legend(
  cca_plots$bar_rl + 
    theme(
      legend.title = element_text(size = rel(1.2)),
      legend.text = element_text(size = rel(1))
    )
)
species_lgd <- cowplot::get_plot_component(
  cca_plots$species + 
    labs(colour = "Species") + 
    theme(
      legend.title = element_text(size = rel(1.2)),
      legend.text = element_text(size = rel(1)),
      legend.position = "bottom",
      legend.background = element_rect(fill = NA, colour = NA)
    ),
  "guide-box",
  return_all = TRUE
) %>% 
  pluck(3)

# line under label for integration method
annot_line <- ggdraw() + 
  draw_line(x = c(0.025,0.99), y = c(0.5,0.5), colour = "black", size = 0.5)

# add tag to each plot
rpca_plots <- add_plot_tags(rpca_plots, tags = c("D", "C", "A", "B"))
harm_plots <- add_plot_tags(harm_plots, tags = c("H", "G", "E", "F"))
cca_plots <- add_plot_tags(cca_plots, tags = c("L", "K", "I", "J"))
```

```{r}
.plt <- plot_grid(
  # RPCA
  ggdraw() + draw_text(text = "RPCA", x = 0.5, y = 0.5, size = 12),
  annot_line,
  plot_grid(
    plotlist = map(rpca_plots[c(3,4,2,1)], \(x) {x <- x + theme(legend.position = "none")}),
    align = "h",
    axis = "t",
    ncol = 4
  ),
  # Harmony
  ggdraw() + draw_text(text = "Harmony", x = 0.5, y = 0.5, size = 12),
  annot_line,
  plot_grid(
    plotlist = map(harm_plots[c(3,4,2,1)], \(x) {x <- x + theme(legend.position = "none")}),
    align = "h",
    axis = "t",
    ncol = 4
  ),
  # CCA
  ggdraw() + draw_text(text = "CCA", x = 0.25, y = 0.5, size = 12),
  ggdraw() + draw_line(x = c(0.025,0.49), y = c(0.5,0.5), colour = "black", size = 0.5),
  plot_grid(
    plotlist = map(cca_plots[c(3,4)], \(x) {x <- x + theme(legend.position = "none")}),
    align = "h",
    axis = "t",
    ncol = 4
  ),
  # overall layout arguments
  nrow = 9,
  rel_heights = rep(c(0.5, 0.3, 9.2), 3)
)

.plt <- cowplot::ggdraw(.plt) + 
  cowplot::draw_plot(rl_lgd, x = 0.51, y = 1/6, width = 0, height = 0) + 
  cowplot::draw_plot(species_lgd, x = 7/8, y = 0.33, width = 0, height = 0)

walk(
  .x = c("png", "pdf"),
  .f = \(device) {
    ggsave(
      filename = paste("final", device, sep = "."),
      plot = .plt,
      path = out_dir,
      width = 9,
      height = 8,
      units = "in",
      dpi = 600
    )
  }
)
```
