---
title: "Pseudotime of the RL lineage"
---

Code for figures of the RL pseudotime results. Note that the code chunks are run interactively using the `monocle3` conda environment. This file was **not** rendered with `quarto render`.

## Initialization

Packages:

```{r}
library(tidyverse)
library(patchwork)
library(monocle3)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))

out_dir <- "./pseudotime"
my_pals <- get_custom_pals()
```


## UMAPs of pseudotime

```{r}
# load cell_data_set object
cds <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240611/human_pseudotime/post_pseudotime_cds.qs")

# set default `cell_size` and `cell_stroke` when plotting
cell_size <- 1
cell_stroke <- 0
```

```{r}
# UMAP showing trajectory with cells coloured by cluster
traj_umap <- plot_cells(
  cds,
  color_cells_by = "cell_type_annot",
  label_cell_groups = FALSE,
  label_branch_points = FALSE,
  label_roots = FALSE,
  label_leaves = FALSE,
  cell_size = cell_size,
  cell_stroke = cell_stroke
) + 
  scale_colour_manual(
    values = c(my_pals$rl_integ_annot, my_pals$ubc_integ_clust)
  ) + 
  guides(
    colour = guide_legend(title = "cell type", override.aes = list(size = 3))
  ) + 
  theme_classic2()

# remove ugly black border around the points
traj_umap$layers[[1]] <- NULL

ggsave(
  filename = "trajectory_umap.png",
  plot = traj_umap,
  path = out_dir,
  width = 6,
  height = 5,
  units = "in",
  dpi = 600
)
```

```{r}
# UMAP with cells coloured by pseudotime
ptime_umap <- plot_cells(
  cds,
  color_cells_by = "pseudotime",
  show_trajectory_graph = FALSE,
  label_cell_groups = FALSE,
  cell_size = cell_size,
  cell_stroke = cell_stroke
) + 
  theme_classic2()

# remove ugly black border around the points
ptime_umap$layers[[1]] <- NULL

ggsave(
  filename = "pseudotime_umap.png",
  plot = ptime_umap,
  path = out_dir,
  width = 6,
  height = 5,
  units = "in",
  dpi = 600
)
```


## Distribution of pseudotime

Get distribution for each age:

```{r}
# read in pseudotime values for each cell
pseudotime_df <- read_csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240611/human_pseudotime/pseudotime_values.csv")

# extract metadata from cell_data_set object
md <- colData(cds) %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell_id") %>%
  mutate(human_age = fct_drop(human_age)) %>%
  # add pseudotime values to metadata
  full_join(
    y = pseudotime_df,
    by = "cell_id"
  )
```

```{r}
# ridgeline plot of pseudotime distribution by age
age_ridge <- ggplot(
  md,
  aes(x = pseudotime, y = fct_rev(human_age))
) + 
  ggridges::geom_density_ridges() + 
  labs(y = "age") + 
  # scale_y_discrete(expand = expansion(mult = c(0, 0.05))) + 
  ggridges::theme_ridges(font_size = 11, center_axis_labels = TRUE) + 
  theme(plot.background = element_rect(fill = "white"))
ggsave(
  filename = "age_ridge.png",
  plot = age_ridge,
  path = out_dir,
  width = 5,
  height = 5,
  units = "in",
  dpi = 600
)
```

Plot distribution for each cell type (RL/UBC only):

```{r}
# filter for RL/UBC
ubc_lineage <- filter(
  md,
  str_detect(string = cell_type_annot, pattern = "(^RL)|(^UBC)")
)

# set factor levels for the plot
ubc_lineage$cell_type_annot <- factor(
  ubc_lineage$cell_type_annot,
  levels = c("RL", paste0("UBC_", c(1, 5, 0, 4, 3, 2)))
)
```

```{r}
# ridgeline plot of pseudotime distribution for RL/UBC clusters
cell_type_ridge <- ggplot(
  ubc_lineage,
  aes(x = pseudotime, y = fct_rev(cell_type_annot))
) + 
  ggridges::geom_density_ridges(
    aes(fill = cell_type_annot),
    scale = 2 # higher means ridges are closer together
  ) + 
  labs(y = "cell type") + 
  scale_y_discrete(expand = expansion(mult = c(0.02, 0.35))) + 
  scale_fill_manual(values = c(
    my_pals$rl_integ_annot[c("RL", "UBC")],
    my_pals$ubc_integ_clust
  )) + 
  guides(fill = guide_legend(title = "cell type")) + 
  ggridges::theme_ridges(font_size = 11, center_axis_label = TRUE) + 
  theme(plot.background = element_rect(fill = "white"))
ggsave(
  filename = "cell_type_ridge.png",
  plot = cell_type_ridge,
  path = out_dir,
  width = 6,
  height = 5,
  units = "in",
  dpi = 600
)
```

## Final figure

```{r}
layout <- c(
  # UMAPs
  area(t = 1, l = 1, r = 8),
  # ridgeline plot of cell types
  area(t = 2, l = 2, r = 7)
)

# combine UMAPs
.plt <- traj_umap + ptime_umap + guide_area() + 
  plot_layout(
    design = c(
      area(t = 1, l = 1, r = 2),
      area(t = 1, l = 3, r = 4),
      area(t = 1, l = 5)
    ),
    guides = "collect"
  )

# add ridgeline plot
.plt <- wrap_plots(.plt, free(cell_type_ridge)) + 
  plot_layout(design = layout, guides = "keep") + 
  plot_annotation(
    title = "Human rhombic lip lineage cells",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(face = "plain", hjust = 0.5, size = rel(1.5))
    )
  ) & 
  theme(
    plot.tag = element_text(face = "bold", hjust = 0.5, vjust = 0.5),
    plot.tag.location = "plot",
  )

walk(
  .x = c("png", "pdf"),
  .f = \(device) {
    ggsave(
      filename = paste0("final.", device),
      plot = .plt,
      path = out_dir,
      width = 9,
      height = 7.5,
      units = "in",
      dpi = 600
    )
  }
)
```
