---
title: "UBC subclusters"
---

Code for figures showing the proportion tests in the UBC subclusters. Note that the code chunks are run interactively using the `scrnaseq_env` conda environment. This file was **not** rendered with `quarto render`.

## Initialization

Packages:

```{r}
library(tidyverse)
library(patchwork)
library(Seurat)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))
source(file.path(thesis_dir, "../software/utilities/plotting.R"))
source(file.path(thesis_dir, "../software/utilities/propeller_helpers.R"))

out_dir <- "./ubc_subclusters"
my_pals <- get_custom_pals()
```


## UMAP of UBC subclusters

```{r}
# load Seurat object for integrated dataset
srat_qs <- get_srat_paths()
ubc_srat <- load_srat(srat_qs["ubc"]) %>%
  pluck("ubc")
```

```{r}
# UMAP coloured by subclusters
subcluster_umap <- DimPlot(
  ubc_srat,
  reduction = "umap",
  group.by = "subclusters",
  cols = my_pals$ubc_integ_clust,
  label = TRUE,
  label.size = 3,
  repel = TRUE
) + 
  labs(title = NULL) + 
  theme_classic2() + 
  theme(legend.position = "none")

# hacky way of setting the seed for the UMAP labels
subcluster_umap[[1]]$layers[[2]]$geom_params$seed <- 42

ggsave(
  filename = "subcluster_umap.png",
  plot = subcluster_umap,
  path = out_dir,
  width = 5,
  height = 5,
  units = "in",
  dpi = 600
)
```


## Proportion of cells in each UBC subcluster

```{r}
# set species factor levels
ubc_srat$species <- factor(ubc_srat$species, levels = c("mouse", "human"))
```

```{r}
# proportion of UBC subclusters in human vs mouse
prop_bar <- cluster_barplot(
  object = ubc_srat,
  split.by = "subclusters",
  group.by = "species",
  position = "fill"
) + 
  labs(fill = "subcluster") + 
  scale_fill_manual(values = my_pals$ubc_integ_clust) + 
  theme_classic2()
ggsave(
  filename = "subcluster_prop_bar.png",
  plot = prop_bar,
  path = out_dir,
  width = 4,
  height = 4,
  units = "in",
  dpi = 600
)
```


## Proportion tests

### `scProportionTest`

```{r}
# load scProportionTest results
scpt_res <- read_csv("/.mounts/labs/pailab/private/llau/results/integrated/20240718/proportion_test/cca/cca_ubc_clusters_proportion_test.csv")

# add significance column and set plot order
log2FD_threshold <- log2(2)
FDR_threshold <- 0.05
scpt_res <- mutate(
  scpt_res,
  significance = case_when(
    FDR < FDR_threshold & obs_log2FD > log2FD_threshold ~ "human",
    FDR < FDR_threshold & obs_log2FD < -log2FD_threshold ~ "mouse",
    .default = "n.s."
  ),
  significance = factor(significance, levels = c("mouse", "n.s.", "human")),
  clusters = fct_reorder(clusters, obs_log2FD, .desc = FALSE)
)

# plot scProportionTest results
scpt_plt <- ggplot(
  data = scpt_res,
  mapping = aes(
    x = obs_log2FD,
    y = clusters,
    xmin = boot_CI_2.5,
    xmax = boot_CI_97.5,
    colour = significance
  )
) + 
  geom_pointrange(size = 0.25) + 
  geom_vline(
    xintercept = c(log2FD_threshold, -log2FD_threshold),
    lty = "dashed"
  ) + 
  geom_vline(xintercept = 0, lty = "solid") + 
  labs(x = "observed log2(FD)", y = "UBC cluster", colour = "enriched") + 
  scale_colour_manual(values = c(my_pals$species[1], "black", my_pals$species[2])) + 
  theme_classic2() + 
  theme(legend.position = "bottom")
ggsave(
  filename = "permutation_plot.png",
  plot = scpt_plt,
  path = out_dir,
  width = 5,
  height = 3,
  units = "in",
  dpi = 600
)
```

### `propeller`

```{r}
# set same cluster plotting order as scProportionTest
cluster_levels <- sort(levels(scpt_res$clusters))

# load `propeller` sample proportions
propeller_props <- read_csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240606/four_dataset_cca_ubc_proportions.csv") %>%
  mutate(
    dataset_name = str_replace(
      string = dataset_name,
      pattern = "_([:alpha:]+)",
      replacement = " (\\1)"
    ),
    species = factor(species, levels = c("mouse", "human")),
    clusters = paste0("UBC_", clusters) %>% factor(levels = cluster_levels)
  )

# load propeller
propeller_res <- read_csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240606/four_dataset_cca_ubc_results.csv")
fdrs <- get_fdrs_for_plotting(propeller_res) %>%
  mutate(
    clusters = paste0("UBC_", clusters) %>% factor(levels = cluster_levels)
  )

# plot `propeller` results for UBCs
propeller_plt <- boxplot_props_per_cluster(
  props = propeller_props,
  facet_var = clusters,
  facet_ncol = 6,
  fdrs = fdrs,
  fdr_label_size = 8.8
) + 
  labs(y = "proportion of all\nRL lineage cells", shape = "dataset") + 
  theme_classic2()
ggsave(
  filename = "propeller_plot.png",
  plot = propeller_plt,
  path = out_dir,
  width = 14,
  height = 4,
  units = "in",
  dpi = 600
)
```


## Final figure

```{r}
layout <- c(
  # UMAP
  area(t = 1, l = 1, r = 3),
  # bar plot
  area(t = 1, l = 4, r = 6),
  # scProportionTest (permutation plot)
  area(t = 1, l = 7, r = 10),
  # propeller plot
  area(t = 2, l = 1, r = 10)
)

.plt <- wrap_plots(
  subcluster_umap,
  free(prop_bar),
  free(scpt_plt),
  free(propeller_plt)
) + 
  plot_layout(design = layout) + 
  plot_annotation(tag_levels = "A")

walk(
  .x = c("png", "pdf"),
  .f = \(device) {
    ggsave(
      filename = paste0("final.", device),
      plot = .plt,
      path = out_dir,
      width = 10,
      height = 6,
      units = "in",
      dpi = 600
    )
  }
)
```
