---
title: "UBC clustering resolutions"
---

Code for figure showing the `clustree` visualization from clustering the UBCs at different resolutions. Note that the code chunks are run interactively using the `clustree` conda environment. This file was **not** rendered with `quarto render`.

The code was adapted from Leo's 2024-05-27 notes (same plot, just need a different size/dpi for thesis).

## Initialization

Packages:

```{r}
library(tidyverse)
library(Seurat)
library(clustree)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))

out_dir <- "./ubc_cluster_res"
# my_pals <- get_custom_pals()
```


## Clustering tree plot

```{r}
# load Seurat object for integrated dataset
srat_qs <- get_srat_paths()
ubc_srat <- load_srat(srat_qs["ubc"]) %>%
  pluck("ubc")
```

```{r}
.plt <- clustree(ubc_srat, prefix = "snn_res.") + 
  labs(title = "Clustering tree of UBCs") + 
  guides(
    colour = guide_legend(title = "cluster\nresolution", order = 1),
    edge_alpha = guide_legend(title = "in-proportion", order = 2),
    edge_colour = guide_legend(title = "number\nof cells", order = 3),
    size = guide_legend(title = "number\nof cells", order = 4)
  ) + 
  theme(
    plot.title = element_text(face = "plain", hjust = 0.5),
    plot.title.position = "plot",
    plot.margin = margin(5.5, 5.5, 5.5, 5.5, "points")
  )

walk(
  .x = c("png", "pdf"),
  .f = \(device) {
    ggsave(
      filename = paste0("final.", device),
      plot = .plt,
      path = out_dir,
      width = 9,
      height = 9,
      units = "in",
      dpi = 600
    )
  }
)
```
