---
title: "UBC signatures"
---

Code for heatmap showing the signatures of the UBC subclusters. Note that the code chunks are run interactively using the `scrnaseq_env` conda environment. This file was **not** rendered with `quarto render`.

See notes from 2024-11-03; however, note that these results use Leo's differential gene expression results from 2024-07-15 rather than Shraddha's results from 2024-10-29.

## Initialization

Packages:

```{r}
library(tidyverse)
library(ComplexHeatmap)
library(Seurat)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))
source(file.path(thesis_dir, "../software/utilities/plotting.R"))

out_dir <- "./ubc_signatures"
my_pals <- get_custom_pals()
```


## UBC signatures heatmap

```{r}
# load Seurat object for human UBCs
human_ubcs <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241103/human_ubc_srat_merged.qs")
```

```{r}
ubc_subclusters <- paste0("UBC_", c(1,5,0,4,3,2))

# load top markers in each cluster across both datasets (Aldinger/Sepp)
# based on Leo's differential expression results from 2024-07-15
ranked_markers <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241029/cluster_marker_ranking.rds")
unique_markers <- map(
  .x = ubc_subclusters,
  .f = \(clust) {
    x <- ranked_markers[[clust]]
    nr <- min(nrow(x), 100)
    x <- pull(x, feature) %>%
      head(nr)
    return(x)
  }
) %>%
  unlist() %>%
  unique()
```

```{r}
# heatmap annotations

# create top annotation rows
ages <- human_ubcs[[]] %>%
  select(subclusters, human_age) %>%
  table() %>%
  prop.table(margin = 1) %>%
  .[ubc_subclusters, ]
sex <- human_ubcs[[]] %>%
  select(subclusters, sex) %>%
  table() %>%
  prop.table(margin = 1) %>%
  .[ubc_subclusters, ]
dset <- human_ubcs[[]] %>%
  select(subclusters, dataset_name) %>%
  mutate(dataset_name = str_remove(dataset_name, "_full_cerebellum_human")) %>%
  table() %>%
  prop.table(margin = 1) %>%
  .[ubc_subclusters, ]
pal1 <- RColorBrewer::brewer.pal(ncol(ages), "YlOrRd") # age
pal2 <- c("pink","blue") # sex
pal3 <- RColorBrewer::brewer.pal(3, "Dark2") # dataset

topannot <- HeatmapAnnotation(
  age = anno_barplot(
    ages, 
    gp = gpar(fill = pal1, border="white", lty="solid"),
    bar_width = 1, 
    height = unit(2, "in"),
    axis=FALSE
  ),
  sex = anno_barplot(
    sex,
    gp = gpar(fill = pal2, border="white"),
    bar_width = 1,
    height = unit(1,"in"),
    axis=FALSE
  ),
  dataset = anno_barplot(
    dset,
    gp = gpar(fill = pal3, border="white"),
    bar_width = 1,
    height = unit(1,"in"),
    axis=FALSE
  )
)
lgd_list <- list(
  Legend(labels=colnames(ages),title="age",legend_gp=gpar(fill=pal1)),
  Legend(labels=colnames(sex),title="sex",legend_gp=gpar(fill=pal2)),
  Legend(labels=colnames(dset),title="dataset",legend_gp=gpar(fill=pal3))
)

# genes to annotate
annoGenes <- map(
  .x = ranked_markers,
  .f = \(df) {
    head(df$feature, 5)
  }
) %>%
  unlist() %>%
  unname() %>%
  c("OTX2", "CBFA2T2", "SOX4", "SOX11", "LMX1A", "EOMES")
```

```{r}
# get data slot, then manually scale the results
xpr <- AverageExpression(
  human_ubcs,
  assays = "SCT",
  features = unique_markers,
  group.by = "subclusters",
  slot = "data"
)[[1]]
xpr <- t(scale(t(xpr)))
# change column order
xpr <- xpr[, ubc_subclusters]

# heatmap colours
col_fun <- circlize::colorRamp2(
  breaks = c(-2,0,2),
  colors = c("#FF00FF","#000000", "#FFFF00")
)

ha <- rowAnnotation(
  foo = anno_mark(at = match(annoGenes, rownames(xpr)), labels = annoGenes)
)

hm <- Heatmap(
  xpr,
  name="scaled\nexpression",
  cluster_rows=FALSE,
  cluster_columns=FALSE,
  col = col_fun,
  show_row_names = FALSE,
  column_title="SCT data slot then manual scale()",
  top_annotation = topannot,
  right_annotation = ha
)

png(
  filename = file.path(out_dir, "final.png"),
  width = 8,
  height = 12,
  units = "in",
  res = 300
)
draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()
```
