---
title: "Upregulated genes and active regulons in human-enriched UBCs"
---

Code for figures showing upregulated genes, pathways, and regulons in the UBC subclusters. Note that the code chunks are run interactively using the `scrnaseq_env` conda environment. This file was **not** rendered with `quarto render`.

## Initialization

Packages:

```{r}
library(tidyverse)
library(patchwork)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))
source(file.path(thesis_dir, "../software/utilities/pyscenic_regulons.R"))

out_dir <- "./ubc_genes_regulons"
my_pals <- get_custom_pals()
```


## Volcano plot from differential gene expression

```{r}
# custom volcano plot function (slightly different from the one in `plotting.R`)
make_volcano_plot <- function(
  data,
  log_fc,
  log_pval,
  direction,
  gene_labels
) {
  .plt <- ggplot(
    data = data,
    mapping = aes(
      x = {{ log_fc }},
      y = {{ log_pval }}
    )
  ) + 
    geom_point(
      aes(colour = {{ direction }}),
      alpha = 0.5,
      shape = "circle",
      size = 2,
      stroke = 0
    ) + 
    ggrepel::geom_text_repel(
      aes(label = {{ gene_labels }}, colour = {{ direction }}),
      size = 2.5,
      min.segment.length = 0.1,
      max.overlaps = Inf,
      show.legend = FALSE,
      seed = 42,
    ) + 
    geom_vline(
      xintercept = 0,
      colour = "grey",
      linewidth = 0.25,
      linetype = "dashed"
    ) + 
    geom_hline(
      yintercept = -log10(0.05),
      colour = "grey",
      linewidth = 0.25,
      linetype = "dashed"
    )

  return(.plt)
}
```

```{r}
# load differential gene expression results (all UBC clusters, both datasets)
ubc_diff_expr <- read_csv("/.mounts/labs/pailab/private/llau/results/integrated/20240715/all_tested_genes.csv") %>%
  rename_with(.fn = str_remove, pattern = "full_cerebellum_")
```

```{r}
# load top markers in each cluster across both datasets (Aldinger/Sepp)
ranked_markers <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241029/cluster_marker_ranking.rds")

ubc_subclusters <- names(ranked_markers)
```

```{r}
# volcano plots
volcano_plots <- map(
  .x = c("aldinger", "sepp"),
  .f = \(dataset) {
    # which column to use (based on dataset)
    logfc_column <- paste0(str_to_title(dataset), "_human_avg_log2FC")
    pval_column <- paste0(str_to_title(dataset), "_human_p_val")

    # set up empty list for plots to return
    .plts <- vector(mode = "list", length = length(ubc_subclusters))
    names(.plts) <- ubc_subclusters

    for (clust in ubc_subclusters) {
      # volcano plot colours
      volcano_cols <- c(
        "down" = my_pals$rl_integ_annot[["UBC"]],
        "n.s." = "black",
        "up" = my_pals$ubc_integ_clust[[clust]]
      )

      # additional columns for the volcano plot
      diff_expr <- ubc_diff_expr %>%
        filter(cluster == clust) %>%
        mutate(direction = case_when(
          rlang::parse_expr(
            paste0(logfc_column, " < 0 & ", pval_column, " < 0.05 ~ 'down'")
          ),
          rlang::parse_expr(
            paste0(logfc_column, " > 0 & ", pval_column, " < 0.05 ~ 'up'")
          ),
          .default = "n.s."
        ))

      # top 10 features for labelling
      features <- ranked_markers %>%
        pluck(clust) %>%
        slice_min(combined_rank, n = 10, with_ties = TRUE) %>%
        pull(feature)
      diff_expr <- diff_expr %>%
        mutate(gene_label = case_when(
          feature %in% features ~ feature,
          .default = ""
        ))

      # label number of genes
      num_up <- nrow(filter(diff_expr, direction == "up"))
      num_down <- nrow(filter(diff_expr, direction == "down"))
      caption <- paste0(
        "total = ", nrow(diff_expr),
        "; down = ", num_down,
        "; up = ", num_up
      )

      # volcano plot
      .plt <- make_volcano_plot(
        data = diff_expr,
        log_fc = .data[[logfc_column]],
        log_pval = -log10(.data[[pval_column]]),
        direction = direction,
        gene_labels = gene_label
      ) + 
        labs(
          title = str_to_title(dataset),
          caption = caption,
          x = "log2(FC)",
          y = "-log10(p-value)"
        ) + 
        scale_colour_manual(values = volcano_cols) +  
        guides(colour = guide_legend(
          title = clust,
          override.aes = list(alpha = 1)
        )) + 
        theme_classic2()
      ggsave(
        filename = paste0(clust, "_", dataset, "_volcano.png"),
        plot = .plt,
        path = out_dir,
        width = 4,
        height = 4,
        units = "in",
        dpi = 600
      )

      .plts[[clust]] <- .plt
    }

    return(.plts)
  }
) %>%
  setNames(nm = c("aldinger", "sepp"))
```


## Pathway enrichment analysis of upregulated genes

```{r}
# load pathway enrichment results for all UBC subclusters
ubc_pathways <- map(
  .x = ubc_subclusters,
  .f = \(clust) {
    ubc_pathways <- read_csv(file.path(
      "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241029",
      clust,
      "enr_pathways.csv"
    ))

    return(ubc_pathways)
  }
) %>%
  setNames(nm = ubc_subclusters)
```

```{r}
# enriched pathways for all UBC subclusters
pathway_bar <- map(
  .x = ubc_subclusters,
  .f = \(clust) {
    # subset pathways for the subcluster
    pway <- ubc_pathways[[clust]]

    # number of significant pathways
    n_sig <- nrow(filter(pway, p_value < 0.05))

    # total number of enriched pathways
    caption <- paste0("# enriched pathways/total = ", n_sig, "/", nrow(pway))

    # how many pathways to show?
    if (n_sig == 0) {
      # if no significant pathways, just show top 10
      n_plot <- 10
    } else {
      # show up to 15 pathways
      n_plot <- min(n_sig, 15)
    }

    # get pathways to plot
    pway <- slice_min(
      pway,
      order_by = p_value,
      n = n_plot
    )

    # set font size and width of terms
    term_font_size <- rel(0.8)
    lab_width <- 30

    # plot enriched pathways
    .plt <- ggplot(
      pway,
      aes(
        x = -log10(p_value),
        y = fct_reorder(.f = term_name, .x = p_value, .desc = TRUE)
      )
    ) + 
      geom_col(fill = my_pals$ubc_integ_clust[[clust]]) + 
      labs(
        title = sprintf("Enriched pathways in %s", clust),
        caption = caption,
        x = "-log10(FDR)",
        y = "pathway term"
      ) + 
      scale_x_continuous(expand = expansion(mult = c(0, 0.05))) + 
      scale_y_discrete(label = scales::label_wrap(lab_width)) + 
      theme_classic2() + 
      theme(
        axis.text.y = element_text(size = term_font_size),
        plot.title = element_text(vjust = 1)
      )
    ggsave(
      filename = paste0(clust, "_pathways.png"),
      plot = .plt,
      path = out_dir,
      width = 5,
      height = 6,
      units = "in",
      dpi = 600
    )

    return(.plt)
  }
) %>%
  setNames(nm = ubc_subclusters)
```


## Regulon specificity score plot

```{r}
# load regulon specificity scores
rss <- read.csv(
  file = "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240704/cell_type_annot/aldinger_sepp_RL.rss.csv",
  row.names = 1
)
```

```{r}
# rank top regulons for all UBCs
ubc_top_rss <- map(
  .x = ubc_subclusters,
  .f = \(cluster) {
    df <- data.frame(
      gene = colnames(rss),
      score = rss[cluster, ] %>% as.numeric(),
      rank = rank(-rss[cluster, ], ties.method = "random") %>% as.integer()
    ) %>%
      mutate(gene_label = case_when(
        rank <= 8 ~ gene,
        .default = ""
      )) %>%
      # change order for plotting
      arrange(desc(rank))

    return(df)
  }
) %>%
  setNames(nm = ubc_subclusters)
```

```{r}
# RSS plots
rss_plots <- map2(
  .x = ubc_top_rss,
  .y = names(ubc_top_rss),
  .f = \(top_rss, clust) {
    .plt <- ggplot(
      top_rss,
      aes(x = rank, y = score)
    ) + 
      geom_point(colour = ifelse(
        top_rss$gene_label == "",
        "black",
        my_pals$ubc_integ_clust[[clust]]
      )) + 
      ggrepel::geom_text_repel(
        aes(label = gene_label),
        colour = my_pals$ubc_integ_clust[[clust]],
        size = 2.5,
        segment.colour = alpha(my_pals$ubc_integ_clust[[clust]], 0.5),
        min.segment.length = 0.1,
        max.overlaps = Inf,
        nudge_y = -0.005,
        show.legend = FALSE,
        seed = 100
      ) + 
      labs(title = "pySCENIC", x = "rank", y = "regulon specificity score") + 
      theme_classic2()
    ggsave(
      filename = paste0(clust, "_rss_plot.png"),
      plot = .plt,
      path = out_dir,
      width = 3,
      height = 4,
      units = "in",
      dpi = 600
    )

    return(.plt)
  }
)
```


## Final figure

```{r}
layout <- c(
  # volcano plots
  area(t = 1, b = 3, l = 1, r = 11),
  # volcano plot legend
  area(t = 2, l = 12),
  # pathway enrichment
  area(t = 4, b = 9, l = 1, r = 6),
  # rss plot
  area(t = 4, b = 6, l = 8, r = 12)
)

walk(
  .x = ubc_subclusters,
  .f = \(clust) {
    # get volcano plots first
    .plt <- volcano_plots$aldinger[[clust]] + volcano_plots$sepp[[clust]] & 
      labs(caption = NULL)

    # add remaining plots
    .plt <- wrap_plots(
      .plt,
      guide_area(),
      free(pathway_bar[[clust]]) + labs(caption = NULL),
      free(rss_plots[[clust]])
    ) + 
      plot_layout(design = layout, guides = "collect") + 
      plot_annotation(tag_levels = list(c("A", "", "B", "C")))

    walk(
      .x = c("png", "pdf"),
      .f = \(device) {
        ggsave(
          filename = paste0(clust, "_final.", device),
          plot = .plt,
          path = out_dir,
          width = 7.5,
          height = 9,
          units = "in",
          dpi = 600
        )
      }
    )
  }
)
```
