---
title: "Medulloblastoma similarity to human-enriched UBCs"
---

Code for UMAP and gene expression of medulloblastoma datasets. Note that the code chunks are run interactively using the `scrnaseq_env` conda environment. This file was **not** rendered with `quarto render`.

## Initialization

Packages:

```{r}
library(tidyverse)
library(patchwork)
library(Seurat)
library(ggVennDiagram)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))

out_dir <- "./mb_similarity"
my_pals <- get_custom_pals()
```

```{r}
mb_subtypes <- c("SHH", "G3", "G4")
```


## UMAP of MB tumour cells

```{r}
# load medulloblastoma Seurat object
srat_qs <- get_srat_paths()
mb_srat <- load_srat(srat_qs["mb"]) %>%
  pluck("mb")

# set levels for MB subtypes
mb_srat$subtype <- factor(
  mb_srat$subtype,
  levels = mb_subtypes
)
```

```{r}
# UMAP coloured by subtype
mb_subtype_umap <- DimPlot(
  mb_srat,
  reduction = "umap",
  group.by = "subtype",
  cols = my_pals$mb_subtype,
  label = TRUE,
  label.size = 3,
  repel = TRUE
) + 
  labs(title = NULL) + 
  theme_classic2() + 
  theme(legend.position = "bottom")

# hacky way of setting the seed for the UMAP labels
mb_subtype_umap[[1]]$layers[[2]]$geom_params$seed <- 100

ggsave(
  filename = "mb_subtype_umap.png",
  plot = mb_subtype_umap,
  path = out_dir,
  width = 5,
  height = 5,
  units = "in",
  dpi = 600
)
```


## Regulon specificity score plot for MB subtypes

```{r}
# load regulon specificity scores
rss <- read.csv(
  file = "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240928/vladoiu_mb_subtype_rss.csv",
  row.names = 1
)
```

```{r}
# rank top regulons for all MB clusters
mb_top_rss <- map(
  .x = mb_subtypes,
  .f = \(cluster) {
    df <- data.frame(
      gene = colnames(rss),
      score = rss[cluster, ] %>% as.numeric(),
      rank = rank(-rss[cluster, ], ties.method = "random") %>% as.integer()
    ) %>%
      mutate(gene_label = case_when(
        rank <= 18 ~ gene,
        .default = ""
      )) %>%
      # change order for plotting
      arrange(desc(rank))

    return(df)
  }
) %>%
  setNames(nm = mb_subtypes)
```

```{r}
# RSS plots
rss_plots <- map2(
  .x = mb_top_rss,
  .y = names(mb_top_rss),
  .f = \(top_rss, subtype) {
    .plt <- ggplot(
      top_rss,
      aes(x = rank, y = score)
    ) + 
      geom_point(colour = ifelse(
        top_rss$gene_label == "",
        "black",
        my_pals$mb_subtype[[subtype]]
      )) + 
      ggrepel::geom_text_repel(
        aes(label = gene_label),
        colour = my_pals$mb_subtype[[subtype]],
        size = 2.5,
        segment.colour = alpha(my_pals$mb_subtype[[subtype]], 0.5),
        min.segment.length = 0.1,
        max.overlaps = Inf,
        nudge_x = 30,
        nudge_y = -0.02,
        show.legend = FALSE,
        seed = 100
      ) + 
      labs(x = "rank", y = "regulon specificity score") + 
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) + 
      theme_classic2()
    ggsave(
      filename = paste0(subtype, "_rss_plot.png"),
      plot = .plt,
      path = out_dir,
      width = 3,
      height = 4,
      units = "in",
      dpi = 600
    )

    return(.plt)
  }
)
```


## Common regulons/TFs in medulloblastoma and UBCs

between the UBC subclusters and MB tumours
- which TFs are upregulated in both?
- what are the common top regulons?

### Common TFs

See full analysis from 2024-09-17.

First, load the list of TFs from Lambert et al.:

```{r}
# load list of TFs
tfs <- read.csv(
  "/.mounts/labs/pailab/src/transcription-factors/human-tfs-lambert/full_database_v1.01.csv",
  row.names = 1,
  check.names = FALSE
) %>%
  # get only confirmed TFs
  dplyr::filter(`Is TF?` == "Yes") %>%
  pull(`HGNC symbol`)
```

```
> # total number of transcription factors
> length(tfs)
[1] 1639
```

Upregulated TFs in UBCs:

```{r}
# load genes that are significantly upregulated in both human datasets
upreg_ubc_genes <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240827/diff_expr_markers/cluster_marker_ranking.rds")

# filter for TFs only
upreg_ubc_tfs <- map(
  .x = upreg_ubc_genes,
  .f = \(x) {
    x <- dplyr::filter(x, feature %in% tfs) %>%
      pull(feature)
    return(x)
  }
)
```

```
> str(upreg_ubc_tfs)
List of 6
 $ UBC_0: chr [1:5] "ZFHX4" "MEIS2" "LMX1A" "EOMES" ...
 $ UBC_1: chr [1:14] "FOXP2" "DACH1" "EBF3" "SOX11" ...
 $ UBC_2: chr [1:22] "ZNF804A" "ZNF521" "ZNF385B" "ZNF804B" ...
 $ UBC_3: chr [1:8] "RFX3" "EOMES" "ZBTB20" "ZFHX4" ...
 $ UBC_4: chr(0) 
 $ UBC_5: chr [1:6] "ZNF423" "EBF3" "ST18" "EBF1" ...
```

Upregulated TFs in MB:

```{r}
# load differential gene expression results for MB subtypes
mb_diff_expr <- read_csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20240917/mb_subtype_diff_exp.csv")

# set factor levels for MB subtypes
mb_diff_expr$cluster <- factor(mb_diff_expr$cluster, levels = mb_subtypes)

# filter for upregulated TFs
upreg_mb_tfs <- map(
  .x = mb_subtypes,
  .f = \(subtype) {
    upreg_tfs <- filter(
      mb_diff_expr,
      cluster == subtype & gene %in% tfs & avg_log2FC > 0 & p_val_adj < 0.05
    ) %>% 
      pull(gene)
    return(upreg_tfs)
  }
) %>%
  setNames(nm = mb_subtypes)
```

```
> str(upreg_mb_tfs)
List of 3
 $ : chr [1:174] "SOX2" "TFDP2" "TSHZ2" "TCF4" ...
 $ : chr [1:224] "ZFHX4" "HLX" "NRL" "NEUROG1" ...
 $ : chr [1:746] "TBR1" "NEUROD1" "SOX4" "ZNF804A" ...
```

Plot the intersection of upregulated TFs in:

- UBC_1 and G4 MB
- UBC_2 and G4 MB

```{r}
tfs_venn <- map(
  .x = paste0("UBC_", 1:2),
  .f = \(ubc_clust) {
    venn_data <- c(upreg_ubc_tfs[ubc_clust], upreg_mb_tfs["G4"]) %>%
      Venn(names = c(ubc_clust, "G4")) %>%
      # use horizontal circles (default is vertical, "201f")
      process_data(shape_id = "201")

    # set colour palette
    fill_cols <- scales::pal_seq_gradient(
      low = my_pals$ubc_integ_clust[ubc_clust],
      high = my_pals$mb_subtype["G4"]
    )(seq(0, 1, length.out = 3)) %>%
      setNames(nm = c("1", "1/2", "2"))

    .plt <- ggplot() + 
      # drawing the areas (insides) of the circles
      geom_polygon(
        mapping = aes(x = X, y = Y, group = id, fill = id),
        data = venn_regionedge(venn_data)
      ) + 
      # drawing the edges of the circles
      geom_path(
        mapping = aes(x = X, y = Y, group = id),
        data = venn_setedge(venn_data),
        colour = "black"
      ) + 
      # label the names of the circles
      geom_text(
        mapping = aes(x = X, y = Y, label = name),
        data = venn_setlabel(venn_data)
      ) + 
      # label the counts for each region
      geom_text(
        mapping = aes(x = X, y = Y, label = count),
        data = venn_regionlabel(venn_data)
      ) + 
      scale_fill_manual(values = fill_cols, guide = "none") + 
      labs(title = "Number of\nupregulated TFs") + 
      # keep fixed aspect ratio
      coord_fixed() + 
      theme_void() + 
      theme(
        plot.background = element_rect(fill = "white", colour = NA),
        plot.title = element_text(
          size = rel(1.2),
          hjust = 0.5,
          margin = margin(b = 5.5)
        ),
        plot.tag = element_text(size = rel(1), face = "bold"),
        plot.tag.location = "plot"
      )

    ggsave(
      filename = paste0(ubc_clust, "_G4_venn.png"),
      plot = .plt,
      path = out_dir,
      width = 3,
      height = 2.75,
      units = "in",
      dpi = 600
    )

    return(.plt)
  }
) %>%
  setNames(nm = c("UBC_1", "UBC_2"))
```


## MB volcano plot for MB subtypes

```{r}
# custom volcano plot function (slightly different from the one in `plotting.R`)
make_volcano_plot <- function(
  data,
  log_fc,
  log_pval,
  direction,
  gene_labels
) {
  .plt <- ggplot(
    data = data,
    mapping = aes(
      x = {{ log_fc }},
      y = {{ log_pval }}
    )
  ) + 
    geom_point(
      aes(colour = {{ direction }}),
      alpha = 0.5,
      shape = "circle",
      size = 2,
      stroke = 0
    ) + 
    ggrepel::geom_text_repel(
      aes(label = {{ gene_labels }}, colour = {{ direction }}),
      size = 2.5,
      min.segment.length = 0.1,
      max.overlaps = Inf,
      show.legend = FALSE,
      seed = 42
    ) + 
    geom_vline(
      xintercept = 0,
      colour = "grey",
      linewidth = 0.25,
      linetype = "dashed"
    ) + 
    geom_hline(
      yintercept = -log10(0.05),
      colour = "grey",
      linewidth = 0.25,
      linetype = "dashed"
    )

  return(.plt)
}
```

```{r}
# set genes with `p_val_adj` of 0 to 1e-300 (the minimum non-zero `p_val_adj` is
# around 3e-294 right now)
min_adj_p <- filter(mb_diff_expr, p_val_adj != 0) %>% pull(p_val_adj) %>% min()

# prep for plotting
mb_diff_expr_tfs <- map(
  .x = mb_subtypes,
  .f = \(subtype) {
    de <- filter(mb_diff_expr, cluster == subtype & gene  %in% tfs) %>%
      mutate(
        direction = case_when(
          avg_log2FC < 0 & p_val_adj < 0.05 ~ "down",
          avg_log2FC > 0 & p_val_adj < 0.05 ~ "up",
          .default = "n.s."
        ),
        p_val_adj = case_when(
          p_val_adj < min_adj_p ~ 1e-300,
          .default = p_val_adj
        ),
        top_tfs_regulons = case_when(
          gene %in% intersect(upreg_ubc_tfs$UBC_1, upreg_mb_tfs[[subtype]]) ~ paste0(subtype, " & UBC_1 TFs"),
          gene %in% intersect(upreg_ubc_tfs$UBC_2, upreg_mb_tfs[[subtype]]) ~ paste0(subtype, " & UBC_2 TFs"),
          gene %in% tail(mb_top_rss[[subtype]]$gene, 18) ~ paste0(subtype, " regulon"),
          .default = NA
        ),
        gene_label = case_when(
          !is.na(top_tfs_regulons) ~ gene,
          .default = ""
        )
      ) %>%
      arrange(gene_label)

    return(de)
  }
) %>%
  setNames(nm = mb_subtypes)
```

```{r}
# volcano plots (all subtypes)
volcano_plots <- map(
  .x = mb_subtypes,
  .f = \(subtype) {
    de <- mb_diff_expr_tfs[[subtype]]

    # label number of genes
    num_up <- nrow(filter(de, direction == "up"))
    num_down <- nrow(filter(de, direction == "down"))
    caption <- paste0(
      "total = ", nrow(de),
      "; down = ", num_down,
      "; up = ", num_up
    )

    .plt <- make_volcano_plot(
      data = de,
      log_fc = avg_log2FC,
      log_pval = -log10(p_val_adj),
      direction = top_tfs_regulons,
      gene_labels = gene_label
    ) + 
      labs(
        caption = caption,
        x = "log2(FC)",
        y = "-log10(adj. p)"
      ) + 
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) + 
      scale_colour_manual(
        values = c(
          my_pals$ubc_integ_clust[c("UBC_1", "UBC_2")],
          my_pals$mb_subtype[subtype]
        ) %>% unname(),
        guide = guide_legend(
          title = "TFs and\nregulons",
          override.aes = list(alpha = 1, size = 3)
        ),
        # don't show NA in legend
        breaks = \(x) {x[!is.na(x)]},
        na.value = "grey"
      ) + 
      theme_classic2() + 
      theme(legend.position = "bottom", legend.box.spacing = unit(1, "points"))

    ggsave(
      filename = paste0(subtype, "_tfs_volcano.png"),
      plot = .plt,
      path = out_dir,
      width = 5,
      height = 5.5,
      units = "in",
      dpi = 600
    )

    return(.plt)
  }
) %>%
  setNames(nm = mb_subtypes)
```


## Final figure

```{r}
layout <- c(
  # tumour UMAP
  area(t = 1, l = 1, r = 7),
  # volcano plot
  area(t = 1, l = 8, r = 20),
  # venn diagrams
  area(t = 2, l = 1, r = 7),
  area(t = 2, l = 8, r = 14),
  # RSS plot
  area(t = 2, l = 15, r = 20)
)

.plt <- wrap_plots(
  mb_subtype_umap,
  volcano_plots$G4 + labs(caption = NULL),
  free(tfs_venn$UBC_1),
  free(tfs_venn$UBC_2),
  free(rss_plots$G4),
  design = layout
) + 
  plot_annotation(tag_levels = list(c(LETTERS[1:3], "", "D")))

# bug in `patchwork` package, need to re-add tag for panel A
.plt[[1]][[1]]$labels$tag <- "A"

walk(
  .x = c("png", "pdf"),
  .f = \(device) {
    ggsave(
      filename = paste0("final.", device),
      plot = .plt,
      path = out_dir,
      width = 9,
      height = 6.75,
      units = "in",
      dpi = 600
    )
  }
)
```
