---
title: "Marker gene expression of RL lineage and control cells"
---

Code for `DotPlot` and `FeaturePlots` showing gene expression of RL lineage and control cells. Note that the code chunks are run interactively using the `scrnaseq_env` conda environment. This file was **not** rendered with `quarto render`.

## Initialization

Packages:

```{r}
library(tidyverse)
library(patchwork)
library(Seurat)
```

Functions/files/directories:

```{r}
thesis_dir <- "../../"
source(file.path(thesis_dir, "./utils.R"))
source(file.path(thesis_dir, "../software/utilities/cell_labelling.R"))

out_dir <- "./rl_gene_expr"
my_pals <- get_custom_pals()
```

Load file and add metadata:

```{r}
# load Seurat object for integrated dataset
srat_qs <- get_srat_paths()
integ_srat <- load_srat(srat_qs["rl"]) %>%
  pluck("rl")

# get cell type annotations (from `cell_labelling.R`)
integ_srat <- label_rl_lineage_integration(integ_srat)

# collapse UBC subclusters into one cluster
integ_srat@meta.data <- mutate(
  integ_srat[[]],
  broad_annot = case_when(
    str_starts(cell_type_annot, "UBC") ~ "UBC",
    .default = cell_type_annot
  ),
  broad_annot = factor(
    broad_annot,
    levels = c("RL", "UBC", "GCP", "GC", "oligodendrocyte/OPC", "microglia", "endothelial")
  ),
  dataset_name = str_remove(
    string = dataset_name,
    pattern = "full_cerebellum_"
  )
)
```


## Preparing integrated data for visualization

The authors mention that the "integrated" assay should not be used for downstream analysis (such as differential expression and visualization). Here, we check two options for visualization gene expression:

- use "RNA" assay (run NormalizeData on raw counts); this seems to be the recommended way for visualizing integrated data (and for performing differential expression)
  - https://satijalab.org/seurat/archive/v3.2/integration
  - https://github.com/satijalab/seurat/issues/1717
  - https://github.com/satijalab/seurat/issues/5881
- use "SCT" assay - this is just another normalization method and apparently it should work for differential expression (and I assume, by extension, visualization); just need to make sure to run `PrepSCTFindmarkers` or to visualize each dataset separately
  - https://satijalab.org/seurat/articles/integration_introduction (see last section regarding integrating SCTransform-normalized datasets)
  - https://github.com/satijalab/seurat/issues/2906
  - see analysis notes from 2024-08-27 where differential expression was run with `FindConservedMarkers` on the SCTransformed datasets - volcano plots and dot plots were generated separately for each of the datasets

### "RNA" assay

The assay looks like it should be normalized, but we'll just re-run it anyways to be sure.

```{r}
# normalize the RNA assay
DefaultAssay(integ_srat) <- "RNA"
integ_srat <- NormalizeData(integ_srat)
```

### "SCT" assay




## Dot plot of RL lineage markers

```{r}
# marker genes for the general cell types; selected from /.mounts/labs/pailab/private/icheong/CBL_scRNAseq/scrnaseq_Leo/utilities/cell_gene_mapping.csv
cell_type_markers <- c(
  "MKI67", # RL, GCP
  "WLS", # RL, UBC progenitors
  "EOMES", "LMX1A", "OTX2", # UBC
  "RBFOX3", "ATOH1", # GCP
  "PAX6", # differentiated cells (UBC/GC)
  "RELN", # GC
  "PDGFRA", "TNR", # OPC/oligodendrocytes
  "TREM2", # microglia
  "CLDN5", "ITM2A" # endothelial
)
```

```{r}
# dot plots
dot_plots <- map(
  .x = list(NULL, "dataset_name"),
  .f = \(split.by) {
    if (is.null(split.by)) {
      col_pal <- c("lightgrey", "blue") # Seurat default
      filename <- "dot_plot_combined.png"
      width <- 5
      height <- 6
    } else if (split.by == "dataset_name") {
      col_pal <- scales::pal_hue()(4)
      filename <- paste0("dot_plot_split_", split.by, ".png")
      width <- 14
      height <- 5
    }

    .plt <- DotPlot(
      integ_srat,
      assay = "RNA",
      features = rev(cell_type_markers),
      cols = col_pal,
      group.by = "broad_annot",
      split.by = split.by
    ) + 
      coord_flip() + 
      theme_classic2() + 
      theme(
        axis.text.x = element_text(hjust = 1, angle = 30)
      )

    ggsave(
      filename = filename,
      plot = .plt,
      path = file.path(out_dir, "rna_assay"),
      width = width,
      height = height,
      units = "in",
      dpi = 600
    )

    return(.plt)
  }
)
```


## Gene expression UMAPs

```{r}
feature_plots <- FeaturePlot(
  integ_srat,
  features = c("EOMES", "LMX1A"),
  order = TRUE,
  min.cutoff = "q10",
  max.cutoff = "q90",
  combine = FALSE
) %>%
  setNames(nm = c("EOMES", "LMX1A")) %>%
  map(.f = \(x) {x + theme_classic2()})

walk2(
  .x = feature_plots,
  .y = names(feature_plots),
  .f = \(.plt, gene) {
    ggsave(
      filename = paste0(gene, ".png"),
      plot = .plt,
      path = file.path(out_dir, "rna_assay"),
      width = 4.5,
      height = 4,
      units = "in",
      dpi = 600
    )
  }
)
```


## Final figure

```{r}
layout <- c(
  area(t = 1, b = 3, l = 1, r = 4),
  area(t = 4, l = 1, r = 2),
  area(t = 4, l = 3, r = 4)
)

.plt <- wrap_plots(
  free(dot_plots[[1]]),
  feature_plots$EOMES,
  feature_plots$LMX1A
) + 
  plot_layout(design = layout) + 
  plot_annotation(tag_levels = "A")

walk(
  .x = c("png", "pdf"),
  .f = \(device) {
    ggsave(
      filename = paste0("final.", device),
      plot = .plt,
      path = out_dir,
      width = 6,
      height = 8,
      units = "in",
      dpi = 600
    )
  }
)
```
