---
title: Comparing top regulons in UBC subclusters and MB tumours
date: 2024-09-29
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To compare the top differentially expressed genes and pySCENIC regulons in UBC subclusters (particularly the human-enriched UBC_1 and UBC_2 cells) to those in medulloblastoma tumours. Similar to the analysis from 2024-08-27, except that the medulloblastoma RSSs were calculated based on the subtypes instead of the cell clusters.

## Files and directories

### Code

All code can be found in this document. The code was run using the `scrnaseq_env` conda environment. The analysis combines results from several previous analyses:

- normal developing RL
  - pySCENIC: `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240704/`
- MB tumours
  - pySCENIC: `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20231018/` and `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240928/`

### Output

The output directory can be found at `<pyscenic results directory>/20240929` with the following output:

- `ranked_ubc_regulons.rds`, list of regulons driving each UBC subcluster with their respective RSS and rank
- `ranked_mb_regulons.rds`, list of top eight regulons driving each MB tumour subtype with their respective RSS and rank
- `num_shared_regulons.csv`, number of common regulons in UBCs and MB
- `shared_regulons_mat.csv`, matrix for generating the heatmap of shared regulons
- `heatmap.png`, final figure


## Results

Load packages, functions, etc.:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(ggrepel)
library(patchwork)
library(Seurat)
library(ComplexHeatmap)

# helper functions
# source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/plotting.R")
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/pyscenic_regulons.R")

# set output directory
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240929/"
```

### Regulons in UBCs and MB

Here, we compare the top regulons driving cell state in cerebellar development (Aldinger/Sepp) and medulloblastoma (Vladoiu).

First, we load and get the top regulons in each cluster.

```{r}
#| code-fold: true

# >>> developing cerebellulm
ubc_reg <- read.csv(
  "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240704/cell_type_annot/aldinger_sepp_RL.rss.csv",
  row.names = 1
)
ubc_subclusters <- str_subset(string = rownames(ubc_reg), pattern = "^UBC") %>%
  sort()

ubc_reg_rank <- map(
  .x = ubc_subclusters,
  .f = \(subcluster) {
    df <- data.frame(
      gene = colnames(ubc_reg),
      score = ubc_reg[subcluster, ] %>% as.numeric(),
      rank = rank(-ubc_reg[subcluster, ], ties.method = "random") %>% as.integer()
    ) %>%
      arrange(rank)

    return(df)
  }
) %>%
  setNames(
    nm = ubc_subclusters
  )

write_rds(
  x = ubc_reg_rank,
  file = file.path(out_dir, "ranked_ubc_regulon.rds")
)
# <<< developing cerebellum

# >>> medulloblastoma
mb_reg <- read.csv(
  "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/pyscenic/20240928/vladoiu_mb_subtype_rss.csv",
  row.names = 1
)
mb_subtypes <- c("SHH", "G3", "G4")

mb_reg_rank <- map(
  .x = mb_subtypes,
  .f = \(subtype) {
    df <- data.frame(
      gene = colnames(mb_reg),
      score = mb_reg[subtype, ] %>% as.numeric(),
      rank = rank(-mb_reg[subtype, ], ties.method = "random") %>% as.integer()
    ) %>%
      arrange(rank)

    return(df)
  }
) %>%
  setNames(
    nm = mb_subtypes
  )

write_rds(
  x = mb_reg_rank,
  file = file.path(out_dir, "ranked_mb_regulons.rds")
)
# <<< medulloblastoma

str(ubc_reg_rank)
str(mb_reg_rank)
```

We then identify the shared regulons across each cell type and plot a heatmap. Previously, we only took the intersection of the top eight regulons. Here, we instead take the intersection of the top 10% of regulons (which works out to the top `r ceiling(0.1*ncol(ubc_reg))`/`r ncol(ubc_reg)` for the UBC subclusters and top `r ceiling(0.1*ncol(mb_reg))`/`r ncol(mb_reg)` for the medulloblastoma tumours).

```{r}
# how many top regulons to take
num_regs <- list(
  ubc = ceiling(0.1*ncol(ubc_reg)),
  mb = ceiling(0.1*ncol(mb_reg))
)

# find shared top regulons in UBCs and MB
num_shared_regs <- expand_grid(ubc_subclusters, mb_subtypes) %>%
  rowwise() %>%
  mutate(
    # get list of shared regulons
    intersection = list(intersect(
      # take `head` because the regulons are already sorted
      ubc_reg_rank[[ubc_subclusters]] %>% pull(gene) %>% head(num_regs$ubc),
      mb_reg_rank[[mb_subtypes]] %>% pull(gene) %>% head(num_regs$mb)
    )),
    # get number of shared regulons
    intersection_size = length(intersection),
    # collapse list of shared regulons into single string
    intersection = case_when(
      intersection_size > 0 ~ paste(intersection, collapse = ", "),
      .default = NA
    )
  )
write_csv(
  x = num_shared_regs,
  file = file.path(out_dir, "num_shared_regulons.csv")
)
```

```{r}
# prep for plotting

# main heatmap matrix
mat <- num_shared_regs %>% 
  select(-intersection) %>% 
  pivot_wider(
    names_from = mb_subtypes,
    values_from = intersection_size
  ) %>% 
  column_to_rownames("ubc_subclusters") %>% 
  as.matrix()
write.csv(
  x = mat,
  file = file.path(out_dir, "shared_regulons_mat.csv"),
  row.names = TRUE
)
```

```{r}
#| label: fig-shared-regs
#| fig-cap: "Heatmap showing the number of regulons that are shared between the UBC subclusters and the MB clusters (out of the top 10% in each respective cluster)."
#| fig-width: 4
#| fig-height: 5

# plotting

# colour function for heatmap
col_fun <- circlize::colorRamp2(
  breaks = c(0, max(mat)),
  hcl_palette = "Viridis"
)

# row/column title format
title_gp <- gpar(fontface = "bold")

# row/column names format
names_gp <- gpar(fontsize = 10)

# row labels
row_labels <- structure(
  rownames(mat),
  names = rownames(mat)
)

# main heatmap
hm <- Heatmap(
  matrix = mat,
  col = col_fun,
  name = "shared\nregulons",
  rect_gp = gpar(col = "white", lwd = 2),
  row_title = "UBC subclusters",
  row_title_side = "left",
  row_title_gp = title_gp,
  column_title = "MB clusters",
  column_title_side = "bottom",
  column_title_gp = title_gp,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_labels = row_labels[rownames(mat)],
  row_names_side = "left",
  row_names_max_width = max_text_width(rownames(mat)),
  row_names_gp = names_gp,
  column_names_rot = 0,
  column_names_gp = names_gp,
  column_names_centered = TRUE,
  width = ncol(mat) * unit(0.5, "in"),
  height = nrow(mat) * unit(0.5, "in")
)
# need `draw()` to calculate heatmap width/height
hm <- draw(hm, merge_legend = TRUE)
# save heatmap to file
png(
  filename = file.path(out_dir, "heatmap.png"),
  width = ComplexHeatmap:::width(hm),
  height = ComplexHeatmap:::height(hm),
  units = "mm",
  res = 600
)
draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()
```

Similar to the results from 2024-08-27, UBC_1 and UBC_4 share the most regulons with Group 4 MB (*SOX4* and *SOX11* for both). The remaining clusters share 0-1 of the top eight regulons. None of the UBC clusters share any top regulons with SHH and Group 3 MB. The full list of common regulons between the UBC clusters and MB subtypes can be found below.

```{r}
filter(num_shared_regs, intersection_size > 0)
```
