---
title: Determine corresponding timepoints across human and mouse cells using pseudoage
date: 2024-05-21
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine human and mouse developmental timepoints/stages correspond best using pseudoage (see Sepp et al. 2024, Extended Data Fig. 2f and methods).

The code from Sepp et al. can be found at [here](https://gitlab.com/kaessmannlab/mammalian-cerebellum). In particular, see `calculate_pseudostage_per_cell.R` and `stage_matching.R`.

Essentially the same as 2024-05-07, but a slightly different integration workflow was used to generate the integrated Seurat object (see below).


## Files and directories

### Code

The code can be found in `<integrated software directory>/calculate_pseudoage.R`. Briefly, this script

1. Subsamples the Sepp mouse dataset to generate a "reference". Each cell in the Sepp mouse dataset was assigned a developmental index based on its real age (e.g. E10.5 = 1, E11.5 = 2, etc.)
1. Determines the 25 nearest reference cells for each cell and calculates the mean developmental index (pseudoage)
1. Bins the cells in each developmental timepoint into 50 bins based on their pseudoage
1. Calculates the proportion of cells in each bin for each developmental timepoint
1. Calculates the Manhattan distance between each developmental timepoint based on the proportion of cells in each bin

The input Seurat object is located in `/.mounts/labs/pailab/private/llau/results/integrated/20240516/fc_subset_analysis/20240514_cca_integ_subset.qs`. This contains the integration of the cells from the full cerebellum followed by subsetting of the RL lineage cells performed by Leo with the following datasets:

- Human - Aldinger et al. (2021), Luo et al. (2022), Sepp et al. (2024)
- Mouse - Vladoiu et al. (2019), Sepp et al. (2024)

#### Difference from 2024-05-07

The previous analysis (from 2024-05-07) used the integration object from 2024-04-15. The integration was performed by first isolating the RL lineage cells then integrating them with CCA. Here, we instead integrate all the cells from the full cerebellum, then subset the RL lineage cells and filter out all human cells before 11 PCW.

To briefly summarize:

```{mermaid}
flowchart TD
    old_date["2024-05-07"] --> old_0[("integration from\n2024-04-15")]
    old_0 --> old_1["isolate RL lineage cells"]
    old_1 --> old_2["CCA integration"]
    new_date["2024-05-21"] --> new_0[("integration from\n2024-05-12")]
    new_0 --> new_1["CCA integration of\nall cerebellar cells"]
    new_1 --> new_2["isolate RL lineage cells,\nfilter out human cells\nbefore 11 PCW"]
    new_2 --> new_3["re-run PCA,\nUMAP, clustering"]
```

### Output

The output directory can be found in `<integrated results directory>/20240521`. The developmental index of each cell in the reference dataset can be found in `sepp_mouse_ref.csv`. The pseudoage of each cell is saved to `pseudoage.csv` (this also contains the actual human/mouse ages of the cells).


## Results

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240521/"
```

### Load packages and Seurat object metadata

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(ComplexHeatmap)
```

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240516/fc_subset_analysis/20240514_cca_integ_subset.qs")
srat
```

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- srat@meta.data
rm(srat)
```

### Manhattan distance for all pairwise developmental stages

Load the file with the Manhattan distances between pairwise developmental stages. This gives ALL pairwise comparisons of the developmental stages, even the intra-species comparisons.

```{r}
pseudoage_dist <- read.csv(
  file = file.path(out_dir, "pseudoage_dist.csv"),
  header = TRUE,
  row.names = 1,
  check.names = FALSE
) %>%
  as.matrix()

pseudoage_dist[1:5, 1:5]
```

### Comparing developmental stages between human and mouse

Show the number of cells at each timepoint:

```{r}
table(md$mouse_age[md$species == "mouse"])
table(md$human_age[md$species == "human"])
```

To compare human vs. mouse developmental stages, we first need to filter out the distance matrix. We remove the human stages from the columns and the mouse stages from the rows.

```{r}
# get human/mouse developmental stages
human_ages <- names(table(md$human_age))
mouse_ages <- names(table(md$mouse_age))

# remove them from distance matrix
hm_dist <- pseudoage_dist[
  !rownames(pseudoage_dist) %in% mouse_ages,
  !colnames(pseudoage_dist) %in% human_ages
]

# sort the columns so "P14" shows up last
hm_dist <- hm_dist[, str_sort(colnames(hm_dist), numeric = TRUE)]
hm_dist
```

Now we use a heatmap to plot the distances between the human and mouse pseudoages.

```{r}
#| label: fig-heatmap-hm-pseudoage
#| fig-cap: "Heatmap showing pairwise Manhattan distances between human and mouse pseudoages."
#| fig-width: 10
#| fig-height: 6

# set colour function for heatmap
breaks <- seq(0, 2, by = 0.25)
col_fun <- circlize::colorRamp2(
  breaks = breaks,
  colors = scales::pal_viridis(direction = -1, option = "inferno")(length(breaks))
)

.plt <- Heatmap(
  matrix = hm_dist,
  col = col_fun,
  row_title = "human",
  row_title_side = "left",
  column_title = "mouse",
  column_title_side = "bottom",
  row_order = rev(rownames(hm_dist)),
  column_order = colnames(hm_dist),
  row_names_side = "left",
  column_names_side = "bottom",
  heatmap_legend_param = list(
    title = "",
    legend_height = unit(4, "in")
  ),
  width = unit(ncol(hm_dist)*0.5, "in"),
  height = unit(nrow(hm_dist)*0.5, "in")
)
.plt
```

### Comparing developmental stages across mouse datasets

The Vladoiu dataset contains mouse cells from integer stages (e.g., E10, E12) while the Sepp dataset contains mouse cells from half stages (e.g., E10.5, E12.5). To see if any of these stages can be combined, we can compare the distances between the mouse stages.

```{r}
# keep only mouse timepoints in the distance matrix
mm_dist <- pseudoage_dist[
  rownames(pseudoage_dist) %in% mouse_ages,
  colnames(pseudoage_dist) %in% mouse_ages
]

mm_dist <- mm_dist[
  # sort the rows so "P14" show up first
  str_sort(rownames(mm_dist), numeric = TRUE), 
  # sort the columns so "P14" shows up last
  str_sort(colnames(mm_dist), numeric = TRUE)
]
mm_dist
```

Now we use a heatmap to plot the distances between the mouse pseudoages.

```{r}
#| label: fig-heatmap-mm-pseudoage
#| fig-cap: "Heatmap showing pairwise Manhattan distances between mouse pseudoages."
#| fig-width: 10
#| fig-height: 10

.plt <- Heatmap(
  matrix = mm_dist,
  col = col_fun, # same colour function from before
  row_title = "mouse",
  row_title_side = "left",
  column_title = "mouse",
  column_title_side = "bottom",
  row_order = rev(rownames(mm_dist)),
  column_order = colnames(mm_dist),
  row_names_side = "left",
  column_names_side = "bottom",
  heatmap_legend_param = list(
    title = "",
    legend_height = unit(4, "in")
  ),
  width = unit(ncol(mm_dist)*0.5, "in"),
  height = unit(nrow(mm_dist)*0.5, "in")
)
.plt
```
