---
title: Comparing GetResidual vs ScaleData for the SCT assay
date: 2024-11-03
metadata-files:
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To compare heatmaps generated using GetResidual vs ScaleData (for the SCT assay).

Briefly, Seurat's `DoHeatmap` function uses the object's `scale.data` slot. For the regular RNA assay, `scale.data` contains the log-normalized and scaled counts. However, in the SCT assay, `scale.data` contains the Pearson residuals, while the `counts` and `data` slots hold the corrected counts and the log1p(corrected counts).

This is important because, by default, not all genes are included in the `scale.data` slot (to save memory), and so some genes won't be plotted when `DoHeatmap` is run. In order for the gene to be plotted on the heatmap, it needs to be added to `scale.data`, so for SCT assay the `GetResidual` function could be used to calculate the residuals for the missing genes. Based on previous posts (see [1](https://github.com/satijalab/seurat/issues/2050), [2](https://github.com/satijalab/seurat/issues/2503), and [3](https://github.com/satijalab/seurat/issues/2399)), plotting the residuals from the SCT assay should be valid.

On the other hand, since `data` in the SCT assay contains log-normalized corrected counts, you should theoretically be able to scale and centre those counts with `ScaleData` so that genes have mean expression of 0 and standard deviation of 1. However, there doesn't seem to be any posts/threads that talk about doing it this way.

All that being said, to avoid the issue with genes not being found in the `scale.data` slot, you could just run `SCTransform` with `return.only.var.genes = FALSE` to return the residuals for ALL genes; however, this will probably take longer than normal to run.


## Files and directories

### Code

The code can be found in this document and all code was run using the `scrnaseq_env` conda environment. Shraddha's differential expression results from 2024-10-29 are used here. These show much greater log2FC than Leo's results from 2024-07-15, although the reason for that difference still needs to be investigated (Shraddha said she copied the code from Leo, so in theory, it should all be the same).

### Output

The output directories can be found at `<integrated results directory>/20241103`.


## Results

- load the Seurat object
- run GetResidual for all genes
  - compare genes that were already there
- run ScaleData for all genes
- compare heatmap of GetResidual vs ScaleData

Load packages and functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(ComplexHeatmap)

# set output directories
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241103/"
```

Load and merge Aldinger/Sepp Seurat objects. `SCTransform` has already been run separately on each of the datasets and we just need to merge them and run `PrepSCTFindMarkers`. (See notes from 2024-08-27 for more details; these were based on Leo's workflow from 2024-07-15.)

```{r}
#| cache: false
#| cache-lazy: false

merge_srat_file <- file.path(out_dir, "human_ubc_srat_merged.qs")
if (!file.exists(merge_srat_file)) {
  message("***Running `merge` and `PrepSCTFindMarkers`***")
  ubc_srat <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240827/diff_expr_markers/human_ubc_srat_list.qs")

  ubc_srat <- merge(ubc_srat[[1]], ubc_srat[[2]]) %>%
    PrepSCTFindMarkers()

  # add sex to metadata
  sex_of_samples <- read_csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/scrnaseq_Leo/utilities/dataset_sexes.csv")
  ubc_srat@meta.data <- ubc_srat@meta.data %>%
    mutate(sex = case_when(
      dataset_name == "Aldinger_full_cerebellum_human" ~ sex,
      dataset_name == "Sepp_full_cerebellum_human" ~ sex_of_samples$sex[
        match(orig.ident, sex_of_samples$orig.ident)
      ]
    ))

  qs::qsave(x = ubc_srat, file = merge_srat_file)
} else {
  message(paste0("***Reading in ", merge_srat_file, "***"))
  ubc_srat <- qs::qread(merge_srat_file)
}

ubc_plot_order <- paste0("UBC_", c(5,1,2,4,0,3)) # plotting order
```

### Run `GetResidual`

Calculate residuals for all genes:

```{r}
#| cache: false
#| cache-lazy: false

# get residuals for all genes
srat_residuals <- GetResidual(
  ubc_srat,
  features = rownames(ubc_srat),
  assay = "SCT",
  replace.value = FALSE
)
```

### Run `ScaleData`

Scale the log1p(corrected counts) of the SCT assay:

```{r}
#| cache: false
#| cache-lazy: false

# scale all genes
srat_scale <- ScaleData(
  ubc_srat,
  features = rownames(ubc_srat),
  assay = "SCT",
  vars.to.regress = "CC.Difference"
)
```

### Get top differentially expressed genes in each cluster

Use markers with nominal p-value < 0.05 in both datasets and a mean(Aldinger log2FC, Sepp log2FC) > 1.5. Take up to 100 genes.

```{r}
spai_diff_expr <- read.delim("/.mounts/labs/pailab/private/projects/HumanMouseUBC/UBC1_marker_analysis/Integrated_RL_Hsonly/DEG_notfiltered_241029.txt", sep = "\t") %>%
  rename_with(.fn = \(x) {str_remove(string = x, pattern = "full_cerebellum_")})

top_genes <- map(
  .x = ubc_plot_order,
  .f = \(clust) {
    df <- filter(
      spai_diff_expr,
      cluster == clust & Aldinger_human_p_val < 0.05 & Sepp_human_p_val < 0.05
    ) %>%
      mutate(
        avg_log2FC = (Aldinger_human_avg_log2FC + Sepp_human_avg_log2FC) / 2
      ) %>%
      arrange(desc(avg_log2FC)) %>%
      filter(avg_log2FC > 1.5)

    # how many top genes to take
    nr <- min(nrow(df), 100)
    df <- slice_head(df, n = nr)

    return(df)
  }
) %>%
  setNames(nm = ubc_plot_order)
unique_markers <- map(.x = top_genes, .f = \(x) {pull(x, feature)}) %>%
  unlist() %>%
  unique()

# save top markers
write_csv(
  x = do.call(what = "rbind", args = top_genes),
  file = file.path(out_dir, "top_ubc_markers.csv")
)

# show the top genes
walk(
  .x = ubc_plot_order,
  .f = \(clust) {
    message(sprintf("Top markers in the %s:", clust))
    print(head(top_genes[[clust]]$feature, 10))
  }
)
```

### Plot heatmaps with top differentially expressed genes

```{r}
#| cache: false
#| cache-lazy: false

srat_residuals$subclusters <- factor(
  srat_residuals$subclusters,
  levels = ubc_plot_order
)
srat_scale$subclusters <- factor(
  srat_scale$subclusters,
  levels = ubc_plot_order
)
```

```{r}
#| label: fig-srat-hm-resid-cell
#| fig-cap: "Heatmap of top markers with their Pearson residuals (from the SCT assay with `GetResidual`) for each cell."
#| fig-width: 10
#| fig-height: 10
#| cache: false
#| cache-lazy: false
#| code-fold: true

.plt <- DoHeatmap(
  srat_residuals,
  features = unique_markers,
  group.by = "subclusters",
  assay = "SCT",
  slot = "scale.data",
  size = 3
) + 
  labs(title = "Heatmap of residuals (GetResidual)") + 
  # don't show gene names, it's too crowded
  theme(axis.text.y = element_blank())

ggsave(
  filename = "hm_residual_by_cell.png",
  plot = .plt,
  path = out_dir,
  width = 10,
  height = 10,
  units = "in",
  dpi = 300
)

.plt
```

```{r}
#| label: fig-srat-hm-scale-cell
#| fig-cap: "Heatmap of top markers with scaled log1p(corrected counts) (from the SCT assay with `ScaleData`) for each cell."
#| fig-width: 10
#| fig-height: 10
#| cache: false
#| cache-lazy: false
#| code-fold: true

.plt <- DoHeatmap(
  srat_scale,
  features = unique_markers,
  group.by = "subclusters",
  assay = "SCT",
  slot = "scale.data",
  size = 3
) + 
  labs(title = "Heatmap of scaled log1p(corrected counts) (ScaleData)") + 
  # don't show gene names, it's too crowded
  theme(axis.text.y = element_blank())

ggsave(
  filename = "hm_scaledata_by_cell.png",
  plot = .plt,
  path = out_dir,
  width = 10,
  height = 10,
  units = "in",
  dpi = 300
)

.plt
```

Comparing the heatmaps of the residuals (@fig-srat-hm-resid-cell) to the scaled log(corrected counts) (@fig-srat-hm-scale-cell), there does not appear to be much of a difference between the two.

```{r}
#| code-fold: true

# annotations for ComplexHeatmap (adapted from Shraddha's `UBCmarkerHeatmap.R`)

# create top annotation rows
ages <- ubc_srat[[]] %>%
  select(subclusters, human_age) %>%
  table() %>%
  prop.table(margin = 1) %>%
  .[ubc_plot_order, ]
sex <- ubc_srat[[]] %>%
  select(subclusters, sex) %>%
  table() %>%
  prop.table(margin = 1) %>%
  .[ubc_plot_order, ]
dset <- ubc_srat[[]] %>%
  select(subclusters, dataset_name) %>%
  mutate(dataset_name = str_remove(dataset_name, "_full_cerebellum_human")) %>%
  table() %>%
  prop.table(margin = 1) %>%
  .[ubc_plot_order, ]
pal1 <- RColorBrewer::brewer.pal(ncol(ages), "YlOrRd") # age
pal2 <- c("pink","blue") # sex
pal3 <- RColorBrewer::brewer.pal(3, "Dark2") # dataset

topannot <- HeatmapAnnotation(
  age = anno_barplot(
    ages, 
    gp = gpar(fill = pal1, border="white", lty="solid"),
    bar_width = 1, 
    height = unit(2, "in"),
    axis=FALSE
  ),
  sex = anno_barplot(
    sex,
    gp = gpar(fill = pal2, border="white"),
    bar_width = 1,
    height = unit(1,"in"),
    axis=FALSE
  ),
  dataset = anno_barplot(
    dset,
    gp = gpar(fill = pal3, border="white"),
    bar_width = 1,
    height = unit(1,"in"),
    axis=FALSE
  )
)
lgd_list <- list(
  Legend(labels=colnames(ages),title="age",legend_gp=gpar(fill=pal1)),
  Legend(labels=colnames(sex),title="sex",legend_gp=gpar(fill=pal2)),
  Legend(labels=colnames(dset),title="dataset",legend_gp=gpar(fill=pal3))
)

# genes to annotate
annoGenes <- map(
  .x = top_genes,
  .f = \(df) {
    head(df$feature, 5)
  }
) %>%
  unlist() %>%
  unname()
```

```{r}
#| label: fig-srat-hm-resid-clust
#| fig-cap: "Heatmap of top markers with their average Pearson residuals (from the SCT assay with `GetResidual`) in each cluster."
#| fig-width: 8
#| fig-height: 12
#| cache: false
#| cache-lazy: false
#| code-fold: true

# get average of Pearson residuals
xpr <- AverageExpression(
  srat_residuals,
  assays = "SCT",
  features = unique_markers,
  group.by = "subclusters",
  slot = "scale.data"
)[[1]]

# heatmap colours
col_fun <- circlize::colorRamp2(
  breaks = c(-2,0,2),
  colors = c("#FF00FF","#000000", "#FFFF00")
)

ha <- rowAnnotation(
  foo = anno_mark(at = match(annoGenes, rownames(xpr)), labels = annoGenes)
)

hm <- Heatmap(
  xpr,
  name="avg_pearson_resid",
  cluster_rows=FALSE,
  cluster_columns=FALSE,
  col = col_fun,
  show_row_names = FALSE,
  column_title="SCT residuals (GetResidual)",
  top_annotation = topannot,
  right_annotation = ha
)

png(
  filename = file.path(out_dir, "hm_residual_by_cluster.png"),
  width = 8,
  height = 12,
  units = "in",
  res = 300
)
draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()

draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()
```

```{r}
#| label: fig-srat-hm-scale-clust
#| fig-cap: "Heatmap of top markers with scaled log1p(corrected counts) (from the SCT assay with `ScaleData`) for each cluster."
#| fig-width: 8
#| fig-height: 12
#| cache: false
#| cache-lazy: false
#| code-fold: true

# get scale.data slot
xpr <- AverageExpression(
  srat_scale,
  assays = "SCT",
  features = unique_markers,
  group.by = "subclusters",
  slot = "scale.data"
)[[1]]

# heatmap colours
col_fun <- circlize::colorRamp2(
  breaks = c(-2,0,2),
  colors = c("#FF00FF","#000000", "#FFFF00")
)

ha <- rowAnnotation(
  foo = anno_mark(at = match(annoGenes, rownames(xpr)), labels = annoGenes)
)

hm <- Heatmap(
  xpr,
  name="avg_scaledata",
  cluster_rows=FALSE,
  cluster_columns=FALSE,
  col = col_fun,
  show_row_names = FALSE,
  column_title="SCT + ScaleData",
  top_annotation = topannot,
  right_annotation = ha
)

png(
  filename = file.path(out_dir, "hm_scaledata_by_cluster.png"),
  width = 8,
  height = 12,
  units = "in",
  res = 300
)
draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()

draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()
```

```{r}
#| label: fig-srat-hm-scale2-clust
#| fig-cap: "Heatmap of top markers with manually scaled log1p(corrected counts) for each cluster."
#| fig-width: 8
#| fig-height: 12
#| cache: false
#| cache-lazy: false
#| code-fold: true

# get data slot, then manually scale the results
xpr <- AverageExpression(
  srat_scale,
  assays = "SCT",
  features = unique_markers,
  group.by = "subclusters",
  slot = "data"
)[[1]]
xpr <- t(scale(t(xpr)))

# heatmap colours
col_fun <- circlize::colorRamp2(
  breaks = c(-2,0,2),
  colors = c("#FF00FF","#000000", "#FFFF00")
)

ha <- rowAnnotation(
  foo = anno_mark(at = match(annoGenes, rownames(xpr)), labels = annoGenes)
)

hm <- Heatmap(
  xpr,
  name="manual_scale",
  cluster_rows=FALSE,
  cluster_columns=FALSE,
  col = col_fun,
  show_row_names = FALSE,
  column_title="SCT data slot then manual scale()",
  top_annotation = topannot,
  right_annotation = ha
)

png(
  filename = file.path(out_dir, "hm_scale_by_cluster.png"),
  width = 8,
  height = 12,
  units = "in",
  res = 300
)
draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()

draw(hm, annotation_legend_list = lgd_list, merge_legend = TRUE)
dev.off()
```

