---
title: Sex differences in the human cerebellum
date: 2024-03-04
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

- Do male and female samples separate out in the RL-VZ/SVZ?
- What genes are differentially expressed in males vs. females?


## Files and directories

### Code

All code can be found in this document and in `<human Aldinger results directory>/20240304/diff_gene_expr_by_sex.R`.

Load packages:

```{r}
#| cache: false
library(tidyverse)
library(ggrepel)
library(Seurat)
```

Load additional functions:

```{r}
source("/CBL_scRNAseq/software/utilities/plotting.R")
```


### Output

All output files can be found in `<human Aldinger results directory>/20240304`. Genes with a negative/positive log2FC are upregulated in females/males.

```{r}
out_dir <- "/CBL_scRNAseq/results/human/Aldinger/20240304/"
```

Figures can be found in this document.


## Data

Load Aldinger dataset (RL lineage only from Hendrikse).

```{r}
#| cache: false
#| cache-lazy: false
srat <- readRDS("/isilon/CBL_scRNAseq-archived/data/human/Aldinger/glutamatergic_dev_Liam.RDS")
```


## Results

### UMAPs

```{r}
#| echo: false
#| label: fig-umap-sex
#| fig-cap: "UMAP of RL lineage showing cells from female (left panel) and male (right panel) samples. Cells are coloured by cell type."
#| fig-width: 10

DimPlot(srat, split.by = "sex")
```

### Differential gene expression by sex

```{r}
#| echo: false
#| output: asis

cell_types <- levels(srat$new_cell_type)

purrr::walk(
  .x = cell_types,
  .f = \(cell_type) {
    cat("#### ", cell_type, "\n", sep = "")
    
    fname <- paste0(
      str_replace(cell_type, " |-", ""),
      "_MvF.csv"
    )
    de_genes <- read_csv(
      file = file.path(out_dir, fname),
      show_col_types = FALSE # suppress warning messages
    )
    
    top_genes <- arrange(de_genes, p_val_adj) %>% 
      .$gene_name %>% 
      head(25)
    
    de_genes <- de_genes %>% 
      mutate(
        label = case_when(
          gene_name %in% top_genes ~ gene_name,
          TRUE ~ NA_character_
        )
      )
    
    res <- knitr::knit_child(
      text = c(
        "```{r}",
        "print(de_genes, n = 25)",
        "```",
        "",
        "A total of `r nrow(de_genes[de_genes$p_val < 0.05, ])` genes are differentially expressed (*p*-value < 0.05). The top 25 genes are shown above.",
        "",
        "```{r}",
        "#| echo: false",
        "#| fig-width: 8",
        "#| fig-height: 8",
        "#| fig-cap: !expr glue::glue('Volcano plot of differentially expressed genes in male vs. female {cell_type}. A positive logFC indicates genes are more highly expressed in males. Unadjusted *p*-values are shown here.')",
        ".plt <- make_volcano(de_genes, log_fc = avg_log2FC, log_pval = -log10(p_val), gene_labels = label, log_fc_thresh = c(-0.5, 0.5), log_pval_thresh = -log10(0.05)) + ",
        "  labs(title = '<-- Female | Male --> ', x = 'log2(FC)', y = '-log10(p-value)')",
        "print(.plt)",
        "```"
      ),
      envir = environment(),
      quiet = TRUE
    )
    
    cat(res, sep = "\n")
    cat("\n")
  }
)
```

