---
title: Comparing normal vs. tumour BRCA1 regulon genes
date: 2023-10-06
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

Previously, BRCA1 was shown to be a top regulon in the RL-SVZ of the normal developing hindbrain and in some Group 4 tumour MB clusters.

Here, the aim is to compare the genes in the BRCA1 regulon in normal vs. tumour.


## Files and directories

### Code

Code can be found in this notebook and in `<pyscenic software directory>/brca1_gost_enrichment.R`.

Normal developing hindbrain regulons are from 2023-08-01.

MB tumour regulons are from 2023-10-05.

The script runs GO/pathway enrichment analysis on the target genes of the BRCA1 regulons.

### Output

Output can be found in this notebook and in `<pyscenic results directory>/20231006`.

The `normal` and `tumour` subdirectories contain the pathway enrichment results for the normal and tumour BRCA1 regulons, respectively.


## Results

### Common target genes

```{r}
#| eval: true
#| echo: true
#| output: false

library(tidyverse)
library(gprofiler2)
```

First, import the BRCA1 regulons.

```{r}
#| eval: true
#| echo: true

rlsvz_brca1 <- read_csv("/CBL_scRNAseq/results/pyscenic/20230801/BRCA1.csv")
g4mb_brca1 <- read_csv("/CBL_scRNAseq/results/pyscenic/20231005/BRCA1.csv")
```

We can see that the BRCA1 regulon in the developing cerebellum contains `r nrow(rlsvz_brca1)` target genes. On the other hand, the tumour BRCA1 regulon only contains `r nrow(g4mb_brca1)` target genes.

Of the target genes, only `r length(intersect(rlsvz_brca1$gene, g4mb_brca1$gene))` are common across both regulons:

```{r}
#| eval: true
#| echo: true

intersect(rlsvz_brca1$gene, g4mb_brca1$gene)
```

### Pathway enrichment analysis

Enrichment analysis was carried out using g:Profiler with the Bader lab's custom "GOBP All Pathways no IEA" gene set from August 8, 2023 as the custom GMT.

```{r}
#| eval: true
#| echo: false
#| label: fig-normal-manhattan
#| fig-cap: "Enriched GO terms and pathways in the normal BRCA1 regulon."

read_rds("/CBL_scRNAseq/results/pyscenic/20231006/normal/normal.gost.rds") %>% 
  gostplot(pal = "#FF9900")
```

```{r}
#| eval: true
#| echo: false
#| label: fig-tumour-manhattan
#| fig-cap: "Enriched GO terms and pathways in the MB BRCA1 regulon."

read_rds("/CBL_scRNAseq/results/pyscenic/20231006/tumour/tumour.gost.rds") %>% 
  gostplot(pal = "#FF9900")
```

