---
title: Run UMAP for subsetted UBCs
date: 2024-08-25
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
execute:
  eval: false
---

## Goal of this analysis

To generate a UMAP showing only the subsetted UBCs. See also Leo's results on 2024-07-31 (however, the Seurat object containing the UMAP embeddings was not saved).


## Files and directories

### Code

All code can be found in this document. The code was run using the `scrnaseq_env` conda environment.

- Code to perform UMAP dimensionality reduction was run interactively and not during rendering (`eval: false` in the YAML header prevents code from running by default)
- Plot are generated upon rendering (`eval: true` overrides the default in certain code chunks)

The input Seurat file can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs`.

### Output

The output directory can be found at `<integrated results directory>/20240825` with the following output:

- `ubc_subset.qs` contains the Seurat object with the new UMAP embeddings
- `*_umap.png` show the UMAPs with cells coloured by different metadata columns


## Results

### Load packages

Load packages and other functions:

```{r}
#| eval: true
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(Seurat)
```

Set output directory:

```{r}
#| eval: true

out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240825/"
```

### Run UMAP

Load Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs")

# make of UBC clusters
srat$subclusters <- paste0("UBC_", srat$snn_res.0.3)
```

Run UMAP with the top 25 PCs (PCs are from the integration of all the datasets):

```{r}
#| cache: false
#| cache-lazy: false

srat <- RunUMAP(srat, reduction = "pca", dims = 1:25)
```

Save the Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

qs::qsave(x = srat, file = file.path(out_dir, "ubc_subset.qs"))
```

### Plot UMAP

First, load the new Seurat object with the new UMAP embeddings (unless it's already been loaded).

```{r}
#| eval: true
#| cache: false
#| cache-lazy: false

if (!exists("srat")) {
  srat <- qs::qread(file.path(out_dir, "ubc_subset.qs"))
}

# clean up dataset column
srat$dataset <- str_replace(
  string = srat$dataset_name,
  pattern = "_full_cerebellum_([:alpha:]+)",
  replacement = " (\\1)"
)
```

Show the UMAP plots:

```{r}
#| eval: true
#| label: fig-ubc-umaps
#| fig-cap: "UMAP visualization of the UBCs with cells coloured by (A) UBC subclusters, (B) species, and (C) dataset."
#| fig-width: 6
#| fig-height: 12
#| cache: false
#| cache-lazy: false
#| code-fold: true

# coloured by UBC subclusters
.plt1 <- DimPlot(
  srat,
  reduction = "umap",
  group.by = "subclusters",
  cols = pals::brewer.set2(6),
  label = TRUE,
  repel = TRUE
) + 
  labs(title = "UBC subclusters")
ggsave(
  filename = "subclusters_umap.png",
  plot = .plt1,
  path = out_dir,
  width = 5.5,
  height = 5,
  units = "in",
  dpi = 600
)

# coloured by species
.plt2 <- DimPlot(
  srat,
  reduction = "umap",
  cells.highlight = WhichCells(srat, expression = species == "human"),
  sizes.highlight = 0.1,
  label = FALSE
) + 
  labs(title = "species") + 
  scale_colour_manual(
    labels = c("mouse", "human"),
    values = c("grey", scales::pal_hue()(1)),
  )
ggsave(
  filename = "species_umap.png",
  plot = .plt2,
  path = out_dir,
  width = 5.5,
  height = 5,
  units = "in",
  dpi = 600
)

# coloured by datasets
.plt3 <- DimPlot(
  srat,
  reduction = "umap",
  group.by = "dataset",
  label = FALSE
) + 
  labs(title = "dataset")
ggsave(
  filename = "dataset_umap.png",
  plot = .plt3,
  path = out_dir,
  width = 6,
  height = 5,
  units = "in",
  dpi = 600
)

# show plot
wrap_plots(
  .plt1, .plt2, .plt3,
  ncol = 1
) + 
  plot_annotation(tag_levels = "A") & 
  theme(plot.title = element_text(hjust = 0.5))
```
