---
title: Comparing differential expression results using SCT vs RNA assay
date: 2024-10-19
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To compare differential gene expression results when using the log-normalized counts ("RNA" assay) vs. the SCTransform-normalized and log-corrected counts ("SCT" assay). This is to ensure the results look similar Differential expression was run for each of the UBC subclusters.


## Files and directories

### Code

The code can be found in this document. Differential gene expression was run by Leo and results were taken from the following directories:

- SCT assay results: `/.mounts/labs/pailab/private/llau/results/integrated/20240715` (volcano plots in 20240719 instead)
- RNA assay results: `/.mounts/labs/pailab/private/llau/results/integrated/20241018`

See Leo's notebook for the differential gene expression code.

### Output

The output can be found in `<integrated results directory>/20241019`.


## Results

Load packages and files:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)

out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241019"

# SCT results
sct_de <- read_csv("/.mounts/labs/pailab/private/llau/results/integrated/20240715/all_tested_genes.csv") %>%
  rename_with(.fn = \(x) {str_remove(string = x, pattern = "full_cerebellum_")})

# RNA results
rna_de <- read_csv("/.mounts/labs/pailab/private/llau/results/integrated/20241018/all_tested_genes.csv") %>%
  rename_with(.fn = \(x) {str_remove(string = x, pattern = "full_cerebellum_")})
```

Functions:

```{r}
merge_sct_rna <- function(
  sct,
  rna,
  column_name,
  column_is_pval = FALSE
) {
  sct <- sct[, c("feature", column_name)]
  rna <- rna[, c("feature", column_name)]

  if (column_is_pval) {
    sct[, column_name] <- -log10(sct[, column_name])
    rna[, column_name] <- -log10(rna[, column_name])
  }

  df <- inner_join(
    x = sct,
    y = rna,
    by = "feature",
    suffix = c("sct", "rna")
  ) %>%
    rename_with(.fn = \(x) {str_remove(string = x, pattern = column_name)})
  return(df)
}

make_scatter <- function(
  data
) {
  .plt <- ggplot(data = data, mapping = aes(x = sct, y = rna)) + 
    geom_point(colour = "grey25") + 
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", linewidth = 0.5, colour = "grey50") + 
    labs(x = "SCT assay", y = "RNA assay")

  return(.plt)
}
```

Make plots to compare the RNA vs. SCT logFC and adjusted p-values:

```{r}
ubc_subclusters <- paste0("UBC_", 0:5)

walk(
  .x = ubc_subclusters,
  .f = \(clust) {
    sct <- filter(sct_de, cluster == clust)
    rna <- filter(rna_de, cluster == clust)

    # get genes tested in both (can't plot if gene was only tested in one)
    # genes <- intersect(sct$feature, rna$feature)

    aldi_logfc <- make_scatter(
      data = merge_sct_rna(
        sct = sct,
        rna = rna,
        column_name = "Aldinger_human_avg_log2FC"
      )
    ) + 
      geom_vline(xintercept = 0, linetype = "solid", linewidth = 0.75) + 
      geom_hline(yintercept = 0, linetype = "solid", linewidth = 0.75) + 
      labs(title = "Aldinger log2FC")

    aldi_qval <- make_scatter(
      data = merge_sct_rna(
        sct = sct,
        rna = rna,
        column_name = "Aldinger_human_p_val_adj",
        column_is_pval = TRUE
      )
    ) + 
      labs(title = "Aldinger -log10(adj. p-value)")

    sepp_logfc <- make_scatter(
      data = merge_sct_rna(
        sct = sct,
        rna = rna,
        column_name = "Sepp_human_avg_log2FC"
      )
    ) + 
      geom_vline(xintercept = 0, linetype = "solid", linewidth = 0.75) + 
      geom_hline(yintercept = 0, linetype = "solid", linewidth = 0.75) + 
      labs(title = "Sepp log2FC")

    sepp_qval <- make_scatter(
      data = merge_sct_rna(
        sct = sct,
        rna = rna,
        column_name = "Sepp_human_p_val_adj",
        column_is_pval = TRUE
      )
    ) + 
      labs(title = "Sepp -log10(adj. p-value)")

    .plt <- wrap_plots(
      aldi_logfc, aldi_qval, sepp_logfc, sepp_qval,
      ncol = 2,
      byrow = TRUE
    ) & 
      theme_minimal() & 
      plot_annotation(
        title = clust,
        theme = theme(
          plot.title = element_text(face = "bold", hjust = 0.5),
          plot.background = element_rect(fill = "white", linewidth = 0)
        )
      )

    ggsave(
      filename = paste0(clust, ".png"),
      plot = .plt,
      path = out_dir,
      width = 8,
      height = 8,
      units = "in",
      dpi = 600
    )
  }
)
```
