---
title: Upregulated genes in non-homologous UBCs and mutated genes in MB
date: 2023-02-23
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
knitr:
  opts_chunk: 
    output: false
---

## Goal of this analysis

Last week, the TF motifs enriched in non-homologous UBCs were identified and linked to mutated genes in MB. Here, we want to include the upregulated non-homologous UBC genes in the comparison as well. Which of the upregulated genes in the non-homologous UBCs are mutated in MB?


## Files and directories

### Code

See `<human Aldinger results directory>/​20230215/​enriched_motifs_nonhomol.csv` for the TF motifs that were enriched in non-homologous UBC genes.

See `<human Aldinger results directory>/​20230205/​de_genes.csv` for the genes that are upregulated in non-homologous UBCs compared to homologous UBCs.

See `<isilon human Hendrikse data directory>/​supp_tab_5.xlsx` for the table of mutations in G3 and G4 MB.

### Output

All output can be found in this document.


## Results

```{r}
library(tidyverse)
```

### Import data and clean-up

Import the MB-mutated genes (see notes from 2023-02-16 for a more detailed explanation).

```{r}
source("/CBL_scRNAseq/software/human/Hendrikse/load_data.R")
mut_genes <- load_mut_genes(
  mut_xlsx = "/isilon/CBL_scRNAseq-archived/data/human/Hendrikse/supp_tab_5.xlsx", # default
  rename_cols = TRUE # default
)

sig_mut_genes <- mut_genes %>% 
  mutate(
    g4_mutated = case_when(
      qval_oncodrive_g4 < 0.05 | qval_mutsig_g4 < 0.05 ~ TRUE,
      TRUE ~ FALSE
    ),
    g3_mutated = case_when(
      qval_oncodrive_g3 < 0.05 | qval_mutsig_g3 < 0.05 ~ TRUE,
      TRUE ~ FALSE
    ),
    mutated_in = case_when(
      g4_mutated & g3_mutated ~ "both",
      g4_mutated ~ "G4",
      g3_mutated ~ "G3",
      TRUE ~ NA_character_
    )
  ) %>% 
  select(gene, Freq_whole_cohort, freq_g4, freq_g3, mutated_in) %>% 
  filter(!is.na(mutated_in))
```

The final result:

```{r}
#| output: true

sig_mut_genes
```

Import the differentially expressed genes from 2023-02-05.

```{r}
#| output: true

de_genes <- read_csv("/CBL_scRNAseq/results/human/Aldinger/20230205/de_genes.csv")

print(de_genes)
```

Let's select for genes that are upregulated in the non-homologous UBCs.

```{r}
nhubc_genes <- filter(
  de_genes,
  p_val_adj < 0.05 & avg_log2FC < 0
)
```


Import the enriched TFs from 2023-02-15.

```{r}
#| output: true

tf_genes <- read_csv(
  "/CBL_scRNAseq/results/human/Aldinger/20230215/enriched_motifs_nonhomol.csv"
) %>% 
  mutate(
    gene = str_split(motif_id, "_", simplify = TRUE)[,1],
    .after = rank
  ) %>% 
  select(-c(motif_db, motif_id, motif_alt_id, consensus))

tf_genes
```

### Compare mutated genes from MB with the upregulated genes and enriched TFs

Supplementary Table 5 only contains some of the mutated genes listed in Figure 1 (looks like it's only small mutations like SNVs and INDELs; e.g. PRDM6 is missing from the file). We need to manually add the genes listed in Figure 1 of the paper.

```{r}
additional_genes <- c(
  "PRDM6",
  "CBFA2T3",
  "KDM6A",
  "RUNX1T1",
  "CBFA2T2",
  "KDM2B",
  "KBTBD4",
  "SMARCA4",
  "KMT2D",
  "KMT2C",
  "SETD2",
  "SRCAP",
  "ZMYM3",
  "EHMT1",
  "CREBBP",
  "EHMT2",
  "KAT6A",
  "KDM6B",
  "KMT2A",
  "CHD7",
  "CHD9",
  "CHD8",
  "CHD3",
  "CHD4",
  "CHD6",
  "CHD2",
  "CHD5",
  "GFI1B",
  "GFI1",
  "OTX2",
  "ZIC1",
  "TBR1",
  "MYC",
  "MYCN",
  "PVT1",
  "CTDNEP1",
  "CDK6",
  "PKD1",
  "STARD9",
  "ASPM",
  "HTT",
  "ANKRD17",
  "EIF2AK4",
  "TSC2",
  "ATM",
  "FANCA",
  "FANCL",
  "PALB2",
  "FANCI",
  "BRCA2",
  "FANCB",
  "FANCG",
  "FANCM",
  "ZFHX2",
  "ZFHX3",
  "ZFHX4",
  "IKBKAP",
  "ELP2",
  "ELP4",
  "DST"
)

all_mut_genes <- union(
  sig_mut_genes$gene,
  additional_genes
)
```

We can now compare the genes that are upregulated in non-homologous UBCs or have enriched TF motifs with the genes that are mutated in MB.

```{r}
#| output: true

common_genes <- intersect(
  union(nhubc_genes$gene, tf_genes$gene),
  all_mut_genes
)

common_genes
```

