---
title: Exploring the Luo et al. 2022 single-cell dataset
date: 2024-03-07
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
editor_options: 
  chunk_output_type: inline
---

## Goal of this analysis

Exploratory analysis of the [Luo et al. 2022](https://doi.org/10.1038/s41586-022-05487-2) (PMID: [36450980](https://pubmed.ncbi.nlm.nih.gov/36450980/)) single-cell dataset to determine what additional downstream processing will be needed prior to integration with other datasets.


## Files and directories

All code and output can be found in this document.


## Results

### Load R packages and dataset as Seurat object

<!-- need to load the libraries every time otherwise an error occurs -->

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(patchwork)
```

```{r}
#| cache: false
#| cache-lazy: false

con <- gzfile("/isilon/CBL_scRNAseq-archived/data/src/neurodev-genomics/scRNAseq/Luo_2022/Luo2022_humanFetalCerebellum_scRNAseq/GSM5952337_Fetal_cerebellum_final_cluster_1.rds.gz")
srat <- readRDS(gzcon(con))
close(con)
```

### First look at Seurat object

```{r}
print(srat)

head(srat[[]])
```

Looks like it contains 95542 cells and 32021 genes. This is consistent with the number of cells in the paper. The t-SNE reduction appears to be missing.

```{r}
table(srat$orig.ident, useNA = "ifany")
table(srat$batch, useNA = "ifany")
table(srat$cluster.names, useNA = "ifany")
table(srat$celltype, useNA = "ifany")
```

### QC plots

Plot cells grouped by batch (looks like it's just the age of the sample).

```{r}
#| fig-width: 6
#| fig-height: 18

.plt <- VlnPlot(
  srat,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
  group.by = "batch",
  combine = FALSE
)

wrap_plots(.plt, nrow = 3) & NoLegend()
```

Plot cells grouped by sample ID (also matches with sample ages).

```{r}
#| fig-width: 6
#| fig-height: 18

.plt <- VlnPlot(
  srat,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
  group.by = "orig.ident",
  combine = FALSE
)

wrap_plots(.plt, nrow = 3) & NoLegend()
```

Plotting all cells in one group.

```{r}
#| cache: false
#| cache-lazy: false
#| fig-width: 6
#| fig-height: 18

# save old cell labels
old_idents <- Idents(srat)
# set common label
Idents(srat) <- "luo2022"

.plt <- VlnPlot(
  srat,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
  combine = FALSE
)

wrap_plots(.plt, nrow = 3) & NoLegend()

summary(srat$nCount_RNA)
summary(srat$nFeature_RNA)
summary(srat$percent.mt)

# reset cell labels
Idents(srat) <- old_idents
```

### UMAP

```{r}
#| fig-width: 8
#| fig-height: 8

DimPlot(srat, group.by = "celltype", label = TRUE, repel = TRUE) + NoLegend()
```

