---
title: Using Cytoscape to generate EnrichmentMaps
date: 2025-03-26
metadata-files:
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To generate Cytoscape EnrichmentMaps of the UBC pathway enrichment results.


## Files and directories

### Code

::: {.callout-important}
# CODE MUST BE RUN LOCALLY

All code MUST be run on a machine that has Cytoscape installed, e.g. on a laptop. It CANNOT be run on the pailab server or the cluster (or maybe it is possible, I just haven't figured out how to run it that way yet...)
:::

The code can be found in `HumanMouseUBC/integrated_human_ubc/01C_cytoscape.R`. This script generates EnrichmentMaps using the pathway enrichment results from `<pailab projects>/HumanMouseUBC/integrated_human_ubc/20250307`.

### Output

The output can be found in `<pailab projects>/HumanMouseUBC/integrated_human_ubc/20250326/`. It includes the following files:

- `*.cys`: Cytoscape session file
- `*.pdf`/`*.svg`: screenshots of the EnrichmentMaps for quick viewing


## Results

```{r}
#| echo: false
#| output: false
#| cache: false
#| cache-lazy: false

library(tidyverse)

# set results directory
results_dir <- "/.mounts/labs/pailab/private/projects/HumanMouseUBC/integrated_human_ubc/20250326/"
```

```{r}
#| echo: false
#| output: asis

ubc_clusts <- paste0("UBC_", c(0:8))

for (clust in ubc_clusts) {
  emap <- file.path(results_dir,  paste0(clust, ".svg"))
  if (file.exists(emap)) {
    cat(sprintf("\n![EnrichmentMap of enriched pathways in %s.](%s/%s.svg){#fig-%s}\n\n", clust, results_dir, clust, clust))
  } else {
    cat(sprintf("\nWARNING: no enriched pathways in %s\n\n", clust))
  }
}
```
