---
title: Abundance of UBCs across developmental ages
date: 2024-05-16
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine proportion of UBCs at each developmental age. Similar to 2024-04-25, but with updated Seurat integration from 2024-05-12.


## Files and directories

All code and output can be found in this document.

The input Seurat object is located in `/.mounts/labs/pailab/private/llau/results/integrated/20240516/fc_subset_analysis/20240514_cca_integ_subset.qs`. This contains the integration of the cells from the full cerebellum performed by Leo with the following datasets:

- Human - Aldinger et al. (2021), Luo et al. (2022), Sepp et al. (2024)
- Mouse - Vladoiu et al. (2019), Sepp et al. (2024)

The cells were then subsetted to keep only the RL lineage cells and to remove all human cells from before 11 PCW.


## Results

### Load packages and dataset (Seurat object)

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(Seurat)
```

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240516/fc_subset_analysis/20240514_cca_integ_subset.qs")
srat
```

Show quick UMAP with the clusters:

```{r}
#| label: fig-umap
#| fig-cap: "UMAP of RL lineage cells with cells coloured by cluster (0.3 resolution)."
#| fig-width: 8
#| fig-height: 6
#| cache: false
#| cache-lazy: false

DimPlot(
  srat,
  group.by = "snn_res.0.3",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
)
```

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- srat@meta.data

# fix factor levels for the 0.3 resolution clusters so it's in numeric order
md$snn_res.0.3 <- fct_relevel(
  md$snn_res.0.3,
  str_sort(levels(md$snn_res.0.3), numeric = TRUE)
)

head(md)
rm(srat)
```

Show ages across datasets.

```{r}
table(md$dataset_name[md$species == "human"], md$human_age[md$species == "human"])
table(md$dataset_name[md$species == "mouse"], md$mouse_age[md$species == "mouse"])
```

### Proportion of UBCs by age

The UBCs are in cluster 1 of @fig-umap.

Calculate proportion of mouse UBCs at each age.

```{r}
num_mouse_ct_unfiltered <- md %>% 
  filter(species == "mouse") %>% 
  group_by(snn_res.0.3, dataset_name, mouse_age) %>% 
  count()
num_mouse_ct_unfiltered

# show cell types that are being filtered out
filter(num_mouse_ct_unfiltered, n < 25) %>% print(n = Inf)

# filter and calculate proportion
prop_mouse_ct_filtered <- num_mouse_ct_unfiltered %>% 
  filter(n >= 25) %>% 
  ungroup(snn_res.0.3) %>% 
  mutate(proportion = n/sum(n)) %>% 
  ungroup()
prop_mouse_ct_filtered

# filter for UBCs
prop_mouse_ubc <- prop_mouse_ct_filtered %>%
  filter(snn_res.0.3 == 1)
prop_mouse_ubc
```

Do the same for the human UBCs.

```{r}
num_human_ct_unfiltered <- md %>% 
  filter(species == "human") %>% 
  group_by(snn_res.0.3, dataset_name, human_age) %>% 
  count()
num_human_ct_unfiltered

# show cell types that are being filtered out
filter(num_human_ct_unfiltered, n < 25) %>% print(n = Inf)

# filter and calculate proportion
prop_human_ct_filtered <- num_human_ct_unfiltered %>% 
  filter(n >= 25) %>% 
  ungroup(snn_res.0.3) %>% 
  mutate(proportion = n/sum(n)) %>% 
  ungroup()
prop_human_ct_filtered

# filter for UBCs
prop_human_ubc <- prop_human_ct_filtered %>%
  filter(snn_res.0.3 == 1)
prop_human_ubc
```

Make line charts showing change in UBC proportion by age.

```{r}
#| label: fig-ubc-prop
#| fig-cap: "Proportion of UBCs at each developmental age for (A) mouse datasets and (B) human datasets."
#| fig-width: 12
#| fig-height: 6

# mouse datasets
.plt1 <- ggplot(
  prop_mouse_ubc,
  aes(
    x = mouse_age,
    y = proportion,
    colour = dataset_name,
    group = dataset_name
  )
) + 
  geom_point() + 
  geom_line() + 
  labs(x = "mouse age", y = "proportion of UBCs") + 
  scale_colour_hue(
    labels = \(.x) {str_remove(.x, "_\\w*")}
  )

# human datasets
.plt2 <- ggplot(
  prop_human_ubc,
  aes(
    x = human_age,
    y = proportion,
    colour = dataset_name,
    group = dataset_name
  )
) + 
  geom_point() + 
  geom_line() + 
  labs(x = "human age", y = "proportion of UBCs") + 
  scale_colour_hue(
    labels = \(.x) {str_remove(.x, "_\\w*")}
  )

.plt <- wrap_plots(
  .plt1,
  .plt2,
  nrow = 1
) + 
  plot_annotation(tag_levels = "A") & 
  guides(colour = guide_legend(nrow = 1)) & 
  theme_minimal() & 
  theme(
    legend.position = "bottom",
    legend.justification = c("center", "top"),
    axis.text = element_text(colour = "black"),
  )

.plt
```
