---
title: Enriched pathways for upregulated genes in the UBC subclusters
date: 2024-10-29
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To perform gene set enrichment analysis on the upregulated genes in each of the UBC subclusters. Similar to Leo's analysis on 2024-07-15 except that:

1. the nominal p-value was used to filter for upregulated genes (instead of the Bonferroni-corrected p-values)
    - i.e. genes with Aldinger p-value < 0.05 & Sepp p-value < 0.05 & log2FC > 0 were considered upregulated in a cluster
1. the Bader lab gene sets were filtered for those with 10-250 genes **prior to running `gost`** (see `_paths_to_datasets.qmd`)


## Files and directories

### Code

The code can be found in this document. It was adapted from `<Leo's utilities software directory>/ubcs_gost_enrichment.R` and run using the `scrnaseq_env` conda environment.

Differential gene expression was run by Leo and the results can be found in `/.mounts/labs/pailab/private/llau/results/integrated/20240715/`. Filtering of the genes was re-run using the nominal p-value.

### Output

The output directory can be found in `<integrated results directory>/20241029`. Within the parent directory, there is:

- `cluster_marker_ranking.rds`, a list of data frames that rank the upregulated genes in both datasets based on their nominal p-value

There is also a directory for each of the UBC clusters, and each directory contains:

- updated differential expression results
  - `background_genes.txt`, list of genes that were tested for differential expression (should be unchanged from Leo's analysis on 2024-07-15)
  - `upregulated_genes.txt`, list of genes that are upregulated in that cluster (nominal p-value < 0.05 and log2FC > 0 for both datasets)
- updated pathway enrichment results
  - `enr_pathways.rds`, raw output from `gprofiler2::gost()`
  - `enr_pathways.csv`, results table extracted from the raw output RDS file
  - `enr_pathways.html` and associated `enr_pathways_files` directory, Manhattan plot showing the enriched pathways
  - `enr_pathways.gem.txt`, a Generic Enrichment Map file used as input to Cytoscape for visualization


## Results

### Redo filtering with nominal p-values

Load packages and set variables:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)

out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241029/"
ubc_clust <- paste0("UBC_", 0:5)

# Leo's differential gene expression results directory
diff_expr_dir <- "/.mounts/labs/pailab/private/llau/results/integrated/20240715/"
```

Load differential gene expression results and re-filter genes (code not run when rendering the notebook, needs to be run separately):

```{r}
#| eval: false
#| code-fold: true

# load full results
diff_expr <- read_csv(file.path(diff_expr_dir, "all_tested_genes.csv")) %>%
  rename_with(.fn = str_remove, pattern = "full_cerebellum_")

# filter for upregulated genes in each cluster and save results
walk(
  .x = ubc_clust,
  .f = \(clust) {
    # make directories
    clust_dir <- file.path(out_dir, clust)
    if (!dir.exists(clust_dir)) {
      dir.create(clust_dir, recursive = TRUE)
    }

    # get differential expression results for the specific cluster
    clust_de <- filter(diff_expr, cluster == clust)

    # save all genes as background (should be identical to background file in
    # Leo's 2024-07-15 results directory)
    # write_lines(
    #   x = clust_de$feature,
    #   file = file.path(clust_dir, "background_genes.txt")
    # )

    # get upregulated genes
    upreg_genes <- filter(
      clust_de,
      Aldinger_human_p_val < 0.05 & Sepp_human_p_val < 0.05 & 
        Aldinger_human_avg_log2FC > 0 & Sepp_human_avg_log2FC > 0
    )
    browser()
    # save upregulated genes
    write_lines(
      x = upreg_genes$feature,
      file = file.path(clust_dir, "upregulated_genes.txt")
    )
  }
)
```

Rank the top markers for each UBC subcluster (same as 2024-08-27):

1. Rank the genes based on each dataset's adj. p-value
1. Combine the rankings from both datasets to determine the top markers

```{r}
# load differential gene expression results
diff_expr <- read_csv(file.path(diff_expr_dir, "all_tested_genes.csv")) %>%
  rename_with(.fn = str_remove, pattern = "full_cerebellum_")

# combine and rank the UBC markers from both Aldinger/Sepp datasets (sum the
# rank of each feature in each dataset, ranked by adjusted p-value)
cluster_markers_rank <- map(
  .x = ubc_clust,
  .f = \(clust) {
    # load list of upregulated genes
    upreg_genes <- read_lines(file.path(
      "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241029",
      clust,
      "upregulated_genes.txt"
    ))

    # rank the genes by (nominal) p-value
    feature_ranking <- diff_expr %>%
      filter(cluster == clust & feature %in% upreg_genes) %>%
      mutate(
        aldinger_rank = min_rank(Aldinger_human_p_val),
        sepp_rank = min_rank(Sepp_human_p_val),
        combined_rank = aldinger_rank + sepp_rank
      ) %>%
      arrange(combined_rank) %>%
      select(feature, aldinger_rank, sepp_rank, combined_rank)

    return(feature_ranking)
  }
) %>%
  setNames(nm = ubc_clust)

# save the UBC marker rankings
saveRDS(
  cluster_markers_rank,
  file = file.path(out_dir, "cluster_marker_ranking.rds")
)

head(cluster_markers_rank, n = 3)
```

### Run pathway enrichment analysis

::: {.callout-important}
The code chunk below needs to be run separately; it is **NOT** run when rendering the notebook. Internet access is required for g:Profiler.
:::

```{r}
#| eval: false
#| code-fold: true

library(tidyverse)
library(gprofiler2)

# load helper functions
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/gprofiler2_helpers.R")

out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20241029/"
ubc_clust <- paste0("UBC_", 0:5)

# loop through each subcluster UBC_0 to UBC_5
walk(
  .x = ubc_clust,
  .f = \(clust) {
    clust_dir <- file.path(out_dir, clust)

    # load upregulated genes
    upreg_genes <- read_lines(file.path(clust_dir, "upregulated_genes.txt"))

    message(sprintf(
      "***%s contains %s upregulated genes***",
      clust,
      length(upreg_genes)
    ))

    # load background genes
    bg <- read_lines(file.path(clust_dir, "background_genes.txt"))

    # run pathway enrichment
    gost_res <- run_gost(
      query = upreg_genes,
      organism = "gp__OUFl_gpIE_RyE", # 10-250 genes in each pathway
      significant = FALSE,
      custom_bg = bg,
      filename = file.path(clust_dir, "enr_pathways")
    )

    # save as Generic Enrichment Map file for downstream Cytoscape visualization
    gost_res2gem(gost_res = gost_res, phenotype = "+1") %>%
      write_gem(file = file.path(clust_dir, "enr_pathways.gem.txt"))
  }
)

message("\n***SESSION INFO***\n")
print(sessionInfo())
```

For all code output, see the `run_gost.log` file in the output directory.

### Show enriched pathways

Print stats for each of the pathways and the top pathways in each UBC cluster. Note that the `p_value` column shows FDR-corrected p-values (g:Profiler only returns the corrected p-values, not the raw p-values).

```{r}
walk(
  .x = ubc_clust,
  .f = \(clust) {
    message(sprintf("RESULTS FOR %s:", clust))
    clust_dir <- file.path(out_dir, clust)

    # load upregulated genes
    upreg_genes <- read_lines(file.path(clust_dir, "upregulated_genes.txt"))

    # print number of upregulated genes
    message(sprintf("%s upregulated genes", length(upreg_genes)))

    # load pathway results
    gost_res <- readRDS(file.path(clust_dir, "enr_pathways.rds"))

    # number of significant pathways
    message("number of significant pathways:")
    print(table(gost_res$result$significant))

    # print top 20 pathways
    gost_res$result %>%
      arrange(p_value) %>%
      slice_head(n = 20) %>%
      select(term_name, p_value, term_size, query_size, intersection_size) %>%
      print()
  }
)
```
