---
title: Running `scProportionTest` using annotated RL cell types
date: 2024-08-22
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine if there is a statistically significant difference in cell type proportions between human and mouse using `scProportionTest`. Similar to Leo's analysis on 2024-06-13 and 2024-07-18 except that the cell type annotations (instead of cluster numbers) are being used.


## Files and directories

### Code

All code can be found in this document. The code was run using the `scrnaseq_env` environment (note that `scProportionTest` needs to be installed through GitHub after setting up the conda environment - see conda README).

### Output

The output directory can be found at `<integrated results directory>/20240822`. There are two output files:

- `rl_broad_annot_scpt.qs` contains the `scProportionTest` object
- `rl_broad_annot_results.csv` contains final table with the permutation test results (contained in the `@results$permutation` slot of the `scProportionTest` object)


## Results

### Load packages

Load packages and other functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(scProportionTest)

# helper functions
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/cell_labelling.R")
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240822/"
```

### Load Seurat object

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240524/25_pc_without_luo/25_pc_rl.qs")
```

Add broad cell type annotations to the metadata:

```{r}
#| cache: false
#| cache-lazy: false

# get cell type annotations (from `cell_labelling.R`)
srat <- label_rl_lineage_integration(srat)

# collapse UBC subclusters into one cluster, set species factor levels
srat@meta.data <- mutate(
  srat[[]],
  broad_annot = case_when(
    str_starts(cell_type_annot, "UBC") ~ "UBC",
    .default = cell_type_annot
  ),
  species = factor(species, levels = c("mouse", "human"))
)
```

Filter out the control cells (endothelial cells, microglia, and oligodendrocytes):

```{r}
#| cache: false
#| cache-lazy: false

srat <- subset(
  x = srat,
  subset = broad_annot %in% c("endothelial", "microglia", "oligodendrocyte/OPC"),
  invert = TRUE
)
```

### Run `scProportionTest`

First, convert the Seurat object to the `scProportionTest` object:

```{r}
rl_scpt <- sc_utils(srat)
```

Then, run the proportion test (basically a permutation test):

```{r}
# set the seed for reproducibility
set.seed(1)

# run proportion test
rl_scpt <- permutation_test(
	rl_scpt,
  cluster_identity = "broad_annot",
	sample_1 = "mouse",
  sample_2 = "human",
	sample_identity = "species",
  n_permutations = 10000
)

# save proportion test results
write_csv(
  x = rl_scpt@results$permutation,
  file = file.path(out_dir, "rl_broad_annot_results.csv")
)

# save scProportionTest object
qs::qsave(
  x = rl_scpt,
  file = file.path(out_dir, "rl_broad_annot_scpt.qs")
)
```

Plot the permutation test results (default plotting function from `scProportionTest`):

```{r}
#| label: fig-scpt-plot
#| fig-cap: "Observed log2(Fold Difference) in cell type proportions between human and mouse. UBCs are significantly enriched in humans while GCPs are significantly enriched in mice (FDR < 0.05 and |log2FD| > 0.58, permutation test, n = 10000 permutations.)"
#| fig-width: 5
#| fig-height: 3
#| code-fold: true

permutation_plot(rl_scpt) + 
  guides(colour = guide_legend(title = NULL, ncol = 1)) + 
  theme(legend.position = "bottom", legend.box.spacing = unit(2.5, "points"))
```

A log2FD > 0 indicates that the cell type is enriched in humans compared to mice, and vice versa for a log2FD < 0.
