---
title: Preparing Luo et al. 2022 dataset for integration
date: 2024-03-15
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To prepare the Luo et al. (2022) dataset for integration.


## Files and directories

### Code

The code can be found in `<Luo human software directory>/prep_for_integration.R`.

Briefly, this script runs `SCTransform` on the dataset (regressing out `CC.Difference`), performs dimensionality reduction (PCA and UMAP), and re-clusters the cells. The original reductions were saved to the "pca_orig" and "umap_orig" slots.

### Output

The output can be found in `<Luo human results directory>/20240315`. The Seurat object is saved as output, and figures of the SCTransformed data can be found below.


## Results

Load Luo dataset and required packages:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(Seurat)

srat <- readRDS("/CBL_scRNAseq/results/human/Luo/20240315/luo2022_sct.rds")

source("/CBL_scRNAseq/software/utilities/plotting.R")
```

### PCA

```{r}
ElbowPlot(srat, ndims = 100)

DimPlot(srat, reduction = "pca", label = TRUE, repel = TRUE) + NoLegend()
```

The UMAP was generated using the top 25 PCs.

### UMAP

```{r}
#| label: fig-umap-sct
#| fig-width: 8
#| fig-height: 18
#| fig-cap: "UMAP after SCTransform with cells coloured by (a) clusters, (b) cell type, and (c) sample age."
#| cache: false
#| cache-lazy: false

.plt_list <- map(
  .x = c("seurat_clusters", "orig_cell_type"),
  .f = \(x) {
    .plt <- DimPlot(
      srat,
      reduction = "umap",
      group.by = x,
      label = TRUE,
      repel = TRUE
    ) + 
      NoLegend()
    
    return(.plt)
  }
)

.plt_list <- append(
  .plt_list,
  list(DimPlot(srat, reduction = "umap", group.by = "batch"))
)

wrap_plots(
  .plt_list,
  nrow = 3
) + 
  plot_annotation(tag_levels = "A")
```

```{r}
#| label: fig-umap-orig
#| fig-width: 6
#| fig-height: 6
#| fig-cap: "Original UMAP from published dataset with cells coloured by cell type."

DimPlot(
  srat,
  reduction = "umap_orig",
  group.by = "orig_cell_type",
  label = TRUE,
  repel = TRUE
) + 
  NoLegend()
```

Overall, the UMAP after SCTransform (@fig-umap-sct) looks very similar to the original published UMAP (@fig-umap-orig). The cell types are clustering together as expected.

### Clustering reflects original cell types

```{r}
#| label: fig-barplot-cell-types
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Bar plot showing the original cell type annotations in each SCTransformed cluster."

cluster_barplot(
  srat,
  split.by = "orig_cell_type",
  group.by = "seurat_clusters"
) + 
  scale_fill_manual(
    values = pals::cols25()
  )
```

The new clusters appear to reflect the original cell type annotations. In general, the majority of cells in each cluster come from 1-3 different cell types (@fig-barplot-cell-types).

- Clusters 0/5 contain a mix of GCP_1 and GCP_2
- Clusters 3/4 contain a mix of GCP_1, granule cells, and differentiated UBCs; cluster 3 also contains TCP cells
- Cluster 8 contains a mix of granule cells and differentiated UBCs
- Cluster 13 contains a mix of GCP progenitors and UBC progenitors


## Next steps

The dataset should be ready for integration with other human/mouse datasets.
