---
title: Pseudobulk differential expression between human and mouse RL and UBC
date: 2023-08-18
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To find human-specific genes in the developing cerebellum by running differential expression of human vs. mouse pseudobulked scRNA-seq data. Inspired by a paper from the Konopka lab (Caglayan et al. 2023, Nature).


## Files and directories

### Code

The code can be found in `<species_dge software directory>/pseudobulk_diff_expr.R>`. The single-cell datasets used were:

- `glutamatergic_dev_Liam.RDS` (human Aldinger dataset from Liam)
- `merged_seurat_RLonly.rds` (mouse Vladoiu dataset processed by me)

Briefly, this script

1. merges the human and mouse single-cell RNA counts and normalizes it, keeping only the orthologous genes
1. runs pseudobulk using `scuttle::aggregateAcrossCells`
1. runs differential gene expression in human vs. mouse using `edgeR::exactTest`
    a. BCV is set to 0.5 (recommended is 0.4 if comparing across human samples)

### Output

The output can be found in `<species_dge results directory>/20230818`. The list of differentially expressed genes can be found saved as a CSV file and as a DGEList object.


## Results

Three separate tests were run comparing human vs. mouse:

1. human RL-VZ vs. mouse RL (`01_rl_vz`)
1. human RL-SVZ vs. mouse RL (`02_rl_svz`)
1. human early + late UBC vs. mouse UBC (`03_ubc`)

Load packages to show results.

```{r}
library(tidyverse)
library(edgeR)
```

### `01_rl_vz`

Load data

```{r}
rl_vz <- readRDS("/CBL_scRNAseq/results/species_dge/20230818/01_rl_vz.rds")
```

```{r}
# Number of genes tested
nrow(rl_vz$table)
```

```{r}
# Number of genes with FDR < 0.05
nrow(filter(rl_vz$table, FDR < 0.05))
```

Print top 20 differentially expressed genes.

```{r}
head(rl_vz$table, 20)
```

### `02_rl_svz`

Load data

```{r}
rl_svz <- readRDS("/CBL_scRNAseq/results/species_dge/20230818/02_rl_svz.rds")
```

```{r}
# Number of genes tested
nrow(rl_svz$table)
```

```{r}
# Number of genes with FDR < 0.05
nrow(filter(rl_svz$table, FDR < 0.05))
```

Print top 20 differentially expressed genes.

```{r}
head(rl_svz$table, 20)
```

### `03_ubc`

Load data

```{r}
ubc <- readRDS("/CBL_scRNAseq/results/species_dge/20230818/03_ubc.rds")
```

```{r}
# Number of genes tested
nrow(ubc$table)
```

```{r}
# Number of genes with FDR < 0.05
nrow(filter(ubc$table, FDR < 0.05))
```

Print top 20 differentially expressed genes.

```{r}
head(ubc$table, 20)
```
