---
title: Running pathway enrichment analysis on UBC subclusters
date: 2025-03-07
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To run a pathway enrichment analysis on the UBC clusters using the upregulated genes from the previous differential gene expression analysis.


## Files and directories

### Code

The code can be found in `HumanMouseUBC/integrated_human_ubc/01B_pathway_enrichment.R`. All code was run using the `scrnaseq_env` conda environment.

`g:Profiler` was run using a custom EnrichmentMap GMT gene set file from the Bader Lab, using only the gene sets with 10-250 genes. See notes [here](_paths_to_datasets.html) for more information.

### Output

The output can be found in `<pailab projects>/HumanMouseUBC/integrated_human_ubc/20250307/`. The results for each UBC cluster can be found in its own subdirectory, with each subdirectory containing the following files:

- gene lists for g:Profiler
  - `bg_genes.txt`: list of genes that were tested for differential expression
  - `upreg_genes.txt`: list of genes that are upregulated in that cluster (nominal p-value < 0.05 and log2FC > 0 for both datasets)
- pathway enrichment results
  - `enriched_pathways.rds`: raw output from `gprofiler2::gost()`
  - `enriched_pathways.csv`: results table extracted from the raw output RDS file
  - `enriched_pathways.html` and associated `enr_pathways_files` directory: Manhattan plot showing the enriched pathways
  - `enriched_pathways.gem.txt`: a Generic Enrichment Map file that can be used as input to Cytoscape for visualization


## Results

```{r}
#| echo: false
#| output: false
#| cache: false
#| cache-lazy: false

library(tidyverse)

# set results directory
results_dir <- "/.mounts/labs/pailab/private/projects/HumanMouseUBC/integrated_human_ubc/20250307/"
```

```{r}
#| echo: false
#| output: asis

ubc_clusts <- paste0("UBC_", c(0:8))

for (clust in ubc_clusts) {
  stats <- list()

  # stats for enriched pathways
  if (file.exists(file.path(results_dir, clust, "enriched_pathways.csv"))) {
    pathways <- read_csv(
      file.path(results_dir, clust, "enriched_pathways.csv"),
      show_col_types = FALSE
    ) %>%
      pull("term_name")
    stats$top_pathways <- head(pathways, n = 10)
    stats$num_pathways <- length(pathways)
  } else {
    pathways <- NULL
  }

  # stats for upregulated and background genes
  stats$num_upreg <- read_lines(file.path(results_dir, clust, "upreg_genes.txt")) %>%
    length()
  stats$num_bg <- read_lines(file.path(results_dir, clust, "bg_genes.txt")) %>%
    length()

  # print heading
  cat(sprintf("\n\n### %s\n\n", clust))

  # show stats
  cat(str_glue(
    "- Number of upregulated genes: {stats$num_upreg}",
    "- Number of genes in background: {stats$num_bg}",
    .sep = "\n"
  ), "\n")
  if (!is.null(pathways)) {
    cat(str_glue(
      "- Number of enriched pathways (FDR < 0.05): {stats$num_pathways}",
      "- Top pathways:\n{paste0('\t1. ', stats$top_pathways, collapse = '\n')}",
      .sep = "\n"
    ), "\n")
  } else {
    cat("- No enriched pathways were found\n")
  }
}
```
