---
title: Testing for differences in proportions of UBC subtypes between species
date: 2024-06-06
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine if there is a statistically significant difference in the proportions of UBC subclusters between human and mouse.


## Files and directories

### Code

All code can be found in this document. The code was run using the `propeller` conda environment.

We perform the proportion test on three integrated objects:

- `/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/vladoiu_liam_RL.rds`: the original two-dataset integration (Aldinger/human + Vladoiu/mouse)
- `/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs`: the UBC cells subsetted from the RL lineage cells of the four-dataset full cerebellum **CCA** integration (Aldinger/human, Sepp/human, Sepp/mouse, Vladoiu/mouse)
- `/.mounts/labs/pailab/private/llau/results/integrated/20240606/rpca_20_analysis/ubc/ubc.qs`: the UBC cells subetted from the RL lineage cells of the four-dataset full cerebellum **RPCA** integration (Aldinger/human, Sepp/human, Sepp/mouse, Vladoiu/mouse)

`propeller` was run using the default logit transformation; according to the author, this better controls the type I error rate with small sample sizes ([Phipson et al. 2022](https://academic.oup.com/bioinformatics/article/38/20/4720/6675456)).

### Output

The output directory can be found at `<integrated results directory>/20240606`. This contains the tables of results that were generated by running `propeller` in this document. There are two types of files:

- `*_proportions.csv` contains the proportion of cells in each cluster for each sample (i.e., the proportions for each sample add up to 1); this is (part of) the output of `getTransformedProps`
- `*_results.csv` contains the output of the `propeller` test for each cluster, including the p-values and the FDRs


## Results

### Load packages

Load packages and other functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(speckle)
library(patchwork)
# library(ComplexHeatmap)

source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/cluster_barplot.R")
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240606/"
```

Some variables and helper functions (click below to expand):

```{r}
#| code-fold: true

# set colour for the species
species_cols <- c("grey", scales::pal_hue()(1))

# make UMAPs and bar plot showing the species and clusters of a Seurat object
make_summary_fig <- function(
  srat,
  cluster
) {
  # UMAP showing clusters
  .plt1 <- DimPlot(
    srat,
    group.by = cluster,
    label = TRUE,
    repel = TRUE,
    raster = FALSE
  )

  # UMAP showing human/mouse cells
  human_cells <- rownames(srat[[]])[srat$species == "human"]
  .plt2 <- DimPlot(
    srat,
    group.by = "species",
    cells.highlight = human_cells,
    sizes.highlight = 0.01,
    label = FALSE,
    raster = FALSE
  ) + 
    labs(title = NULL) + 
    scale_colour_manual(labels = c("mouse", "human"), values = species_cols)

  # cluster_barplot.R
  srat$species <- fct_relevel(srat$species, "mouse", "human")
  .plt3 <- cluster_barplot(
    srat,
    split.by = "species",
    group.by = cluster,
    position = "stack"
  ) + 
    scale_fill_manual(values = species_cols)

  .plt <- (.plt1 + .plt2) / .plt3 + 
    plot_annotation(tag_levels = "A") & 
    theme(plot.tag = element_text(face = "bold"))

  return(.plt)
}

# get proportions for plotting
get_props_per_sample <- function(
  clusters,
  sample,
  dataset_sample_matching_df
) {
  props <- getTransformedProps(
    clusters = clusters,
    sample = sample,
    transform = "logit"
  ) %>%
    pluck("Proportions") %>% # get this list item
    as.data.frame() %>%
    rename(proportion = Freq) %>%
    merge(x = ., y = dataset_sample_matching_df, by = "sample", all.x = TRUE)

  return(props)
}

# make bar plot showing the proportion of cells in each cluster for each sample
plot_props_per_sample <- function(
  props
) {
  # make the plot
  .plt <- ggplot(
    props,
    aes(x = sample, y = proportion, fill = clusters)
  ) + 
    geom_col(position = "stack") + 
    facet_grid(cols = vars(dataset_name), scales = "free", space = "free") + 
    scale_y_continuous(expand = expansion(0, 0.05)) + 
    scale_fill_manual(
      values = pals::cols25(n = length(unique(props[["clusters"]])))
    ) + 
    theme_minimal() + 
    theme(
      axis.text.x = element_text(hjust = 1, angle = 60)
    )

  return(.plt)
}
```

### UBCs in the original two-dataset integration

Here, we run propeller to test for differences in UBC proportions in my original Aldinger + Vladoiu CCA integration.

Load Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

two_dataset_srat <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/vladoiu_liam_RL.rds")
two_dataset_srat
```

Show a quick UMAP and bar plot of all the RL cells.

```{r}
#| label: fig-two-dataset-summary
#| fig-cap: "Overview of the RL lineage cells from the integration of the Aldinger and Vladoiu datasets. (A-B) UMAP with cells coloured by (A) cluster or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

make_summary_fig(two_dataset_srat, "seurat_clusters")
```

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- two_dataset_srat@meta.data
head(md)

# delete Seurat object
rm(two_dataset_srat)
```

#### Run `propeller` on all cells of the RL lineage

What's the total number of human vs. mouse cells?

```{r}
table(md$species)
```

To run `propeller`, we need the cluster, sample, and group of each cell. The cluster information is in the `seurat_clusters` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset, "Aldinger")],
  md$orig.ident[str_which(md$dataset, "Aldinger")]
)
table(md$dataset, md$orig.ident)

# collect info for `propeller`
x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset, "Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = seurat_clusters,
  dataset_name = case_when(
    str_detect(dataset, "Aldinger") ~ "Aldinger_human",
    str_detect(dataset, "Vladoiu") ~ "Vladoiu_mouse"
  )
) %>%
  select(cluster, sample, species, dataset_name)

samples_to_filter <- c(
  # filter out samples with fewer than 25 cells, similar to Sepp et al.
  names(table(x$sample)[table(x$sample) < 25]),
  # also filter out samples from 9 and 10 PCW for consistency with the new four-dataset integration
  {
    table(md$sample_id, md$age) %>%
      as.data.frame() %>%
      filter(Freq != 0) %>%
      rename(sample = Var1, age = Var2) %>%
      filter(age %in% c("9 PCW", "10 PCW")) %>%
      pull(sample) %>%
      as.character
  }
)
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
two_dataset_rl_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = two_dataset_rl_results,
  file = file.path(out_dir, "two_dataset_rl_results.csv")
)

two_dataset_rl_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-two-dataset-rl
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 8
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset and species
dataset_sample_matching <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(species = fct_relevel(species, "mouse", "human"))

# get proportion of cells in each cluster for each sample
two_dataset_rl_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  dataset_sample_matching_df = dataset_sample_matching
)
write_csv(
  two_dataset_rl_props,
  file = file.path(out_dir, "two_dataset_rl_proportions.csv")
)

dim(two_dataset_rl_props)
head(two_dataset_rl_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(two_dataset_rl_props) + 
  labs(y = "proportion of all RL lineage cells")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-two-dataset-rl
#| fig-cap: "Proportion of cells in each of the UBC clusters (7, 19, and 20) compared to all the cells in the RL. Each dot represents a sample from the dataset. Only the UBC clusters (7, 19, and 20) are shown in this graph."
#| fig-width: 8
#| fig-height: 6
#| code-fold: true

# only plot the UBCs (clusters 7, 19, and 20)
ubc_clusters <- c("7", "19", "20")

# FDRs to label on the plot
fdrs <- two_dataset_rl_results[, c("BaselineProp.clusters", "FDR")] %>%
  rename(clusters = BaselineProp.clusters) %>%
  filter(clusters %in% ubc_clusters) %>%
  arrange(clusters) %>%
  remove_rownames()

# plot proportions
.plt <- ggplot(
  two_dataset_rl_props[two_dataset_rl_props$clusters %in% ubc_clusters, ],
  aes(x = species, y = proportion)
) + 
  geom_boxplot(width = 0.6, outlier.shape = NA) + 
  geom_jitter(aes(shape = dataset_name), width = 0.2) + 
  geom_text(
    data = fdrs,
    mapping = aes(x = "mouse", y = 0.28, label = sprintf("adj. p = %.3f", FDR)),
    nudge_x = 0.5,
    inherit.aes = FALSE
  ) + 
  facet_wrap(vars(clusters)) + 
  labs(y = "proportion of all RL lineage cells") + 
  theme_light()

.plt
```

When comparing across all RL lineage cells, the UBC clusters (7, 19, and 20) all have a higher proportion of human cells than mouse cells, but only clusters 7 and 19 have an FDR < 0.05 (@fig-box-two-dataset-rl).

#### Run `propeller` on the UBCs only

Recall that the UBCs are in **clusters 7, 19, and 20**, with cluster 7 being "homologous/conserved", and clusters 19 and 20 being "non-homologous/human-specific". Here, we subset the UBCs and run `propeller` on just those clusters.

What's the number of human vs. mouse UBCs?

```{r}
md <- filter(md, seurat_clusters %in% c(7, 19, 20))
table(md$species)
```

As before, we need the cluster, sample, and group of each cell to run `propeller`.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset, "Aldinger")],
  md$orig.ident[str_which(md$dataset, "Aldinger")]
)
table(md$dataset, md$orig.ident)

# collect info for `propeller`
x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset, "Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = fct_drop(seurat_clusters),
  dataset_name = case_when(
    str_detect(dataset, "Aldinger") ~ "Aldinger_human",
    str_detect(dataset, "Vladoiu") ~ "Vladoiu_mouse"
  )
) %>%
  select(cluster, sample, species, dataset_name)

samples_to_filter <- c(
  # filter out samples with fewer than 10 cells, similar to Sepp et al.
  names(table(x$sample)[table(x$sample) < 10]),
  # also filter out samples from 9 and 10 PCW for consistency with the new four-dataset integration
  {
    table(md$sample_id, md$age) %>%
      as.data.frame() %>%
      filter(Freq != 0) %>%
      rename(sample = Var1, age = Var2) %>%
      filter(age %in% c("9 PCW", "10 PCW")) %>%
      pull(sample) %>%
      as.character
  }
)
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
two_dataset_ubc_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = two_dataset_ubc_results,
  file = file.path(out_dir, "two_dataset_ubc_results.csv")
)

two_dataset_ubc_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-two-dataset-ubc
#| fig-cap: "Bar plot showing UBC cluster proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 6
#| fig-height: 6
#| code-fold: true

# get proportion of cells in each cluster for each sample
two_dataset_ubc_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  dataset_sample_matching_df = dataset_sample_matching
)
write_csv(
  two_dataset_ubc_props,
  file = file.path(out_dir, "two_dataset_ubc_proportions.csv")
)

dim(two_dataset_ubc_props)
head(two_dataset_ubc_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(two_dataset_ubc_props) + 
  labs(y = "proportion of all UBCs")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-two-dataset-ubc
#| fig-cap: "Proportion of cells in each of the UBC clusters (7, 19, and 20) compared to all the UBCs (as opposed to @fig-box-two-dataset-rl which shows the proportions compared to all RL lineage cells). Each dot represents a sample from the dataset."
#| fig-width: 8
#| fig-height: 6
#| code-fold: true

# FDRs to label on the plot
fdrs <- two_dataset_ubc_results[, c("BaselineProp.clusters", "FDR")] %>%
  rename(clusters = BaselineProp.clusters) %>%
  arrange(clusters) %>%
  remove_rownames()

# plot proportions
.plt <- ggplot(two_dataset_ubc_props, aes(x = species, y = proportion)) + 
  geom_boxplot(width = 0.6, outlier.shape = NA) + 
  geom_jitter(aes(shape = dataset_name), width = 0.2) + 
  geom_text(
    data = fdrs,
    mapping = aes(x = "mouse", y = 1, label = sprintf("adj. p = %.3f", FDR)),
    nudge_x = 0.5,
    inherit.aes = FALSE
  ) + 
  facet_wrap(vars(clusters)) + 
  labs(y = "proportion of all RL lineage cells") + 
  theme_light()

.plt
```

As oppposed to comparing proportions in the entire RL lineage (@fig-box-two-dataset-rl), when comparing only the UBCs, none of the clusters have a significantly higher proportion of human or mouse cells (@fig-box-two-dataset-ubc). However, there does appear to be a lot of variability in the mouse cell proportions.

### UBCs in the updated four-dataset CCA integration

Load Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

four_dataset_srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs")
four_dataset_srat
```

Show quick UMAP and bar plot of the UBCs with the subclusters.

```{r}
#| label: fig-four-dataset-summary-cca
#| fig-cap: "Overview of the UBCs (CCA integration). (A-B) UMAP of UBCs with cells coloured by (A) cluster (0.3 resolution) or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

make_summary_fig(four_dataset_srat, "snn_res.0.3") + 
  labs(caption = "(CCA integration)")
```

What's the total number of human vs. mouse UBCs?

```{r}
table(four_dataset_srat$species)
```

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- four_dataset_srat@meta.data

# make sure factor levels for the 0.3 resolution clusters are in numeric order
md$snn_res.0.3 <- fct_relevel(
  md$snn_res.0.3,
  str_sort(levels(md$snn_res.0.3), numeric = TRUE)
)

head(md)

# delete Seurat object
rm(four_dataset_srat)
```

#### Run `propeller` on the subsetted UBC clusters

To run `propeller`, we need the cluster, sample, and group of each cell. The cluster information is in the `snn_res.0.3` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset_name, "^Aldinger")],
  md$orig.ident[str_which(md$dataset_name, "^Aldinger")]
)
table(md$dataset_name, md$orig.ident)

x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = snn_res.0.3
) %>%
  select(cluster, sample, species, dataset_name)

# filtering out samples with fewer than 10 cells, similar to Sepp et al.
samples_to_filter <- names(table(x$sample)[table(x$sample) < 10])
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
four_dataset_cca_ubc_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = four_dataset_cca_ubc_results,
  file = file.path(out_dir, "four_dataset_cca_ubc_results.csv")
)

four_dataset_cca_ubc_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-four-dataset-cca
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 12
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset
dataset_sample_matching <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get proportion of cells in each cluster for each sample
four_dataset_cca_ubc_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  dataset_sample_matching_df = dataset_sample_matching
)
write_csv(
  four_dataset_cca_ubc_props,
  file = file.path(out_dir, "four_dataset_cca_ubc_proportions.csv")
)

dim(four_dataset_cca_ubc_props)
head(four_dataset_cca_ubc_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(four_dataset_cca_ubc_props) + 
  labs(caption = "(CCA integration)")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-four-dataset-cca
#| fig-cap: "Proportion of cells in each of the UBC clusters compared to all the UBCs. Each dot represents a sample from the dataset."
#| fig-width: 8
#| fig-height: 10
#| code-fold: true

# FDRs to label on the plot
fdrs <- four_dataset_cca_ubc_results[, c("BaselineProp.clusters", "FDR")] %>%
  rename(clusters = BaselineProp.clusters) %>%
  arrange(clusters) %>%
  remove_rownames()

# plot proportions
.plt <- ggplot(four_dataset_cca_ubc_props, aes(x = species, y = proportion)) + 
  geom_boxplot(width = 0.6, outlier.shape = NA) + 
  geom_jitter(aes(shape = dataset_name), width = 0.2) + 
  geom_text(
    data = fdrs,
    mapping = aes(x = "mouse", y = 0.95, label = sprintf("adj. p = %.3f", FDR)),
    nudge_x = 0.5,
    inherit.aes = FALSE
  ) + 
  facet_wrap(vars(clusters), ncol = 3) + 
  labs(y = "proportion of all RL lineage cells", caption = "(CCA integration)") + 
  theme_light()

.plt
```

In @fig-box-four-dataset-cca, the only cluster that appears to have a significantly higher proportion of human cells is cluster 2. Meanwhile, clusters 3 and 5 have a significantly higher proportion of mouse cells.

### UBCs in the updated four-dataset RPCA integration

Load Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

four_dataset_srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240606/rpca_20_analysis/ubc/ubc.qs")
four_dataset_srat
```

Show quick UMAP and bar plot of the UBCs with the subclusters.

```{r}
#| label: fig-four-dataset-summary-rpca
#| fig-cap: "Overview of the UBCs (RPCA integration). (A-B) UMAP of UBCs with cells coloured by (A) cluster (0.6 resolution) or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

make_summary_fig(four_dataset_srat, "snn_res.0.6") + 
  labs(caption = "(RPCA integration)")
```

What's the total number of human vs. mouse UBCs?

```{r}
table(four_dataset_srat$species)
```

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- four_dataset_srat@meta.data

# make sure factor levels for the 0.6 resolution clusters are in numeric order
md$snn_res.0.6 <- fct_relevel(
  md$snn_res.0.6,
  str_sort(levels(md$snn_res.0.6), numeric = TRUE)
)

head(md)

# delete Seurat object
rm(four_dataset_srat)
```

#### Run `propeller` on the subsetted UBC clusters

To run `propeller`, we need the cluster, sample, and group of each cell. The cluster information is in the `snn_res.0.6` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset_name, "^Aldinger")],
  md$orig.ident[str_which(md$dataset_name, "^Aldinger")]
)
table(md$dataset_name, md$orig.ident)

x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = snn_res.0.6
) %>%
  select(cluster, sample, species, dataset_name)

# filtering out samples with fewer than 10 cells, similar to Sepp et al.
samples_to_filter <- names(table(x$sample)[table(x$sample) < 10])
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
four_dataset_rpca_ubc_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = four_dataset_rpca_ubc_results,
  file = file.path(out_dir, "four_dataset_rpca_ubc_results.csv")
)

four_dataset_rpca_ubc_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-four-dataset-rpca
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 12
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset
dataset_sample_matching <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get proportion of cells in each cluster for each sample
four_dataset_rpca_ubc_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  dataset_sample_matching_df = dataset_sample_matching
)
write_csv(
  four_dataset_rpca_ubc_props,
  file = file.path(out_dir, "four_dataset_rpca_ubc_proportions.csv")
)

dim(four_dataset_rpca_ubc_props)
head(four_dataset_rpca_ubc_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(four_dataset_rpca_ubc_props) + 
  labs(caption = "(RPCA integration)")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-four-dataset-rpca
#| fig-cap: "Proportion of cells in each of the UBC clusters compared to all the UBCs. Each dot represents a sample from the dataset."
#| fig-width: 10
#| fig-height: 10
#| code-fold: true

# FDRs to label on the plot
fdrs <- four_dataset_rpca_ubc_results[, c("BaselineProp.clusters", "FDR")] %>%
  rename(clusters = BaselineProp.clusters) %>%
  arrange(clusters) %>%
  remove_rownames()

# plot proportions
.plt <- ggplot(four_dataset_rpca_ubc_props, aes(x = species, y = proportion)) + 
  geom_boxplot(width = 0.6, outlier.shape = NA) + 
  geom_jitter(aes(shape = dataset_name), width = 0.2) + 
  geom_text(
    data = fdrs,
    mapping = aes(x = "mouse", y = 0.95, label = sprintf("adj. p = %.3f", FDR)),
    nudge_x = 0.5,
    inherit.aes = FALSE
  ) + 
  facet_wrap(vars(clusters), ncol = 4) + 
  labs(y = "proportion of all RL lineage cells", caption = "(RPCA integration)") + 
  theme_light() + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.875, 0.25)
  )

.plt
```
