---
title: Running PCA and UMAP on the Vladoiu dataset
date: 2024-08-10
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To run and visualize UMAP for the Vladoiu mouse dataset.


## Files and directories

### Code

All code can be found in this file.

### Output

The output directory can be found at `<Vladoiu mouse results directory>/20240810`.

- `seurat_with_umap.qs` contains an updated Seurat object with the UMAP reduction


## Results

Load packages and set output directory:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(Seurat)

out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/mouse/Vladoiu/20240810/"
if (!dir.exists(out_dir)) { 
  dir.create(out_dir, recursive = TRUE)
}
```

Load the Vladoiu dataset:

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/data/Vladoiu_2019/Vladoiu_full_cerebellum_mouse_20240403_stand.qs")
```

Show the PCs:

```{r}
#| label: fig-elbow
#| fig-cap: "Elbow plot showing the variance explained by the top 50 PCs."
#| fig-width: 8
#| fig-height: 6
#| cache: false
#| cache-lazy: false
#| code-fold: true

.plt <- ElbowPlot(srat, ndims = 50)
.plt
```

Run UMAP:

```{r}
#| cache: false
#| cache-lazy: false

srat <- RunUMAP(srat, dims = 1:25)
```

Plot UMAP:

```{r}
#| label: fig-umap
#| fig-cap: "UMAP visualization of the Vladoiu dataset with cells coloured by (A) original cell type and (B) re-annotated common cell type (for consistency with the other datasets). The top 25 PCs were used to generate the UMAP."
#| fig-width: 16
#| fig-height: 12
#| cache: false
#| cache-lazy: false
#| code-fold: true

srat$cell_type_tmp <- str_remove(string = srat$cell_type, pattern = "_Vladoiu")

.plt_list <- map(
  .x = c("cell_type_tmp", "common_cell_name"),
  .f = \(x) {
    if (x == "common_cell_name") {
      legend_ncol <- 3
      plt_title <- "common cell type"
    } else if (x == "cell_type_tmp") {
      legend_ncol <- 2
      plt_title <- "original cell type"
    }

    .plt <- DimPlot(
      srat,
      reduction = "umap",
      group.by = x,
      label = TRUE,
      repel = TRUE
    ) + 
      labs(title = plt_title) + 
      guides(
        colour = guide_legend(ncol = legend_ncol, override.aes = list(size = 2))
      ) + 
      theme(legend.position = "bottom")
    return(.plt)
  }
)

.plt <- wrap_plots(.plt_list, ncol = 2) + 
  plot_annotation(tag_levels = "A")

ggsave(
  filename = "umaps.png",
  plot = .plt,
  path = out_dir,
  width = 16,
  height = 12,
  units = "in",
  dpi = 600
)

.plt
```

Save new Seurat object with UMAPs:

```{r}
#| cache: false
#| cache-lazy: false

srat$cell_type_tmp <- NULL

qs::qsave(
  x = srat,
  file = file.path(out_dir, "seurat_with_umap.qs")
)
```

```{r}
sessionInfo()
```
