---
title: Age as confounder for the human-specific UBCs
date: 2023-05-18
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
execute: 
  echo: true
  eval: true
  output: true
---

## Goal of this analysis

Need to check if the age of the samples is a potential confounder for why the human-specific UBCs are clustering separately from the non-specific/mouse UBCs.


## Files and directories

### Code

Analysis was run in this notebook (all code is below).

### Output

All output is in this notebook.


## Results

First, import packages, custom functions, and Seurat object from file.

::: {.callout-note}
The following code cell options are required when loading in large files like a Seurat object:

```
#| cache: false
#| cache-lazy: false
```
:::

```{r}
#| output: false
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(patchwork)

source("/CBL_scRNAseq/software/mouse/Vladoiu/add_annotations.R")
source("/CBL_scRNAseq/software/utilities/cell_labelling.R")
source("/CBL_scRNAseq/software/utilities/cluster_barplot.R")

srat_integ <- readRDS("/CBL_scRNAseq/results/integrated/vladoiu_liam_RL.rds")
```

Add cell-level annotations to the cells (so we know which are homologous/non-homologous).

```{r}
#| cache: false
#| cache-lazy: false

# add Vladoiu mouse cell annotations
srat_integ <- label_vladoiu_cells(srat_integ)

# add annotations for the Seurat clusters after integration
srat_integ <- label_integ_clusters(srat_integ, integ_method = "cca")
```

We can then generate a figure showing the distribution of ages in each of the UBC clusters.

- Cluster 7 contains non-specific UBCs.
- Cluster 19 and 20 contain human-specific UBCs.

```{r}
#| label: fig-age-per-cluster
#| fig-cap: (a) Number and (b) proportion of cells in the UBC clusters split by age.
#| cache: false
#| cache-lazy: false

.plt <- purrr::map(
  .x = c("stack", "fill"),
  .f = \(position) {
    cluster_barplot(
      srat_integ,
      split.by = "age",
      group.by = "integ_clusters",
      position = position,
      width = 0.5,
      filter_data = "integ_clusters %in% c('7-Homol UBC', '19-NonHomol UBC', '20-NonHomol UBC')"
    )
  }
)

wrap_plots(.plt) + 
  plot_layout(guides = "collect") & 
  plot_annotation(tag_levels = "A") & 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

From @fig-age-per-cluster&#8288;B, it looks like clusters 7 and 19 have a similar distribution of ages, while cells in cluster 20 are mostly from 17 and 20 PCW. Because we are comparing non-specific vs. human-specific UBCs (i.e., 7 vs. {19, 20}), the age of the cells is unlikely to explain the differences between these two clusters.

However, a statistical test will need to be performed at a later date to confirm that the proportions are not significantly different across the clusters.
