---
title: Running `propeller` using annotated RL cell types
date: 2024-08-23
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine if there is a statistically significant difference in cell type proportions between human and mouse using `propeller`. Similar to my analysis on 2024-06-17 except that the cell type annotations (instead of cluster numbers) are being used.


## Files and directories

### Code

All code can be found in this document. The code was run using the `propeller` conda environment.

`propeller` was run using the default logit transformation; according to the author, this better controls the type I error rate with small sample sizes ([Phipson et al. 2022](https://academic.oup.com/bioinformatics/article/38/20/4720/6675456)).

### Output

The output directory can be found at `<integrated results directory>/20240823`. This contains the tables of results that were generated by running `propeller` in this document. There are two files:

- `rl_proportions.csv` contains the proportion of cells in each cluster for each sample (i.e., the proportions for each sample add up to 1); this is (part of) the output of `getTransformedProps`
- `rl_results.csv` contains the output of the `propeller` test for each cluster, including the p-values and the FDRs


## Results

### Load packages and functions

Load packages and other functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(speckle)
library(patchwork)

# helper functions
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/cell_labelling.R")
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/cluster_barplot.R")
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/propeller_helpers.R")
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240823/"
```

Some variables and helper functions (click below to expand):

```{r}
#| code-fold: true

# set colour for the species
species_cols <- c("grey", scales::pal_hue()(1))

# make UMAPs and bar plot showing the species and clusters of a Seurat object
make_summary_fig <- function(
  srat,
  cluster
) {
  # UMAP showing clusters
  .plt1 <- DimPlot(
    srat,
    group.by = cluster,
    label = TRUE,
    repel = TRUE,
    raster = FALSE
  )

  # UMAP showing human/mouse cells
  human_cells <- rownames(srat[[]])[srat$species == "human"]
  .plt2 <- DimPlot(
    srat,
    group.by = "species",
    cells.highlight = human_cells,
    sizes.highlight = 0.01,
    label = FALSE,
    raster = FALSE
  ) + 
    labs(title = NULL) + 
    scale_colour_manual(labels = c("mouse", "human"), values = species_cols)

  # cluster_barplot.R
  srat$species <- fct_relevel(srat$species, "mouse", "human")
  .plt3 <- cluster_barplot(
    srat,
    split.by = "species",
    group.by = cluster,
    position = "stack"
  ) + 
    scale_fill_manual(values = species_cols)

  .plt <- (.plt1 + .plt2) / .plt3 + 
    plot_annotation(tag_levels = "A") & 
    theme(plot.tag = element_text(face = "bold"))

  return(.plt)
}
```

### Load Seurat object

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240524/25_pc_without_luo/25_pc_rl.qs")
```

Add broad cell type annotations to the metadata:

```{r}
#| cache: false
#| cache-lazy: false

# get cell type annotations (from `cell_labelling.R`)
srat <- label_rl_lineage_integration(srat)

# collapse UBC subclusters into one cluster, set species factor levels
srat@meta.data <- mutate(
  srat[[]],
  broad_annot = case_when(
    str_starts(cell_type_annot, "UBC") ~ "UBC",
    .default = cell_type_annot
  ),
  species = factor(species, levels = c("mouse", "human"))
)

table(srat$broad_annot)
```

### Cell type proportions

Show quick UMAP and bar plot of the RL lineage cells (including controls cells - OPCs, microglia, endothelial).

```{r}
#| label: fig-summary
#| fig-cap: "Overview of the RL lineage cells. (A-B) UMAP of RL lineage cells with cells coloured by (A) general cell type or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

make_summary_fig(srat, "broad_annot")
```

What's the total number of human vs. mouse cells?

```{r}
table(srat$species)
```

Only the Seurat metadata is required - save it and delete the Seurat object. The following "control" cells are also filtered out, leaving only the RL lineage cells:

- endothelial cells
- microglia
- oligodendrocytes/OPCs

```{r}
md <- srat@meta.data %>%
  # filter out non-RL lineage cells (oligodendrocytes, endothelial, microglia)
  filter(!(broad_annot %in% c("endothelial", "microglia", "oligodendrocyte/OPC")))

head(md)

# delete Seurat object
rm(srat)
```

After filtering the control cells, `r table(md$species)["human"]` human cells and `r table(md$species)["mouse"]` mouse cells remain.

To run `propeller`, we need the cluster, sample, and group of each cell. The cluster information is in the `broad_annot` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset_name, "^Aldinger")],
  md$orig.ident[str_which(md$dataset_name, "^Aldinger")]
)
table(md$dataset_name, md$orig.ident)

# prep for propeller
x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = broad_annot
) %>%
  select(cluster, sample, species, dataset_name)

# filtering out samples with fewer than 50 cells, similar to Sepp et al.
samples_to_filter <- names(table(x$sample)[table(x$sample) < 50])
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
prop_res <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = prop_res,
  file = file.path(out_dir, "rl_results.csv")
)

prop_res
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 12
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset
sample_metadata <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get proportion of cells in each cluster for each sample
rl_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  sample_meta_df = sample_metadata
)
write_csv(
  rl_props,
  file = file.path(out_dir, "rl_proportions.csv")
)

dim(rl_props)
head(rl_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(rl_props, facet_cols = dataset_name)
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box
#| fig-cap: "Proportion of cells in each of the clusters out of all RL lineage cells. Each dot represents a sample from the dataset."
#| fig-width: 8
#| fig-height: 5
#| code-fold: true

# FDRs to label on the plot
fdrs <- get_fdrs_for_plotting(prop_res)

# plot proportions
boxplot_props_per_cluster(
  props = rl_props,
  facet_var = clusters,
  fdrs = fdrs
) + 
  theme(legend.position = "bottom")
```

