---
title: Enriched TF motifs and mutated genes in MB
date: 2023-02-16
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

Which of the TFs from yesterday's analysis are mutated in medulloblastoma?

References for gene mutations:

- Hendrikse et al., Nature 2022
- Cavalli et al., Cancer Cell 2017


## Files and directories

### Data

See `<human Aldinger results directory>/​20230215/​enriched_motifs_nonhomol.csv` for the TF motifs that were enriched in non-homologous UBC genes.

See `<isilon human Hendrikse data directory>/​supp_tab_5.xlsx` for the table of mutations in G3 and G4 MB.

### Output

All output can be found in this document.


## Results

### Import data and clean-up

Let's first import the mutations.

```{r}
#| output: false

library(tidyverse)

mut_xlsx <- "/isilon/CBL_scRNAseq-archived/data/human/Hendrikse/supp_tab_5.xlsx"
```

We need to take the second sheet of the file and start on the second row.

```{r}
mut_genes <- readxl::read_excel(
  path = mut_xlsx,
  sheet = 2,
  skip = 1
)

mut_genes
```

Note that some of the genes were stored as dates in Excel spreadsheet.

```{r}
mut_genes$Gene %>% sort() %>% head(10)
```

Columns need to be renamed and non-significant genes need to be filtered out. The first set of columns are for Group 4 MB and the second set of columns are for Group 3 MB. A column needs to be added to indicate tumour type in which the gene was found to be mutated.

```{r}
sig_mut_genes <- rename(
  mut_genes,
  gene = Gene,
  freq_g4 = Frequency...3,
  pval_oncodrive_g4 = pval_OncoDriveFML...4,
  qval_oncodrive_g4 = qval_OncoDriveFML...5,
  pval_mutsig_g4 = pval_MutSig...6,
  qval_mutsig_g4 = qval_MutSig...7,
  freq_g3 = Frequency...8,
  pval_oncodrive_g3 = pval_OncoDriveFML...9,
  qval_oncodrive_g3 = qval_OncoDriveFML...10,
  pval_mutsig_g3 = pval_MutSig...11,
  qval_mutsig_g3 = qval_MutSig...12,
) %>% 
  mutate(
    g4_mutated = case_when(
      qval_oncodrive_g4 < 0.05 | qval_mutsig_g4 < 0.05 ~ TRUE,
      TRUE ~ FALSE
    ),
    g3_mutated = case_when(
      qval_oncodrive_g3 < 0.05 | qval_mutsig_g3 < 0.05 ~ TRUE,
      TRUE ~ FALSE
    ),
    mutated_in = case_when(
      g4_mutated & g3_mutated ~ "both",
      g4_mutated ~ "G4",
      g3_mutated ~ "G3",
      TRUE ~ NA_character_
    )
  ) %>% 
  select(gene, Freq_whole_cohort, freq_g4, freq_g3, mutated_in) %>% 
  filter(!is.na(mutated_in))

sig_mut_genes
```

Import the TFs from yesterday.

```{r}
tf_genes <- readr::read_csv(
  "/CBL_scRNAseq/results/human/Aldinger/20230215/enriched_motifs_nonhomol.csv"
) %>% 
  mutate(
    gene = str_split(motif_id, "_", simplify = TRUE)[,1],
    .after = rank
  ) %>% 
  select(-c(motif_db, motif_id, motif_alt_id, consensus))

tf_genes
```

### Compare mutated genes and enriched TF binding sites

We can find the common genes of both `sig_mut_genes` and `tf_genes` by using the `intersect` function.

```{r}
intersect(sig_mut_genes$gene, tf_genes$gene)
```

ZIC1 seems to be well known in cerebellar development: [https://pubmed.ncbi.nlm.nih.gov/29442326/](https://pubmed.ncbi.nlm.nih.gov/29442326/).

However, looking more closely at IRF7...

```{r}
mut_genes %>% filter(Gene == "IRF7") %>% print(width = Inf)
```

The value in `qval_OncoDriveFML...10` doesn't make any sense since it's less than the unadjusted *p*-value of 0.0355.

Guess that just leaves ZIC1 then...
