---
title: Upregulated transcription factors in UBC subclusters and MB tumours
date: 2024-09-17
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To compare the top upregulated transcription factors (TFs) in UBC subclusters (particularly the human-enriched UBC_1 and UBC_2 cells) to those in MB tumours. Similar to the analysis on 2024-08-27 except that the focus here is on TFs that come up in differential gene expression.


## Files and directories

### Code

Code can be found in this document and in `<Vladoiu tumour software directory>/diff_exp_genes.R`. The `diff_exp_genes.R` script runs differential gene expression on the scRNA-seq dataset from Vladoiu (note the script takes several hours to run on the lab server). All code was run using the `scrnaseq_env` conda environment.

### Output

The output directories can be found at `<integrated results directory>/20240917` and `<Vladoiu tumour results directory>/20240917`.

- The `<integrated results directory>` contains:
  - volcano plots and dot plots of the differentially expressed TFs in each of the UBC subclusters
  - `num_shared_tfs.csv`, intersection of upregulated TFs between UBC subclusters and MB subtypes
- The `<Vladoiu tumour results directory>` contains:
  - differential gene expression results between different MB tumour subtypes (SHH, G3, G4) and different MB tumour clusters (from Louvain clustering)
  - volcano plots showing differentially expressed TFs for the MB subtypes


## Results

Load packages and functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
# library(patchwork)
library(Seurat)

# helper functions
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/plotting.R")

# set output directories
integ_out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240917/"
tumour_out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/tumour/Vladoiu/20240917/"

```

### Upregulated TFs in UBC subclusters

First, import the differential gene expression results (some from Leo's folders):

- `all_tested_genes.csv` contains the full differential gene expression results from both datasets and all UBC subclusters
- `cluster_marker_ranking.rds` contains the genes from differential gene expression where the Aldinger adj. p-value < 0.05 AND the Sepp adj. p-value < 0.05 AND the log2FC > 0 in both datasets

```{r}
# load differential gene expression results (all UBC clusters, both datasets)
ubc_diff_expr <- read_csv("/.mounts/labs/pailab/private/llau/results/integrated/20240715/all_tested_genes.csv") %>%
  rename_with(.fn = str_remove, pattern = "full_cerebellum_")
print(ubc_diff_expr, width = Inf)

# get significantly upregulated UBC markers (all UBC clusters, both datasets)
cluster_marker_ranking <- readRDS("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240827/diff_expr_markers/cluster_marker_ranking.rds")
head(cluster_marker_ranking, n = 3)

# get the UBC clusters
ubc_subclusters <- names(cluster_marker_ranking)
```

Filter the genes for TFs:

```{r}
# load list of TFs
tfs <- read.csv(
  "/.mounts/labs/pailab/src/transcription-factors/human-tfs-lambert/full_database_v1.01.csv",
  row.names = 1,
  check.names = FALSE
)
table(tfs$`Is TF?`)

# get only confirmed TFs
tfs <- dplyr::filter(tfs, `Is TF?` == "Yes") %>%
  pull(`HGNC symbol`)

# get differentially expressed TFs
ubc_diff_expr <- dplyr::filter(
  ubc_diff_expr,
  feature %in% tfs
)

# get upregulated TFs
cluster_marker_ranking <- map(
  .x = cluster_marker_ranking,
  .f = \(x) {
    x <- dplyr::filter(x, feature %in% tfs)
    return(x)
  }
)

# print total number of upregulated TFs in each of the UBC subclusters
walk2(
  .x = cluster_marker_ranking,
  .y = ubc_subclusters,
  .f = \(upreg_tfs, subcluster) {
    message(sprintf(
      "%s: %s upregulated TFs",
      subcluster,
      nrow(upreg_tfs)
    ))
  }
)
```

Load the Seurat objects for the dot plots:

```{r}
#| cache: false
#| cache-lazy: false

ubc_srat <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240827/diff_expr_markers/human_ubc_srat_list.qs")
```

Using the upregulated TFs in each UBC subcluster, we make:

- A volcano plot for each dataset with those TFs labelled
- A dot plot for each dataset showing the scaled expression for each of those TFs

```{r}
#| cache: false
#| cache-lazy: false
#| code-fold: true

# make volcano plots and dot plots for the markers in each cluster
walk2(
  .x = ubc_srat,
  .y = names(ubc_srat),
  .f = \(srat, dataset) {
    # which column to use (based on dataset)
    logfc_column <- paste0(str_to_title(dataset), "_human_avg_log2FC")
    adj_pval_column <- paste0(str_to_title(dataset), "_human_p_val_adj")

    # additional columns for the volcano plot
    diff_expr <- ubc_diff_expr %>%
      mutate(direction = case_when(
        rlang::parse_expr(
          paste0(logfc_column, " < 0 & ", adj_pval_column, " < 0.05 ~ 'down'")
        ),
        rlang::parse_expr(
          paste0(logfc_column, " > 0 & ", adj_pval_column, " < 0.05 ~ 'up'")
        ),
        .default = "n.s."
      ))

    # loop through each UBC cluster
    for (subcluster in ubc_subclusters) {

      # get upregulated TFs
      features <- cluster_marker_ranking[[subcluster]]$feature

      # make and save the volcano plot
      .plt <- diff_expr %>%
        filter(cluster == subcluster) %>%
        # label upregulated TFs
        mutate(gene_label = case_when(
          feature %in% features ~ feature,
          .default = ""
        )) %>%
        make_volcano(
          log_fc = .data[[logfc_column]],
          log_pval = -log10(.data[[adj_pval_column]]),
          direction = direction,
          gene_labels = gene_label,
          log_fc_thresh = 0,
          log_pval_thresh = -log10(0.05)
        ) + 
        labs(y = "-log10(adj. p)") + 
        scale_colour_manual(
          values = c("down" = "blue", "n.s." = "black", "up" = "red")
        ) + 
        theme_classic()
      # set text label size
      .plt$layers[[2]]$aes_params$size <- 2.5
      ggsave(
        filename = paste0(dataset, "_volcano.png"),
        plot = .plt,
        path = file.path(integ_out_dir, paste0(subcluster, "_TFs")),
        width = 6,
        height = 5,
        units = "in",
        dpi = 600
      )

      # skip dot plot if there are no upregulated TFs
      if (length(features) == 0) next

      # make and save the dot plot
      .plt <- DotPlot(
        srat,
        features = rev(features),
        assay = "SCT",
        group.by = "subclusters",
        scale = TRUE
      ) + 
        coord_flip() + 
        theme(
          axis.text.x = element_text(hjust = 1, angle = 45),
          plot.background = element_rect(fill = "#FFFFFF")
        )
      # variable plot height depending on number of markers in the cluster
      height <- max(length(features) * 0.5 + 1, 4)
      ggsave(
        filename = paste0(dataset, "_dotplot.png"),
        plot = .plt,
        path = file.path(integ_out_dir, paste0(subcluster, "_TFs")),
        width = 6,
        height = height,
        units = "in",
        dpi = 600
      )
    }
  }
)
```

### Upregulated TFs in MB tumours

Load the MB differential gene expression results:

```{r}
# load differential expression results
mb_diff_exp <- read_csv(file.path(tumour_out_dir, "mb_subtype_diff_exp.csv"))

# set factor levels for MB subtypes
mb_subtypes <- c("SHH", "G3", "G4")
mb_diff_exp$cluster <- factor(mb_diff_exp$cluster, levels = mb_subtypes)

# set genes with `p_val_adj` of 0 to 1e-300 (the minimum non-zero `p_val_adj` is
# around 3e-294 right now)
min_adj_p <- filter(mb_diff_exp, p_val_adj != 0) %>% pull(p_val_adj) %>% min()
mb_diff_exp <- mutate(
  mb_diff_exp,
  p_val_adj = case_when(
    p_val_adj < min_adj_p ~ 1e-300,
    .default = p_val_adj
  )
)

# show total number of genes tested in each subtype
table(mb_diff_exp$cluster)
```

The minimum non-zero adjusted p-value is `r sprintf("%.2e", min_adj_p)`. All adjusted p-values of 0 were set to 1e-300 so they show up on the volcano plot.

Make volcano plots for each of the subtypes:

```{r}
#| cache: false
#| cache-lazy: false
#| code-fold: true

# add columns for volcano plot
mb_diff_exp_tfs <- map(
  .x = mb_subtypes,
  .f = \(subtype) {
    de <- filter(mb_diff_exp, cluster == subtype & gene %in% tfs) %>%
      mutate(
        direction = case_when(
          avg_log2FC > 0 & p_val_adj < 0.05 ~ "up",
          avg_log2FC < 0 & p_val_adj < 0.05 ~ "down",
          .default = "n.s."
        )
      ) %>%
      arrange(desc(avg_log2FC), p_val_adj)

    # label top 40 upreg and top 20 downreg TFs
    genes_to_label <- c(head(de$gene, n = 40), tail(de$gene, n = 20))
    de <- mutate(
      de,
      gene_label = case_when(
        gene %in% genes_to_label ~ gene,
        .default = ""
      )
    )

    return(de)
  }
) %>%
  setNames(nm = mb_subtypes)

# volcano plots
walk2(
  .x = mb_diff_exp_tfs,
  .y = names(mb_diff_exp_tfs),
  .f = \(de, subtype) {
    # volcano plot
    .plt <- make_volcano(
      data = de,
      log_fc = avg_log2FC,
      log_pval = -log10(p_val_adj),
      direction = direction,
      gene_labels = gene_label,
      log_fc_thresh = 0,
      log_pval_thresh = -log10(0.05),
      seed = 42
    ) + 
      labs(
        x = "log2(FC)",
        y = "-log10(adj. p-value)",
        caption = paste0(subtype, " (TFs only)")
      ) + 
      scale_colour_manual(
        values = c("down" = "blue", "n.s." = "black", "up" = "red")
      ) + 
      theme(plot.background = element_rect(fill = "white"))
    # set text label size
    .plt$layers[[2]]$aes_params$size <- 1.5

    ggsave(
      filename = paste0(subtype, "_TF_volcano.png"),
      plot = .plt,
      path = tumour_out_dir,
      width = 12,
      height = 8,
      units = "in",
      dpi = 600
    )

    return(de)
  }
)
```

### Common upregulated TFs in UBCs and MB

Get the list of upregulated TFs for each UBC subclsuter:

```{r}
ubc_upreg_tfs <- map(
  .x = cluster_marker_ranking,
  .f = \(x) {
    upreg_tfs <- x$feature
    return(upreg_tfs)
  }
)
```

Get the list of upregulated TFs for each MB subtype:

```{r}
mb_upreg_tfs <- map(
  .x = mb_diff_exp_tfs,
  .f = \(de) {
    upreg_tfs <- filter(
      de,
      avg_log2FC > 0 & p_val_adj < 0.05
    ) %>%
      pull(gene)
    return(upreg_tfs)
  }
)
```

Get intersection of TFs:

```{r}
shared_tfs <- expand_grid(ubc_subclusters, mb_subtypes) %>%
  rowwise() %>%
  mutate(
    # get list of shared TFs
    intersection = list(intersect(
      ubc_upreg_tfs[[ubc_subclusters]],
      mb_upreg_tfs[[mb_subtypes]]
    )),
    # get number of shared TFs
    intersection_size = length(intersection),
    # collapse list of shared TFs into single string
    intersection = case_when(
      intersection_size > 0 ~ paste(intersection, collapse = ", "),
      intersection_size == 0 ~ NA
    )
  )
write_csv(
  x = shared_tfs,
  file = file.path(integ_out_dir, "num_shared_tfs.csv")
)

print(shared_tfs, width = Inf)
```
