---
title: Exploring the Sepp et al. 2024 mouse single-cell dataset
date: 2024-04-01
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

Exploratory analysis of the [Sepp et al. 2024](https://doi.org/10.1038/s41586-023-06884-x) (PMID: [36450980](https://pubmed.ncbi.nlm.nih.gov/38029793/)) mouse single-cell dataset to determine what additional downstream processing will be needed prior to integration with other datasets.


## Files and directories

All code and output can be found in this document.


## Results

### Load R packages and dataset as SingleCellExperiment object

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(Seurat)
library(SingleCellExperiment)

options(ggrepel.max.overlaps = Inf)
```

```{r}
#| cache: false
#| cache-lazy: false

sce <- readRDS("/isilon/CBL_scRNAseq-archived/data/src/neurodev-genomics/scRNAseq/Sepp_2024/mou_sce_final.rds")
```

### First look at SingleCellExperiment object

```{r}
#| cache: false
#| cache-lazy: false

print(sce)

head(colData(sce))
```

### Convert SCE to Seurat

```{r}
#| cache: false
#| cache-lazy: false

srat <- as.Seurat(
  sce,
  counts = "umi",
  data = NULL,
  project = "sepp_mouse"
)

srat <- RenameAssays(srat, originalexp = "RNA")
print(srat)
```

Looks like there are 115282 cells and 20684 genes. This is consistent with the number of post-QC cells in the paper. There are two UMAP reductions and no t-SNE reductions.

```{r}
head(srat[[]])

table(srat$orig.ident, useNA = "ifany")
table(srat$broad_lineage, useNA = "ifany")
table(srat$cell_type, useNA = "ifany")
table(srat$dev_state, useNA = "ifany")
table(srat$subtype, useNA = "ifany")
table(srat$precisest_label, useNA = "ifany")
table(srat$Stage, useNA = "ifany")
```

The samples are aged E10.5-adult (P63). We should probably remove all adult cells as those are not relevant to the integration, leaving us with around 105k cells from E10.5-P14. The paper also mentions that the mouse P7/P14 stages correspond to the human newborn/infant stages, respectively. Although all the cells after newborn were filtered out in the human dataset (see 2024-03-26 and 2023-03-29), the P7/P14 cells won't be filtered out for now because the Vladoiu mouse dataset also contains cell up to P14.

### QC plots

When running `RenameAssays` above, some of the metadata columns were duplicated (`nCount_originalexp` to `nCount_RNA`; `nFeature_originalexp` to `nFeature_RNA`). The duplicated columns can be removed later on.

```{r}
sum(srat$nCount_originalexp == srat$nCount_RNA)
sum(srat$nFeature_originalexp == srat$nFeature_RNA)
```

Plot the cells in one group.

```{r}
#| fig-width: 6
#| fig-height: 12

.plt <- VlnPlot(
  srat,
  features = c("nCount_RNA", "nFeature_RNA"),
  combine = FALSE,
  raster = FALSE
)

wrap_plots(.plt, nrow = 2) & NoLegend()
```

```{r}
summary(srat$nCount_RNA)
summary(srat$nFeature_RNA)
```

Some of the `nCount_RNA` and `nFeature_RNA` values appear to be relatively high, but the authors already filtered the cells to get a total of 115282 mouse cells, so no further filtering should be needed at this point.

The `percent.mt` column is missing, so this will need to be calculated from the data and checked later on.

### UMAP

The UMAP embeddings have already been calculated.

```{r}
#| fig-width: 8
#| fig-height: 8.25

DimPlot(
  srat,
  reduction = "umap2d",
  group.by = "broad_lineage",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
) + 
  theme(legend.position = "bottom", legend.justification = "center") + 
  guides(colour = guide_legend(nrow = 1, override.aes = list(size = 3)))
```

```{r}
#| fig-width: 8
#| fig-height: 9.5

DimPlot(
  srat,
  reduction = "umap2d",
  group.by = "cell_type",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
) + 
  theme(legend.position = "bottom", legend.justification = "center") + 
  guides(colour = guide_legend(nrow = 6, override.aes = list(size = 3)))
```

```{r}
#| fig-width: 8
#| fig-height: 8.5

DimPlot(
  srat,
  reduction = "umap2d",
  group.by = "Stage",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
) + 
  theme(legend.position = "bottom", legend.justification = "center") + 
  guides(colour = guide_legend(nrow = 2, override.aes = list(size = 3)))
```

```{r}
#| fig-width: 8
#| fig-height: 8.25

DimPlot(
  srat,
  reduction = "umap2d",
  group.by = "Capture.System",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
) + 
  theme(legend.position = "bottom", legend.justification = "center")
```

The UMAP visualizations shown here appears to match the one in Figure 1c of the paper.

## Next steps for processing

To save the data in a format ready for integration, the following will need to be done:

- convert SCE to Seurat
- filter out cells that were sampled adults
- convert gene names from Ensembl gene IDs to MGI symbols
- remove duplicated metadata columns
- calculate `percent.mt` and check to make sure there aren't any cells with high mitochondrial content
- run `SCTransform` and regress out cell cycle
