---
title: Test pseudoage algorithm on several datasets
date: 2024-05-15
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

Use pseudoage to determine correspondance between human and mouse developmental timepoints. This is the same code as 2024-05-07, but we're checking to make sure the code is working by using different datasets:

1. Sepp et al. (2024) human + mouse dataset, full cerebellum (this is the original paper with the pseudoage method)
1. The fully integrated dataset (3 human + 2 mouse datasets, full cerebellum)


## Files and directories

### Code

The full cerebellum integration Seurat object can be found at `/.mounts/labs/pailab/private/llau/results/integrated/20240512/cca/20240514_cca_integ.qs`.

The code to isolate the Sepp datasets can be found in `<integrated results directory>/20240515/isolate_sepp_from_integ.R`. This script loads the integrated Seurat object, subsets the Sepp cells, and re-runs PCA and UMAP.

The code to calculate pseudoage can be found in `<integrated software directory>/calculate_pseudoage.R`. See notes from 2024-05-07 for more details regarding the script.

The `run_all.sh` script in the output results folder was run interactively on the cluster as it kept crashing when run on the lab server. This script:

1. Creates a subsetted Seurat object with the Sepp human and mouse cells
1. Calculates pseudoage for all cells in the Sepp dataset
1. Calculates pseudoage for all cells in the fully integrated dataset

### Output

The output directory can be found in `<integrated results directory>/20240515`. The Sepp-only Seurat object is saved as `sepp_srat.qs`.

The pseudoages for the Sepp cells can be found in the `sepp/` subdirectory. The pseudoages for all the cells of the integration can be found in the `full/` subdirectory.


## Results

Functions for binning pseudoage and calculating Manhattan distance between true ages. Code is copied from 2024-05-07 notes.

```{r}
#' Bin cells by pseudoage and calculate proportion of cells in each bin at each 
#' developmental timepoint.
#'
#' @param md Metadata table from Seurat containing the `pseudoage` and the
#'   actual human/mouse ages of each cell.
#'
#' @return A dataframe with the proportion of cells in each pseudoage bin
#'   (columns) for each developmental timepoint (rows). Rows should sum to 1.
#'
bin_pseudoage <- function(
  md
) {
  # bin cells by pseudoage
  x <- mutate(
    md,
    pseudoage_bin = cut(pseudoage, breaks = 50, labels = FALSE) %>% as.factor()
  ) %>%
    # keep only certain columns
    select(
      species,
      mouse_age,
      human_age,
      pseudoage_bin
    )

  # calculate proportion of cells in each bin for each developmental timepoint
  x <- x %>%
    mutate(
      all_age = case_when(
        species == "mouse" ~ mouse_age,
        species == "human" ~ human_age
      )
    ) %>%
    group_by(species, all_age, pseudoage_bin) %>%
    count() %>%
    ungroup(pseudoage_bin) %>%
    mutate(
      prop = n / sum(n)
    )

  # reformat so each developmental timepoint is in a row and each bin is in a
  # column
  x <- x %>%
    select(-n) %>%
    pivot_wider(
      names_from = pseudoage_bin,
      values_from = prop,
      names_sort = TRUE,
      values_fill = 0
    ) %>%
    ungroup() %>%
    select(-species) %>%
    column_to_rownames("all_age") %>%

  return(x)
}
```

Load packages:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(patchwork)
library(ComplexHeatmap)
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240515/"
```

### Matching ages using Sepp dataset

Load Seurat object with Sepp human and mouse cells:

```{r}
#| cache: false
#| cache-lazy: false

# sepp_srat <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240515/sepp_srat_10k_test.qs")
sepp_srat <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240515/sepp_srat.qs")
sepp_srat
```

#### Subsetting the Sepp dataset

The cells from the Sepp dataset were subsetted from the integration of the full cerebella from five datasets located at `/.mounts/labs/pailab/private/llau/results/integrated/20240512/cca/20240514_cca_integ.qs`. This was followed by PCA and UMAP.

Plot the PCA elbow plot for all the PCs.

```{r}
#| label: fig-sepp-pca
#| fig-cap: "Elbow plot with the top 100 PCs in the Sepp-only dataset."
#| fig-width: 8
#| fig-height: 6

ElbowPlot(sepp_srat, ndims = 100)
```

Plot the UMAPs of the Sepp dataset to make sure it looks fine before calculating pseudoages.

```{r}
#| label: fig-sepp-umaps
#| fig-cap: "UMAPs of Sepp human and mouse cells. Cells are coloured by species (A), original cell type as annotated by the authors (B), and re-annotated cell type (C)."
#| fig-width: 24
#| fig-height: 12
#| cache: false
#| cache-lazy: false

.plt <- map(
  .x = c("species", "cell_type", "common_cell_name"),
  .f = function(x) {
    p <- DimPlot(
      sepp_srat,
      group.by = x,
      label = TRUE,
      repel = TRUE,
      raster = FALSE
    ) + 
      guides(colour = guide_legend(ncol = 2, override.aes = list(size = 2))) +
      theme(
        legend.position = "bottom",
        legend.justification = c("center", "top")
      )
    
    return(p)
  }
)

wrap_plots(.plt, nrow = 1) + 
  plot_annotation(tag_levels = "A")
```

#### Determining pseudoage similarity

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- sepp_srat@meta.data
rm(sepp_srat)
```

Add pseudoage to the metadata table (`pseudoage.csv` generated with `calculate_pseudoage.R`).

```{r}
pseudoage <- read_csv(file.path(out_dir, "sepp/pseudoage.csv")) %>%
  column_to_rownames("cell_id")
md <- merge(md, pseudoage, by = "row.names", all = TRUE) %>%
  column_to_rownames("Row.names")
head(md)
```

Bin pseudoages and get proportion of cells in each bin.

```{r}
prop_bin <- bin_pseudoage(md)
head(prop_bin)
```

Calculate Manhattan distance for all pairwise developmental timepoints.

```{r}
pseudoage_dist <- prop_bin %>%
  dist(method = "manhattan", diag = TRUE, upper = TRUE) %>%
  as.matrix()
pseudoage_dist[1:5, 1:5]
write.csv(
  pseudoage_dist,
  file = file.path(out_dir, "sepp/pseudoage_dist.csv")
)
```

#### Comparing developmental stages between human and mouse

To compare human vs. mouse developmental stages, we first need to filter the distance matrix. We remove the human stages from the columns and the mouse stages from the rows.

```{r}
# get human/mouse developmental stages
human_ages <- names(table(md$human_age))
mouse_ages <- names(table(md$mouse_age))

# remove them from distance matrix
hm_dist <- pseudoage_dist[
  !rownames(pseudoage_dist) %in% mouse_ages,
  !colnames(pseudoage_dist) %in% human_ages
]

hm_dist <- hm_dist[
  # sort the rows so "7/8/9 PCW" show up first
  str_sort(rownames(hm_dist), numeric = TRUE), 
  # sort the columns so "P14" shows up last
  str_sort(colnames(hm_dist), numeric = TRUE)
]
hm_dist
```

Now we use a heatmap to plot the distances between the human and mouse pseudoages.

```{r}
#| label: fig-sepp-heatmap-pseudoage
#| fig-cap: "Heatmap showing pairwise Manhattan distances between human and mouse pseudoages in the Sepp dataset."
#| fig-width: 7
#| fig-height: 4

# set colour function for heatmap
breaks <- seq(0, 2, by = 0.25)
col_fun <- circlize::colorRamp2(
  breaks = breaks,
  colors = scales::pal_viridis(direction = -1, option = "inferno")(length(breaks))
)

.plt <- Heatmap(
  matrix = hm_dist,
  col = col_fun,
  row_title = "human",
  row_title_side = "left",
  column_title = "mouse",
  column_title_side = "bottom",
  row_order = rev(rownames(hm_dist)),
  column_order = colnames(hm_dist),
  row_names_side = "left",
  column_names_side = "bottom",
  heatmap_legend_param = list(
    title = "",
    legend_height = unit(2.5, "in")
  ),
  width = unit(ncol(hm_dist)*0.5, "in"),
  height = unit(nrow(hm_dist)*0.5, "in")
)
.plt
```

In general, the heatmap in @fig-sepp-heatmap-pseudoage appears to match the original heatmap ([Extended Data Fig. 2f, Sepp et al. 2024](https://www.nature.com/articles/s41586-023-06884-x/figures/6)) relatively well, suggesting that the code being used to determine pseudoage similarity across developmental timepoints is working.

### Matching ages using the full integrated dataset

Here, we use the integration of the full cerebellum from 5 datasets to determine the best matching developmental timepoints between human and mouse. The datasets are from:

- Aldinger et al. (human)
- Luo et al. (human)
- Sepp et al. (two datasets - one human and one mouse)
- Vladoiu et al. (mouse)

The metadata from the Seurat object was extracted using the cluster since the full Seurat object is too big to be loaded on the lab server.

```{r}
#| eval: false

# run separately on cluster
full_srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240512/cca/20240514_cca_integ.qs")
qs::qsave(
  x = full_srat[[]],
  file = file.path(out_dir, "full_cerebellum_metadata.qs")
)
```

Load the Seurat object metadata:

```{r}
md <- qs::qread(file.path(out_dir, "full_cerebellum_metadata.qs"))
head(md)
dim(md)
```

The full cerebellum has `r nrow(md)` cells.

#### Determining pseudoage similarity

Add pseudoage to the metadata table (`pseudoage.csv` generated with `calculate_pseudoage.R`).

```{r}
pseudoage <- read_csv(file.path(out_dir, "full/pseudoage.csv")) %>%
  column_to_rownames("cell_id")
md <- merge(md, pseudoage, by = "row.names", all = TRUE) %>%
  column_to_rownames("Row.names")
head(md)
```

Show actual human/mouse ages.

```{r}
table(md$human_age, useNA = "always")
table(md$mouse_age, useNA = "always")
```

Bin pseudoages and get proportion of cells in each bin.

```{r}
prop_bin <- bin_pseudoage(md)
dim(prop_bin)
head(prop_bin)
```

Calculate Manhattan distance for all pairwise developmental timepoints.

```{r}
pseudoage_dist <- prop_bin %>%
  dist(method = "manhattan", diag = TRUE, upper = TRUE) %>%
  as.matrix()
dim(pseudoage_dist)
pseudoage_dist[1:5, 1:5]
write.csv(
  pseudoage_dist,
  file = file.path(out_dir, "full/pseudoage_dist.csv")
)
```

#### Comparing developmental stages between human and mouse

To compare human vs. mouse developmental stages, we first need to filter the distance matrix. We remove the human stages from the columns and the mouse stages from the rows.

```{r}
# get human/mouse developmental stages
human_ages <- names(table(md$human_age))
mouse_ages <- names(table(md$mouse_age))

# remove them from distance matrix
hm_dist <- pseudoage_dist[
  !rownames(pseudoage_dist) %in% mouse_ages,
  !colnames(pseudoage_dist) %in% human_ages
]

hm_dist <- hm_dist[
  # sort the rows so "7/8/9 PCW" show up first
  str_sort(rownames(hm_dist), numeric = TRUE), 
  # sort the columns so "P14" shows up last
  str_sort(colnames(hm_dist), numeric = TRUE)
]
hm_dist
```

Now we use a heatmap to plot the distances between the human and mouse pseudoages.

```{r}
#| label: fig-full-heatmap-pseudoage
#| fig-cap: "Heatmap showing pairwise Manhattan distances between human and mouse pseudoages across all integrated datasets."
#| fig-width: 10
#| fig-height: 8

# set colour function for heatmap
breaks <- seq(0, 2, by = 0.25)
col_fun <- circlize::colorRamp2(
  breaks = breaks,
  colors = scales::pal_viridis(direction = -1, option = "inferno")(length(breaks))
)

.plt <- Heatmap(
  matrix = hm_dist,
  col = col_fun,
  name = "full_cerebellum",
  row_title = "human",
  row_title_side = "left",
  column_title = "mouse",
  column_title_side = "bottom",
  row_order = rev(rownames(hm_dist)),
  column_order = colnames(hm_dist),
  row_names_side = "left",
  column_names_side = "bottom",
  heatmap_legend_param = list(
    title = "",
    legend_height = unit(4, "in")
  ),
  width = unit(ncol(hm_dist)*0.5, "in"),
  height = unit(nrow(hm_dist)*0.5, "in")
)
.plt

decorate_heatmap_body(
  heatmap = "full_cerebellum",
  code = {
    x = unit(c(7, 9, 12) / ncol(hm_dist), "npc")
    y = unit(c(6, 10, 14) / nrow(hm_dist), "npc")
    w = unit(c(2, 3, 4) / ncol(hm_dist), "npc")
    h = unit(c(2, 4, 4) / nrow(hm_dist), "npc")

    # draw 3 boxes
    for (i in c(1:3)) {
      grid.rect(
        x = x[i],
        y = y[i],
        width = w[i],
        height = h[i],
        just = c("left", "top"),
        gp = gpar(col = "#4DFFFF", fill = "#00000000", lty = "solid", lwd = 4)
      )
    }
  }
)
```

Overall, matching developmental stages using pseudoage with all the datasets (@fig-full-heatmap-pseudoage) appears to be consistent with the previous analysis from Sepp et al. (Extended Data Fig. 2f). Although shown here, for our integration, we removed all human cells from 7-10 PCW as: (1) one of the Luo samples was labelled as "male" yet showed *XIST* expression, and (2) the UBCs don't develop until later (around E12.5 in mice according to Englund et al. 2006).

For our purposes, we can try grouping the cells into:

- 1 - E14.5-E15.5 and 11-12 PCW
- 2 - E16-E18 and 13-16 PCW
- 3 - P0-P7 and 17-21 PCW

This roughly matches the corresponding stages from Sepp et al. (Fig. 1), although they are missing samples from 11-17 PCW and E16-E17.

```{r}
table(
  md$dataset_name[str_starts(md$dataset_name, "Sepp")],
  md$mouse_age[str_starts(md$dataset_name, "Sepp")]
)
table(
  md$dataset_name[str_starts(md$dataset_name, "Sepp")],
  md$human_age[str_starts(md$dataset_name, "Sepp")]
)
```

Interestingly, the mouse cells from P4 and P14 don't appear to match very well with any of the other human timepoints (@fig-full-heatmap-pseudoage). While all the P4 cells are from the Sepp dataset, the P14 cells are from both the Sepp and the Vladoiu datasets.

```{r}
table(
  md$dataset_name[md$mouse_age %in% c("P4", "P14")],
  md$mouse_age[md$mouse_age %in% c("P4", "P14")]
)
```
