---
title: Testing for differences in cell type proportions between species
date: 2024-06-17
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine if there is a statistically significant difference in cell type proportions between human and mouse. Similar to 2024-06-06, except that here we look at the cell type proportion of all RL lineage cells (**excluding** the controls, like OPCs, microglia, and endothelial cells).


## Files and directories

### Code

All code can be found in this document. The code was run using the `propeller` conda environment.

We perform the proportion test on two integrated objects:

- `/.mounts/labs/pailab/private/llau/results/integrated/20240524/25_pc_without_luo/25_pc_rl.qs`: the RL lineage cells of the four-dataset full cerebellum **CCA** integration (Aldinger/human, Sepp/human, Sepp/mouse, Vladoiu/mouse)
  - `propeller` was run both with the UBC subclusters and without the UBC subclusters (UBCs as one cluster)
- `/.mounts/labs/pailab/private/llau/results/integrated/20240606/FC_integ_analysis/rpca_20/rl.qs`: the RL lineage cells of the four-dataset full cerebellum **RPCA** integration (Aldinger/human, Sepp/human, Sepp/mouse, Vladoiu/mouse)

`propeller` was run using the default logit transformation; according to the author, this better controls the type I error rate with small sample sizes ([Phipson et al. 2022](https://academic.oup.com/bioinformatics/article/38/20/4720/6675456)).

### Output

The output directory can be found at `<integrated results directory>/20240617`. This contains the tables of results that were generated by running `propeller` in this document. There are two types of files:

- `*_proportions.csv` contains the proportion of cells in each cluster for each sample (i.e., the proportions for each sample add up to 1); this is (part of) the output of `getTransformedProps`
- `*_results.csv` contains the output of the `propeller` test for each cluster, including the p-values and the FDRs


## Results

### Load packages

Load packages and other functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(speckle)
library(patchwork)

# helper functions
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/cluster_barplot.R")
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/propeller_helpers.R")
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240617/"
```

Some variables and helper functions (click below to expand):

```{r}
#| code-fold: true

# set colour for the species
species_cols <- c("grey", scales::pal_hue()(1))

# make UMAPs and bar plot showing the species and clusters of a Seurat object
make_summary_fig <- function(
  srat,
  cluster
) {
  # UMAP showing clusters
  .plt1 <- DimPlot(
    srat,
    group.by = cluster,
    label = TRUE,
    repel = TRUE,
    raster = FALSE
  )

  # UMAP showing human/mouse cells
  human_cells <- rownames(srat[[]])[srat$species == "human"]
  .plt2 <- DimPlot(
    srat,
    group.by = "species",
    cells.highlight = human_cells,
    sizes.highlight = 0.01,
    label = FALSE,
    raster = FALSE
  ) + 
    labs(title = NULL) + 
    scale_colour_manual(labels = c("mouse", "human"), values = species_cols)

  # cluster_barplot.R
  srat$species <- fct_relevel(srat$species, "mouse", "human")
  .plt3 <- cluster_barplot(
    srat,
    split.by = "species",
    group.by = cluster,
    position = "stack"
  ) + 
    scale_fill_manual(values = species_cols)

  .plt <- (.plt1 + .plt2) / .plt3 + 
    plot_annotation(tag_levels = "A") & 
    theme(plot.tag = element_text(face = "bold"))

  return(.plt)
}
```

### Cell type proportions in CCA

Load Seurat objects:

```{r}
#| cache: false
#| cache-lazy: false

cca_srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240524/25_pc_without_luo/25_pc_rl.qs")
cca_srat

# also need the UBC subclusters to copy them over
cca_srat_ubc_sub <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240527/ubc_subset.qs")
cca_srat_ubc_sub
```

Show quick UMAP and bar plot of the RL lineage cells (including controls cells - OPCs, microglia, endothelial).

```{r}
#| label: fig-summary-cca
#| fig-cap: "Overview of the RL lineage cells (CCA integration). (A-B) UMAP of RL lineage cells with cells coloured by (A) cluster (0.4 resolution) or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

make_summary_fig(cca_srat, "snn_res.0.4") + 
  labs(caption = "(CCA integration)")
```

What's the total number of human vs. mouse cells?

```{r}
table(cca_srat$species)
```

Only the Seurat metadata is required - save it and delete the Seurat object. The following "control" cells are also filtered out, leaving only the RL lineage cells:

- clusters 11/15 - oligodendrocytes
- cluster 13 - endothelial cells
- cluster 14 - microglia

```{r}
md <- cca_srat@meta.data %>%
  # rename column for easier access to cell type clusters
  rename(
    rl_clusters = snn_res.0.4
  )

# make sure factor levels for the 0.4 resolution clusters are in numeric order
md$rl_clusters <- fct_relevel(
  md$rl_clusters,
  str_sort(levels(md$rl_clusters), numeric = TRUE)
)

# add UBC subclusters to the metadata
md <- merge(
  x = md,
  y = select(cca_srat_ubc_sub[[]], snn_res.0.3) %>%
    rename(ubc_subclusters = snn_res.0.3),
  by = "row.names",
  all.x = TRUE
) %>%
  # filter out non-RL lineage cells (oligodendrocytes, endothelial, microglia)
  mutate(
    # combine UBC subclusters with other clusters
    final_clusters = case_when(
      !is.na(ubc_subclusters) ~ paste0("UBC_", ubc_subclusters),
      .default = rl_clusters
    ),
    # set factor levels for the new column
    final_clusters = fct_relevel(
      final_clusters,
      str_sort(names(table(final_clusters)), numeric = TRUE)
    ),
    rl_clusters = fct_drop(rl_clusters)
  ) %>%
  column_to_rownames(var = "Row.names") %>%
  # filter out non-RL lineage cells (oligodendrocytes, endothelial, microglia)
  filter(!(final_clusters %in% c(11, 13, 14, 15)))
md$final_clusters <- fct_drop(md$final_clusters)

head(md)

# delete Seurat objects
rm(cca_srat, cca_srat_ubc_sub)
```

Show the new column with the UBC subclusters:

```{r}
table(md$final_clusters)
```

#### Run `propeller` on the RL lineage cells (with UBC subclusters)

To run `propeller`, we need the cluster, sample, and group of each cell. The cluster information is in the `final_clusters` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset_name, "^Aldinger")],
  md$orig.ident[str_which(md$dataset_name, "^Aldinger")]
)
table(md$dataset_name, md$orig.ident)

# prep for propeller
x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = final_clusters
) %>%
  select(cluster, sample, species, dataset_name)

# filtering out samples with fewer than 50 cells, similar to Sepp et al.
samples_to_filter <- names(table(x$sample)[table(x$sample) < 50])
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
cca_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = cca_results,
  file = file.path(out_dir, "cca_results_with_ubc_subclusters.csv")
)

cca_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-cca
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 12
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset
sample_metadata <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get proportion of cells in each cluster for each sample
cca_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  sample_meta_df = sample_metadata
)
write_csv(
  cca_props,
  file = file.path(out_dir, "cca_proportions_with_ubc_subclusters.csv")
)

dim(cca_props)
head(cca_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(cca_props, facet_cols = dataset_name) +  
  labs(caption = "(CCA integration with UBC subclusters)")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-cca
#| fig-cap: "Proportion of cells in each of the clusters out of all RL lineage cells. Each dot represents a sample from the dataset."
#| fig-width: 8
#| fig-height: 20
#| code-fold: true

# FDRs to label on the plot
fdrs <- get_fdrs_for_plotting(cca_results)

# plot proportions
boxplot_props_per_cluster(
  props = cca_props,
  facet_var = clusters,
  fdrs = fdrs
) + 
  labs(caption = "(CCA integration with UBC subclusters)") + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(7/8, 1/12)
  )
```

#### Run `propeller` on the RL lineage cells (without UBC subclusters)

Here, we run `propeller` with just the broad RL clusters (i.e., UBCs as cluster 2, not as subclusters).

```{r}
# prep for propeller
x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = rl_clusters
) %>%
  select(cluster, sample, species, dataset_name)

# filtering out samples with fewer than 50 cells, similar to Sepp et al.
samples_to_filter <- names(table(x$sample)[table(x$sample) < 50])
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
cca_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = cca_results,
  file = file.path(out_dir, "cca_results_without_ubc_subclusters.csv")
)

cca_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-cca-no-sub
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample. Similar to @fig-prop-cca except that the UBCs here are not split into subclusters."
#| fig-width: 12
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset
sample_metadata <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get proportion of cells in each cluster for each sample
cca_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  sample_meta_df = sample_metadata
)
write_csv(
  cca_props,
  file = file.path(out_dir, "cca_proportions_without_ubc_subclusters.csv")
)

dim(cca_props)
head(cca_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(cca_props, facet_cols = dataset_name) +  
  labs(caption = "(CCA integration)")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-cca-no-sub
#| fig-cap: "Proportion of cells in each of the clusters out of all RL lineage cells. Each dot represents a sample from the dataset. Similar to @fig-box-cca except that the UBCs here are not split into subclusters."
#| fig-width: 8
#| fig-height: 16
#| code-fold: true

# FDRs to label on the plot
fdrs <- get_fdrs_for_plotting(cca_results)

# plot proportions
boxplot_props_per_cluster(
  props = cca_props,
  facet_var = clusters,
  fdrs = fdrs
) + 
  labs(caption = "(CCA integration)") + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(7/8, 1/8)
  )
```

### Cell type proportions in RPCA

Load Seurat objects:

```{r}
#| cache: false
#| cache-lazy: false

rpca_srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240606/rpca_20_analysis/rl/rl.qs")
rpca_srat

# also need the UBC subclusters to copy them over
rpca_srat_ubc_sub <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240606/rpca_20_analysis/ubc/ubc.qs")
rpca_srat_ubc_sub
```

Show quick UMAP and bar plot of the RL lineage cells (including controls cells - OPCs, microglia, endothelial).

```{r}
#| label: fig-summary-rpca
#| fig-cap: "Overview of the RL lineage cells (RPCA integration). (A-B) UMAP of RL lineage cells with cells coloured by (A) cluster (0.2 resolution) or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

make_summary_fig(rpca_srat, "snn_res.0.2") + 
  labs(caption = "(RPCA integration)")
```

What's the total number of human vs. mouse cells?

```{r}
table(rpca_srat$species)
```

Only the Seurat metadata is required - save it and delete the Seurat object. The following "control" cells are also filtered out, leaving only the RL lineage cells:

- clusters 8/12 - oligodendrocytes
- cluster 10 - endothelial cells
- clluster 11 - microglia

```{r}
md <- rpca_srat@meta.data %>%
  # rename column for easier access to cell type clusters
  rename(
    rl_clusters = snn_res.0.2
  )

# make sure factor levels for the 0.4 resolution clusters are in numeric order
md$rl_clusters <- fct_relevel(
  md$rl_clusters,
  str_sort(levels(md$rl_clusters), numeric = TRUE)
)

# add UBC subclusters to the metadata
md <- merge(
  x = md,
  y = select(rpca_srat_ubc_sub[[]], snn_res.0.6) %>%
    rename(ubc_subclusters = snn_res.0.6),
  by = "row.names",
  all.x = TRUE
) %>%
  mutate(
    # combine UBC subclusters with other clusters
    final_clusters = case_when(
      !is.na(ubc_subclusters) ~ paste0("UBC_", ubc_subclusters),
      .default = rl_clusters
    ),
    # set factor levels for the new column
    final_clusters = fct_relevel(
      final_clusters,
      str_sort(names(table(final_clusters)), numeric = TRUE)
    )
  ) %>%
  column_to_rownames(var = "Row.names") %>%
  # filter out non-RL lineage cells (oligodendrocytes, endothelial, microglia)
  filter(!(final_clusters %in% c(8, 10, 11, 12)))
md$final_clusters <- fct_drop(md$final_clusters)

head(md)

# delete Seurat objects
rm(rpca_srat, rpca_srat_ubc_sub)
```

Show the new column with the UBC subclusters:

```{r}
table(md$final_clusters)
```

#### Run `propeller` on the RL lineage cells

To run `propeller`, we need the cluster, sample, and group of each cell. The cluster information is in the `final_clusters` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset_name, "^Aldinger")],
  md$orig.ident[str_which(md$dataset_name, "^Aldinger")]
)
table(md$dataset_name, md$orig.ident)

# prep for propeller
x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = final_clusters
) %>%
  select(cluster, sample, species, dataset_name)

# filtering out samples with fewer than 50 cells, similar to Sepp et al.
samples_to_filter <- names(table(x$sample)[table(x$sample) < 50])
x <- filter(x, !sample %in% samples_to_filter)

# show samples that were filtered out
samples_to_filter
```

Run `propeller`:

```{r}
rpca_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = rpca_results,
  file = file.path(out_dir, "rpca_results.csv")
)

rpca_results
```

We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-rpca
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 12
#| fig-height: 6
#| code-fold: true

# get a table that matches each sample to a dataset
sample_metadata <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get proportion of cells in each cluster for each sample
rpca_props <- get_props_per_sample(
  clusters = x$cluster,
  sample = x$sample,
  sample_meta_df = sample_metadata
)
write_csv(
  rpca_props,
  file = file.path(out_dir, "rpca_proportions.csv")
)

dim(rpca_props)
head(rpca_props)

# plot proportion of cells in each cluster for each sample
plot_props_per_sample(rpca_props, facet_cols = dataset_name) + 
  labs(caption = "(RPCA integration)")
```

We can also plot the proportions of each sample in each cluster:

```{r}
#| label: fig-box-rpca
#| fig-cap: "Proportion of cells in each of the clusters out of all RL lineage cells. Each dot represents a sample from the dataset."
#| fig-width: 8
#| fig-height: 20
#| code-fold: true

# FDRs to label on the plot
fdrs <- get_fdrs_for_plotting(rpca_results)

# plot proportions
boxplot_props_per_cluster(
  props = rpca_props,
  facet_var = clusters,
  fdrs = fdrs
) + 
  labs(caption = "(RPCA integration)") + 
  theme(
    legend.position = "inside",
    legend.position.inside = c(7/8, 1/12)
  )
```
