---
title: Potential confounders for UBC subclusters
date: 2024-09-16
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To investigate the sex, age, and dataset breakdown of the UBC subclusters.


## Files and directories

### Code

All code can be found in this document. The code was run using the `scrnaseq_env` conda environment.

### Output

The output directory can be found at `<integrated results directory>/20240916`


## Results

Load packages and functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(Seurat)

# helper functions
source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/plotting.R")
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240916/"
```

Load files:

```{r}
#| cache: false
#| cache-lazy: false

# load Seurat object with integrated UBCs
ubc_srat <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240825/ubc_subset.qs")

# drop factor levels with no cells
ubc_srat$human_age <- fct_drop(ubc_srat$human_age)
ubc_srat$mouse_age <- fct_drop(ubc_srat$mouse_age)
ubc_srat$combined_age <- fct_drop(ubc_srat$combined_age)

# clean up dataset column
ubc_srat$dataset_name <- str_remove(ubc_srat$dataset_name, "full_cerebellum_")

# load sex metadata by sample (compiled by Leo, see his notes from 2024-07-05
# and 2024-07-23)
sample_sex_df <- read_csv("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/scrnaseq_Leo/utilities/dataset_sexes.csv") %>%
  mutate(dataset_name = str_remove(dataset_name, "full_cerebellum_"))

# add sample and sex to Seurat object
ubc_srat@meta.data <- mutate(
  ubc_srat[[]],
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  sex = case_when(
    !str_detect(dataset_name, "^Aldinger") ~ 
      sample_sex_df$sex[match(sample, sample_sex_df$orig.ident)],
    .default = sex
  )
)
```

### Breakdown by sex

```{r}
#| label: fig-sex-breakdown
#| fig-cap: "Bar plot showing the proportion of cells in each UBC subcluster by sex."
#| fig-width: 8
#| fig-height: 6
#| code-fold: true

.plt <- cluster_barplot(
  object = ubc_srat,
  split.by = "sex",
  group.by = "subclusters"
) + 
  labs(title = "sex")

ggsave(
  filename = "sex_breakdown.png",
  plot = .plt,
  path = out_dir,
  width = 8,
  height = 6,
  units = "in",
  dpi = 600
)

.plt
```

### Breakdown by age

```{r}
#| label: fig-age-breakdown
#| fig-cap: "Bar plots showing the proportion of cells in each UBC subcluster by age."
#| fig-width: 8
#| fig-height: 12
#| code-fold: true

mouse_age_bar <- cluster_barplot(
  object = ubc_srat,
  split.by = "mouse_age",
  group.by = "subclusters"
) + 
  labs(title = "mouse")

human_age_bar <- cluster_barplot(
  object = ubc_srat,
  split.by = "human_age",
  group.by = "subclusters"
) + 
  labs(title = "human")

.plt <- mouse_age_bar / human_age_bar

ggsave(
  filename = "age_breakdown.png",
  plot = .plt,
  path = out_dir,
  width = 8,
  height = 12,
  units = "in",
  dpi = 600
)

.plt
```

### Breakdown by dataset

```{r}
#| label: fig-dataset-breakdown
#| fig-cap: "Bar plot showing the proportion of cells in each UBC subcluster by dataset."
#| fig-width: 8
#| fig-height: 6
#| code-fold: true

.plt <- cluster_barplot(
  object = ubc_srat,
  split.by = "dataset_name",
  group.by = "subclusters"
) + 
  labs(title = "dataset")

ggsave(
  filename = "dataset_breakdown.png",
  plot = .plt,
  path = out_dir,
  width = 8,
  height = 6,
  units = "in",
  dpi = 600
)

.plt
```
