---
title: Differential gene expression along pseudotime trajectory
date: 2024-07-15
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine differentially expressed genes along the Monocle pseudotime trajectories.


## Files and directories

### Code

The code can be found in this document.

### Output


## Results

### Load packages

Load packages:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(patchwork)
library(monocle3)
```

Load Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

cds <- qs::qread("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240611/human_pseudotime/post_pseudotime_cds.qs")
cds
```

### Differential gene expression for cell-type clusters

```{r}
# Moran's I
knn_graph_test_res <- graph_test(cds, neighbor_graph = "knn", cores = 4)
head(knn_graph_test_res)
```

There seems to be some error when running `graph_test` above causing the test to fail, need to fix that issue.

### Differential gene expression for pseudotime

```{r}
# Moran's I
pg_graph_test_res <- graph_test(cds, neighbor_graph = "principal_graph", cores = 4)
head(pg_graph_test_res)
```

