---
title: Testing for differences in cell type proportion between species
date: 2024-05-28
metadata-files: 
  - _metadata.yml
  - ../_quarto.yml
---

## Goal of this analysis

To determine if there is a statistically significant difference in UBC proportion between human and mouse.


## Files and directories

### Code

All code can be found in this document. This code was run using the `propeller` conda environment.

The input Seurat object is located in `/.mounts/labs/pailab/private/llau/results/integrated/20240524/25_pc_without_luo/25_pc_rl.qs`. This contains the RL lineage cells subsetted from the integration of four full cerebellum datasets (Aldinger/human, Sepp/human, Sepp/mouse, Vladoiu/mouse).

### Output

The output directory can be found at `<integrated results directory>/20240528`. This contains the tables of results that were generated by running `propeller` in this document. There are two types of files:

- `*_proportions.csv` contains the proportion of cells in each cluster for each sample (i.e., the proportions for each sample add up to 1); this is (part of) the output of `getTransformedProps`
- `*_results.csv` contains the output of the `propeller` test for each cluster, including the p-values and the FDRs


## Results

### Load packages and Seurat object

Load packages and other functions:

```{r}
#| cache: false
#| cache-lazy: false

library(tidyverse)
library(Seurat)
library(speckle)
library(patchwork)
library(ComplexHeatmap)

source("/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/software/utilities/cluster_barplot.R")
```

Load Seurat object:

```{r}
#| cache: false
#| cache-lazy: false

srat <- qs::qread("/.mounts/labs/pailab/private/llau/results/integrated/20240524/25_pc_without_luo/25_pc_rl.qs")
srat
```

Set output directory:

```{r}
out_dir <- "/.mounts/labs/pailab/private/icheong/CBL_scRNAseq/results/integrated/20240528/"
```

Show quick UMAP and bar plot with the clusters. The UBCs are in cluster 2.

```{r}
#| label: fig-summary
#| fig-cap: "Overview of the RL lineage cells. (A-B) UMAP of RL lineage cells with cells coloured by (A) cluster (0.4 resolution) or (B) species. (C) Bar plot showing number of human and mouse cells in each cluster."
#| fig-width: 12
#| fig-height: 9
#| cache: false
#| cache-lazy: false

# set colour for the species
species_cols <- c("grey", scales::pal_hue()(1))

.plt1 <- DimPlot(
  srat,
  group.by = "snn_res.0.4",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
)

human_cells <- rownames(srat[[]])[srat$species == "human"]
.plt2 <- DimPlot(
  srat,
  group.by = "species",
  cells.highlight = human_cells,
  sizes.highlight = 0.01,
  label = FALSE,
  raster = FALSE
) + 
  labs(title = NULL) + 
  scale_colour_manual(labels = c("mouse", "human"), values = species_cols)

# cluster_barplot.R
srat$species <- fct_relevel(srat$species, "mouse", "human")
.plt3 <- cluster_barplot(
  srat,
  split.by = "species",
  group.by = "snn_res.0.4",
  position = "stack"
) + 
  scale_fill_manual(values = species_cols)

(.plt1 + .plt2) / .plt3 + 
  plot_annotation(tag_levels = "A") & 
  theme(plot.tag = element_text(face = "bold"))
```

What's the total number of human vs. mouse cells?

```{r}
table(srat$species)
```

Only the Seurat metadata is required - save it and delete the Seurat object.

```{r}
md <- srat@meta.data

# make sure factor levels for the 0.4 resolution clusters are in numeric order
md$snn_res.0.4 <- fct_relevel(
  md$snn_res.0.4,
  str_sort(levels(md$snn_res.0.4), numeric = TRUE)
)

head(md)

# delete Seurat object
rm(srat)
```

### Running propeller on the entire RL dataset {#sec-human-vs-mouse}

The input to `propeller` requires three pieces of information:

- cluster (i.e., cell type clusters)
- sample (i.e., biological replicates)
- group (i.e., human vs. mouse in this case)

Here, we first check if there is a significant difference between the species across all timepoints. The cluster information is in the `snn_res.0.4` column and the group information is in the `species` column. The sample information needs to be collected from various columns.

::: {.callout-important}
Note that for the Vladoiu mouse dataset, only the ages of the cells are available - the number of samples from each age was not specified in the paper. For this analysis, we assume each age comes from a single sample.
:::

```{r}
table(
  md$sample_id[str_which(md$dataset_name, "^Aldinger")],
  md$orig.ident[str_which(md$dataset_name, "^Aldinger")]
)
table(md$dataset_name, md$orig.ident)

x <- mutate(
  md,
  sample = case_when(
    str_detect(dataset_name, "^Aldinger") ~ sample_id,
    .default = orig.ident
  ),
  cluster = snn_res.0.4
) %>%
  select(cluster, sample, species, dataset_name) %>%
  # filtering out samples with fewer than 50 cells, similar to Sepp et al.
  filter(!sample %in% names(table(md$orig.ident)[table(md$orig.ident) < 50]))

# show samples that were filtered out
names(table(md$orig.ident)[table(md$orig.ident) < 50])
```

Run `propeller` with the default logit transformation (according to the author, this better controls the type I error rate with small sample sizes ([Phipson et al. 2022](https://academic.oup.com/bioinformatics/article/38/20/4720/6675456))).

```{r}
all_timepoints_results <- propeller(
  clusters = x$cluster,
  sample = x$sample,
  group = x$species,
  transform = "logit"
)

write_csv(
  x = all_timepoints_results,
  file = file.path(out_dir, "all_timepoints_results.csv")
)

all_timepoints_results
```

For the UBC cluster (cluster 2), there appears to be a significantly higher proportion of human cells compared to mouse cells (FDR = `r formatC(all_timepoints_results["2", "FDR"], digits = 2, format = "e")`). We can plot the actual proportions for each sample:

```{r}
#| label: fig-prop-all-timepoints
#| fig-cap: "Bar plot showing cell type proportions for each sample. See one of the tables above for the number of cells in each sample."
#| fig-width: 12
#| fig-height: 6

# get a table that matches each sample to a dataset
dataset_sample_matching <- table(
  x$dataset_name, x$sample, x$species
) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(dataset_name = Var1, sample = Var2, species = Var3) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum"),
    species = fct_relevel(species, "mouse", "human")
  )

# get the cluster proportions per sample for plotting and save to CSV
all_timepoints_props <- getTransformedProps(
  clusters = x$cluster,
  sample = x$sample,
  transform = "logit"
) %>%
  pluck("Proportions") %>% # get this list item
  as.data.frame() %>%
  rename(proportion = Freq) %>%
  merge(x = ., y = dataset_sample_matching, by = "sample", all.x = TRUE)
write_csv(
  all_timepoints_props,
  file = file.path(out_dir, "all_timepoints_proportions.csv")
)

dim(all_timepoints_props)
head(all_timepoints_props)

# make the plot
.plt <- ggplot(
  all_timepoints_props,
  aes(x = sample, y = proportion, fill = clusters)
) + 
  geom_col(position = "stack") + 
  facet_grid(cols = vars(dataset_name), scales = "free", space = "free") + 
  scale_y_continuous(expand = expansion(0, 0.05)) + 
  scale_fill_manual(
    values = pals::cols25(n = length(unique(all_timepoints_props$clusters)))
  ) + 
  theme_minimal() + 
  theme(
    axis.text.x = element_text(hjust = 1, angle = 60)
  )

.plt
```

We can also plot the actual proportions of UBCs:

```{r}
#| label: fig-ubc-prop-all-timepoints
#| fig-cap: "Proportion of UBCs in each sample across all developmental timepoints."
#| fig-width: 4
#| fig-height: 6

# select UBCs
ubc_props <- all_timepoints_props[all_timepoints_props$clusters == "2", ] %>%
  remove_rownames()

.plt <- ggplot(ubc_props, aes(x = species, y = proportion)) + 
  geom_boxplot(width = 0.6, outlier.shape = NA) + 
  geom_jitter(aes(shape = dataset_name), width = 0.2) + 
  geom_text(
    data = data.frame(x = "mouse", y = 0.62, ubc_fdr = all_timepoints_results["2", "FDR"]),
    aes(x = "mouse", y = 0.62, label = sprintf("adj. p = %.3g", ubc_fdr)),
    nudge_x = 0.5,
    inherit.aes = FALSE
  ) + 
  theme_light() + 
  theme(
    axis.text = element_text(colour = "black")
  )

.plt

# number of samples from each dataset and species
table(dataset_sample_matching$dataset_name, dataset_sample_matching$specie)
```

However, the question remains, is there a specific timepoint at which this is happening?

### Running propeller on the RL dataset for each corresponding age

Using the pseudoage analysis from 2024-05-27, we match the human and mouse ages to a common "developmental stage":

- 1 - E14.5-E15.5 and 11 PCW
- 2 - E16-E18 and 12-16 PCW
- 3 - P0-P7 and 17-21 PCWs

Group the cells into these three common developmental stages and filter out any cells that don't fall into these stages (i.e., cells before E14.5, and cells from P4 and P14):

```{r}
y <- mutate(
  md,
  common_stage = case_when(
    mouse_age %in% c("E14.5", "E15.5") | human_age %in% c("11 PCW") ~ 1,
    mouse_age %in% c("E16", "E17.5", "E18") | human_age %in% paste(c(12, 14, 16), "PCW") ~ 2,
    mouse_age %in% c("P0", "P5", "P7") | human_age %in% paste(c(17, 18, 20, 21), "PCW") ~ 3,
    .default = NA
  )
)
```

Make sure all the cells were assigned to the right stage:

```{r}
table(y$human_age, y$common_stage)
table(y$mouse_age, y$common_stage)
```

Now merge the stage info with our table from before:

```{r}
y <- select(.data = y, common_stage) %>% filter(!is.na(common_stage))
x <- merge(x = x, y = y, by = "row.names", all.y = TRUE) %>%
  column_to_rownames("Row.names")
head(x)
```

Here are the number of cells from each timepoint:

```{r}
table(md$human_age)
table(md$mouse_age)
```

And here are the number of cells in each of the three common developmental stages:

```{r}
table(x$species, x$common_stage)
```

In total, `r nrow(md) - nrow(x)` mouse cells from E10-E14, P4, and P14 were removed for the following proportion test, leaving `r nrow(x)` human and mouse cells in total.

Before running `propeller`, we generate a dataframe with the required sample information.

```{r}
# species, common stage, and dataset of each sample for input to design matrix
sample_info <- table(x$sample, x$species, x$common_stage, x$dataset_name) %>%
  as.data.frame() %>%
  filter(Freq != 0) %>%
  select(-Freq) %>%
  rename(sample = Var1, species = Var2, common_stage = Var3, dataset_name = Var4) %>%
  mutate(
    dataset_name = str_remove(dataset_name, "_full_cerebellum")
  ) %>%
  arrange(sample) # sort by sample to match output of `propeller` test

# show number of samples for each species and developmental stage
table(sample_info$species, sample_info$common_stage)

dim(sample_info)

sample_info
```

We now run `propeller` to compare if there are any significant differences in cell type proportions between human and mouse for each of the three developmental stages.

```{r}
# run the `propeller` test for each of the three developmental stages
per_stage_props <- list()
per_stage_results <- list()
for (dev_stage in c(1:3)) {
  # get all cells for the developmental stage
  sub_x <- x[x$common_stage == dev_stage, ]

  message(sprintf("`propeller` results for developmental stage %s", dev_stage))
  per_stage_props[[dev_stage]] <- getTransformedProps(
    clusters = sub_x$cluster,
    sample = sub_x$sample,
    transform = "logit"
  ) %>%
    pluck("Proportions") %>% # get this list item
    as.data.frame() %>%
    rename(proportion = Freq) %>%
    merge(x = ., y = sample_info, by = "sample", all.x = TRUE)
  per_stage_results[[dev_stage]] <- propeller(
    clusters = sub_x$cluster,
    sample = sub_x$sample,
    group = sub_x$species,
    transform = "logit"
  )

  # save proportions and results as CSV
  write_csv(
    x = per_stage_props[[dev_stage]],
    file = file.path(out_dir, paste0("dev_stage_", dev_stage, "_proportions.csv"))
  )
  write_csv(
    x = per_stage_results[[dev_stage]],
    file = file.path(out_dir, paste0("dev_stage_", dev_stage, "_results.csv"))
  )

  print(head(per_stage_props[[dev_stage]]))
  print(per_stage_results[[dev_stage]])
}
```

Let's plot the proportion of UBCs (cluster 2) in human and mice.

```{r}
#| label: fig-ubc-prop-per-stage
#| fig-cap: "Proportion of UBCs in each sample for each common developmental stage."
#| fig-width: 8
#| fig-height: 6

# select UBCs
ubc_props <- map(
  .x = per_stage_props,
  .f = \(prop) {
    # get UBC proportions
    ubc_props <- prop[prop$clusters == "2", ]
    return(ubc_props)
  }
) %>%
  list_rbind()

# FDRs to label on the plot
ubc_fdrs <- map(
  .x = per_stage_results,
  .f = \(res) {
    # get UBC FDRs
    ubc_fdrs <- res["2", ] %>%
      select(FDR)
    return(ubc_fdrs)
  }
) %>%
  list_rbind() %>%
  remove_rownames %>%
  cbind(., common_stage = 1:3)

# plot UBC proportions
.plt <- ggplot(ubc_props, aes(x = species, y = proportion)) + 
  geom_boxplot(width = 0.6, outlier.shape = NA) + 
  geom_jitter(aes(shape = dataset_name), width = 0.2) + 
  geom_text(
    data = ubc_fdrs,
    mapping = aes(x = "mouse", y = 0.62, label = sprintf("adj. p = %s", round(FDR, digits = 3))),
    nudge_x = 0.5,
    inherit.aes = FALSE
  ) + 
  facet_wrap(vars(common_stage)) + 
  theme_light()

.plt
```

We can also plot the FDRs from each stage to see which clusters show significantly different proportions during development.

```{r}
#| label: fig-fdrs-per-stage
#| fig-cap: "Heatmap showing the FDRs for each cluster after running `propeller` on each developmental stage. \\*FDR < 0.05, \\**FDR < 0.01, \\***FDR < 0.001."
#| fig-width: 3
#| fig-height: 10

# collect FDRs from each stage into a matrix
per_stage_fdrs <- map(
  .x = per_stage_results,
  .f = \(res) {
    fdrs <- res %>%
      # re-order rows by cluster number
      arrange(BaselineProp.clusters) %>%
      # take the FDR column
      select(FDR)
    
    return(fdrs)
  }
) %>%
  list_cbind() %>%
  rename(`1` = FDR...1, `2` = FDR...2, `3` = FDR...3) %>%
  as.matrix()
rownames(per_stage_fdrs) <- c(0:17)

# log-transform the FDRs
per_stage_fdrs <- -log10(per_stage_fdrs)
per_stage_fdrs

# colour function
breaks <- seq(0, 3, by = 0.5)
col_fun <- circlize::colorRamp2(
  breaks = breaks,
  colors = scales::pal_viridis(direction = 1, option = "magma")(length(breaks))
)

# label function (for labelling significance of values)
label_fdr <- function(j, i, x, y, w, h, fill) {
  # generate matrix with the labels
  sig_label <- as.data.frame(per_stage_fdrs) %>%
    mutate(across(
      everything(),
      ~ case_when(
        . > -log10(0.001) ~ "***",
        . > -log10(0.01) ~ "**",
        . > -log10(0.05) ~ "*",
        .default = ""
      )
    )) %>%
    as.matrix()
  
  # draw the labels
  grid.text(
    label = sprintf("%s", pindex(sig_label, i, j)),
    x = x,
    y = y,
    gp = gpar(fontsize = 12)
  )
}

.plt <- Heatmap(
  matrix = per_stage_fdrs,
  col = col_fun,
  layer_fun = label_fdr,
  row_title = "cluster",
  row_title_side = "left",
  column_title = "developmental stage",
  column_title_side = "bottom",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  column_names_side = "bottom",
  column_names_rot = 0,
  heatmap_legend_param = list(
    title = "-log10(FDR)",
    title_gp = gpar(fontsize = 10, fontface = "plain"),
    grid_width = unit(0.25, "in"),
    at = breaks,
    labels = c(breaks[-length(breaks)], ">=3"),
    legend_height = unit(4, "in")
  ),
  width = unit(ncol(per_stage_fdrs)*0.5, "in"),
  height = unit(nrow(per_stage_fdrs)*0.5, "in")
)
.plt
```

### Running propeller on the RL dataset controlling for age

Here, we run `propeller` again for all human vs. mouse samples (similar to @sec-human-vs-mouse) except that we control for the developmental stage (see the design matrix in the code below).

::: {.callout-note}
Note that only the 35 samples that have a common developmental stage are included in this analysis (i.e., mouse samples from E10-E14, P4, and P14 are not included here).
:::

```{r}
# create design matrix (each row corresponds to a sample)
species <- sample_info$species
common_stage <- sample_info$common_stage
controlled_design <- model.matrix(~ 0 + species + common_stage)
controlled_design

# run `propeller`
controlled_props <- getTransformedProps(
  clusters = x$cluster,
  sample = x$sample,
  transform = "logit"
)
controlled_results <- propeller.ttest(
  prop.list = controlled_props,
  design = controlled_design,
  contrasts = limma::makeContrasts(specieshuman-speciesmouse, levels = controlled_design),
  robust = TRUE, # must specify argument
  trend = FALSE, # must specify argument
  sort = TRUE # must specify argument
)

write_csv(
  {
    controlled_props %>%
      pluck("Proportions") %>% # get this list item
      as.data.frame() %>%
      rename(proportion = Freq) %>%
      merge(x = ., y = sample_info, by = "sample", all.x = TRUE)
  },
  file = file.path(out_dir, "controlled_proportions.csv")
)
write_csv(
  controlled_results,
  file = file.path(out_dir, "controlled_results.csv")
)

controlled_results
```

We see that even after controlling for the developmental stage, the average proportion of UBCs (cluster 2) is still significantly higher in humans than mice (FDR = `r sprintf("%.2g", controlled_results["2", "FDR"])`).

